\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{Cosets}

\begin{document}
\begin{abstract}
We introduce cosets of subgroups, and show that they partition the
group.
\end{abstract}
\maketitle


\begin{definition}
  Let $S$ be a subgroup of $G$. A \textbf{left }\dfn{coset} of $S$ is a set of
  the form
  \[
  gS := \{g\cdot s : s\in S\}.
  \]
  where $g\in G$. A \textbf{right }\dfn{coset} of $S$ is a set of the form
  \[
  Sg := \{s\cdot g : s\in S\}.
  \]
\end{definition}

\begin{remark}
  Below we will work exculsively with left cosets. However, all of the
  same results are true for right cosets, and in fact the proofs are
  essentially the same.
\end{remark}



\begin{lemma}[Cosets are the size of the subgroup]\label{P:CSG}
  Let $S$ be a subgroup of $G$. For any $g\in G$,
  \[
  |S| = |gS|.
  \]
  \begin{proof}
    Define a function
    \begin{align*}
      \eta: S &\to g S\\
      s &\mapsto g\cdot s.
    \end{align*}
    We claim that $\eta$ is bijective. First note that by definition
    of cosets $\eta$ is surjective.

    Now we must show that $\eta$ is injective. Suppose that $\eta(s) =
    \eta(t)$. Write with me:
    \begin{align*}
      \eta(s) &= \eta(t) \\
      gs &= gt\\
      g^{-1} g s &= g^{-1} g t\\
      s &= t.
    \end{align*}
    Thus if two elements are equal in the image of $\eta$, the must
    have come from equal elements in the preimage of $\eta$. So we see
    that $\eta$ is injective.

    Since $\eta$ is bijective, $|S|= |gS|$.
  \end{proof}
\end{lemma}

\begin{corollary}[Cosets have the same order]
  Let $S$ be a subgroup of $G$. All left cosets of $G$ have the same
  order.
\end{corollary}

\begin{remark}
  The same result holds replacing ``left'' cosets with ``right''
  cosets.
\end{remark}



\begin{lemma}[Cosets are equal or disjoint]\label{L:cosetED}
  Let $S$ be a subgroup of $G$. Left cosets of $S$ are either equal or
  disjoint.
  \begin{proof}
    We will prove that if left cosets are not disjoint, then they are
    equal. Suppose that there exists $c$ such that $c\in aS \cap
    bS$. Write with me
    \[
    as_1 = c = bs_2.
    \]
    So, $as_1 = bs_2$, this means
    \[
    as_1 S = bs_2 S.
    \]
    But $as_1S = aS$ and $bs_2 S = bS$, so $aS = bS$.
  \end{proof}
\end{lemma}



\begin{theorem}[Cosets partition the group]\label{T:CPG}
  Let $S$ be a subgroup of $G$, 
  \[
  G = \union_{g\in G} gS
  \]
  and for all $a,b\in G$, if $aS \ne bS$,
  \[
  aS\cap bS = \emptyset.
  \]
  \begin{proof}
    All that remains to be shown is that $G = \union_{g\in G}
    gS$. Since $e\in S$, this is true. The rest follows from
    Lemma~\ref{L:cosetED}.
  \end{proof}
\end{theorem}




%% \begin{lemma}[Criterion for coset equality]
%%   Let $S$ be a subgroup of $G$. If $a,b\in G$,
%%   \[
%%   aS = bS \Leftrightarrow ab^{-1}\in S.
%%   \]
%%   \begin{proof}
%%     $(\Rightarrow)$ If $aS = bS$, then there exists $s\in S$ with
%%     \begin{align*}
%%       a &= b\cdot s,\\
%%       ab^{-1} &= s,
%%     \end{align*}
%%     so $ab^{-1}\in S$.


%%     $(\Leftarrow)$ If $ab^{-1}\in S$, then for some $s\in S$, $a = bs$
%%     and so $aS =bS$.
%%   \end{proof}
%% \end{lemma}






\end{document}
