\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{Lagrange's theorem}

\begin{document}
\begin{abstract}
  We prove Lagrange's theorem and explore its consequences.
\end{abstract}
\maketitle

\begin{definition}\index{index@$[G:S]$}
  Let $S$ be a subgroup of a finite group $G$, the number of left
  cosets of $S$ is called the \dfn{index} of $S$ in $G$, denoted
  \[
  [G:S] = \text{(number of left cosets)}.
  \]
\end{definition}



\begin{example}[Index of $\boldsymbol{2\pmb\Z$ in $\pmb\Z}$]
  The left cosets of $2\Z\subgp \Z$ are
  \begin{align*}
    0+2\Z &= \{\dots, -8,-6,-4,-2,0,2,4,6,8,\dots\},\\
    1+2\Z &= \{\dots, -7,-5,-3,-1,1,3,5,7,9,\dots\},
  \end{align*}
  hence $[\Z:2\Z] = 2$.
\end{example}

\begin{example}[Index of $\boldsymbol{5\pmb\Z$ in $\pmb\Z}$]
  The left cosets of $5\Z\subgp \Z$ are
  \begin{align*}
    0+5\Z &= \{\dots, -20,-15,-10,-5,0,5,10,15,20,\dots\},\\
    1+5\Z &= \{\dots, -19,-14,-9 ,-4,1,6,11,16,21,\dots\},\\
    2+5\Z &= \{\dots, -18,-13,-8 ,-3,2,7,12,17,22,\dots\},\\
    3+5\Z &= \{\dots, -17,-12,-7 ,-2,3,8,13,18,23,\dots\},\\
    4+5\Z &= \{\dots, -16,-11,-6 ,-1,4,9,14,19,24,\dots\},
  \end{align*}
  hence $[\Z:5\Z] = 5$.
\end{example}

\begin{exercise}
  Prove that $[\Z:n\Z]= n$.
\end{exercise}



\begin{example}[Index of  $\boldsymbol{\<r\>$ in $D_3}$]
  Consider again our partial Cayley table for $D_3$ with $\< r\>$ in
  the upper left-hand corner:
  \[
    \renewcommand{\arraystretch}{1.6}
    \begin{array}{c!{\vline width 2pt}cccccc}
      (D_3,\circ)& e                              & r                              & r^2                            & f                             & r f                                   & r^2 f  \\  \Xhline{2pt}
      e          & e\cellcolor{blue!12!white}     & r\cellcolor{blue!12!white}    & r^2\cellcolor{blue!12!white}   & f \cellcolor{red!12!white}    & r f \cellcolor{blue!33!red!24!white}   & r^2f \cellcolor{blue!66!red!24!white} \\  
      r                                  & r\cellcolor{blue!12!white}    & r^2\cellcolor{blue!12!white}   & e \cellcolor{blue!12!white}    & rf\cellcolor{red!12!white} & r^2f \cellcolor{blue!33!red!24!white}    & f \cellcolor{blue!66!red!24!white}   \\  
      r^2                                 & r^2\cellcolor{blue!12!white}   & e\cellcolor{blue!12!white}     & r\cellcolor{blue!12!white}    & r^2 f \cellcolor{red!12!white}   & f \cellcolor{blue!33!red!24!white} & rf \cellcolor{blue!66!red!24!white}   \\  
      f        & f \cellcolor{red!12!white}    & r^2f \cellcolor{red!12!white}   & r f \cellcolor{red!12!white} &     &   &   \\  
      r f       & rf \cellcolor{blue!33!red!24!white}   & f \cellcolor{blue!33!red!24!white} & r^2f \cellcolor{blue!33!red!24!white}   &    &      &     \\  
      r^2 f       & r^2 f\cellcolor{blue!66!red!24!white} & rf \cellcolor{blue!66!red!24!white}    & f \cellcolor{blue!66!red!24!white}  &     &     &      \\  
    \end{array}
    \]
    Recall the left cosets are below the subgroup. Note there are only
    two distinct left cosets. Here they are
    \begin{align*}
      \<r\> &= \{e, r, r^2\},\\
      f\<r\> &= \{f, rf, r^2f\}.
    \end{align*}
    From this we see that $[D_3: \<r\>] = 2$.
\end{example}


\begin{example}[Index of  $\boldsymbol{\<f\>$ in $D_3}$]
  Consider again our partial Cayley table for $D_3$ with $\< r\>$ in
  the upper left-hand corner:
      \[
    \renewcommand{\arraystretch}{1.6}
    \begin{array}{c!{\vline width 2pt}cccccc}
      (D_3,\circ)& e                         & f                              & rf                            & r^2f                             & r                                    & r^2  \\  \Xhline{2pt}
      e          & e\cellcolor{blue!12!white}     & f\cellcolor{blue!12!white}    & rf\cellcolor{red!12!white}   & r^2f \cellcolor{blue!33!red!24!white}     & r  \cellcolor{blue!66!red!24!white}   & r^2 \cellcolor{blue!36!white} \\  
      f                & f\cellcolor{blue!12!white}    & e\cellcolor{blue!12!white}   & r^2 \cellcolor{red!12!white}    & f\cellcolor{blue!33!red!24!white} & r^2f \cellcolor{blue!66!red!24!white}    & rf \cellcolor{blue!36!white}   \\  
      rf                & rf\cellcolor{red!12!white}   & r\cellcolor{red!12!white}     &    &    &  &     \\  
      r^2f       & r^2f \cellcolor{blue!33!red!24!white}   & r^2 \cellcolor{blue!33!red!24!white}   &   &      &   &   \\  
      r      & r \cellcolor{blue!66!red!24!white}   & rf \cellcolor{blue!66!red!24!white} &    &   &      &     \\  
      r^2       & r^2 \cellcolor{blue!36!white} & r^2f \cellcolor{blue!36!white}    &   &    &     &      \\  
    \end{array}
    \]
    Recall the left cosets are below the subgroup. Note, there are
    only $3$ distinct left cosets.  Here they are
    \begin{align*}
      \<f\> &= \{e, f\},\\
      r\<f\> &= \{r, rf\},\\
      r^2\<f\> &= \{r^2, r^2f\}.
    \end{align*}
    From this we see that $[D_3: \<f\>] = 3$.
\end{example}


\begin{exercise}
  Prove that if $G$ is a group and $S\subgp G$, then $[S:\<e\>] = |S|$.
\end{exercise}


Now we can decorate our lattice diagrams with the index of the
subgroup.

\begin{example}[Index of subgroups for $\boldsymbol{\pmb\Z_6}$]
   Here is the subgroup lattice diagram for $\Z_6$ along with the
   index of each subgroup:   
  \[
  \begin{tikzcd}
    & \Z_6 \ar[ld,-,"{2=[\Z_6:\<2\>]}",swap]  \ar[rdd,-,"{3=[\Z_6:\<3\>]}"] &       \\
    \<2\> \ar[rdd,-,"{3=[\<2\>:\<0\>]}",swap] &  &\\
   & &  \< 3\> \ar[ld,-,"{2=[\<3\>:\<0\>]}"]        \\   
    & \<0\> &
  \end{tikzcd}
  \]
\end{example}



\begin{example}[Index of subgroups for $\boldsymbol{\pmb\Z_8}$]
   Here is the subgroup lattice diagram for $\Z_8$ along with the
   index of each subgroup:   
   \[
   \begin{tikzcd}
     \Z_8  \ar[d,-,"{2 = [\Z_8:\<2\>]}"] \\
     \<2\> \ar[d,-,"{2 = [\<2\>:\<4\>]}"] \\
     \<4\> \ar[d,-,"{2 = [\<4\>:\<0\>]}"] \\   
     \<0\> 
\end{tikzcd}
\]
\end{example}


\begin{example}[Index of subgroups for $\boldsymbol{D_3}$]\index{symmetries!of the equilateral triangle}
  Here is the subgroup lattice diagram for $D_3$ along with the
   index of each subgroup:   
  \[
  \begin{tikzcd}
    & D_3 \ar[rdd,-,"3"]\ar[rrdd,-,"3"]\ar[rrrdd,-,"3"]  \ar[ld,-,"2",swap] & & &      \\
    \<r\> \ar[rdd,-,"3",swap]       &       &  & &  \\
    &       &  \< f\> \ar[ld,-,"2",swap]   &  \< rf\> \ar[lld,-,"2",swap]       &  \< r^2f\> \ar[llld,-,"2"]        \\   
    & \<e\> & & &
  \end{tikzcd}
  \]
\end{example}


\begin{example}[Index of subgroups for $\boldsymbol{D_4}$]
  Here is the subgroup lattice diagram for $D_4$ along with the index
  of each subgroup:
  \[
\begin{tikzcd}
   &    & D_4 \ar[rd,-,"2"] \ar[d,-,"2"] \ar[ld,-,"2",swap] &       \\
&\<r^2,f\>  \ar[rd,-,"2"] \ar[d,-,"2"] \ar[ld,-,"2",swap]& \<r \>  \ar[d,-,"2"]     & \<r^2,rf\> \ar[ld,-,"2",swap] \ar[d,-,"2"] \ar[rd,-,"2"]\\
\<f\>\ar[rrd,-,"2",swap]& \<r^2f\> \ar[rd,-,"2"]&\< r^2\> \ar[d,-,"2"]  &   \<rf\> \ar[ld,-,"2",swap]& \<r^3f\>\ar[lld,-,"2"]     \\   
  &     & \<e\> &  &
\end{tikzcd}
\]
\end{example}




\begin{example}[Index of subgroups for $\boldsymbol{Q_8}$]\index{quaternion group}
  Here is the subgroup lattice diagram for $Q_8$ along with the index
  of each subgroup:
  \[
  \begin{tikzcd}
    & Q_8 \ar[rd,-,"2"] \ar[d,-,"2"] \ar[ld,-,"2",swap] &       \\
    \<i\>  \ar[rd,-,"2",swap] & \<j\>  \ar[d,-,"2"]     & \<k\> \ar[ld,-,"2"]\\
    &\< -1\> \ar[d,-,"2"]&        \\   
    & \<1\> &
  \end{tikzcd}
  \]
\end{example}




\begin{example}[Index of subgroups for $\boldsymbol{A_4}$]\index{alternating group}
  Setting $\sigma = (1 \ 2 \ 3)$ and $\tau = (2 \ 3 \ 4)$, here is the
  subgroup lattice diagram for $A_4$ along with the index
  of each subgroup:
  \[
  \begin{tikzpicture}
  \node[scale=.6] at (0,0) {\begin{tikzcd}
                   &   & & A_4   \ar[lld,-,"3",swap] \ar[rdd,-,"4",swap] \ar[rrdd,-,"4",swap]  \ar[rrrdd,-,"4",swap]  \ar[rrrrdd,-,"4"]   & &  & &        \\
                   & \<\sigma\tau, \sigma^2\tau^2\> \ar[ldd,-,"2",swap] \ar[dd,-,"2"] \ar[rdd,-,"2"] &  &              & &  & &  \\
                   &                                  &            &   & \<\sigma\> \ar[ldd,-,"3",swap]& \<\tau\> \ar[lldd,-,"3",swap] & \<\ \tau\sigma^2\> \ar[llldd,-,"3",swap]& \< \sigma \tau\sigma^2 \>\ar[lllldd,-,"3"]       \\
   \<\sigma\tau\>\ar[drrr,-,"2",swap]  & \<\sigma^2\tau^2\> \ar[drr,-,"2"] & \<\sigma^2\tau\sigma^2\> \ar[dr,-,"2"]& & &  & &    \\
                   &                               &   & \< e\>       & & & &  
  \end{tikzcd}};
  \end{tikzpicture}
  \]
\end{example}




\begin{exercise}
  Draw lattice diagrams for $\Z_n$ for $n=2,3\dots, 12$, include the
  index of each subgroup containment.
\end{exercise}



And now we are ready to state and prove Lagrange's theorem.



\begin{theorem}[Lagrange's theorem]\index{Lagrange's theorem}\label{T:lagrange}
  If $G$ is a finite group with subgroup $S$, then 
  \[
  |G| = [G:S]\cdot |S|.
  \]
  \begin{proof}
    First note that by Lemma~\ref{L:CSG}, $|S| = |aS|$ for all
    $a\in G$. By Theorem~\ref{T:CPG}, $G$ is partitioned by the left
    cosets of $S$. This means that every element of $G$ is in exactly
    one left coset. Hence $|G| = [G:S]\cdot |S|$.
  \end{proof}
\end{theorem}

We can witness the importance of Lagrange's theorem by its
corollaries.

\begin{corollary}[Subgroup's order divides the group's order]
  If $G$ is a finite group with subgroup $S$, then the order of $S$
  divides the order of $G$.
\end{corollary}

\begin{corollary}[Element's order divides the group's order]\label{C:Leodgo}
  Let $G$ be a finite group with $a\in G$. In this case, the order of
  $a$ divides the order of $G$.
\end{corollary}


\begin{corollary}[Prime order group implies trivial subgroups]
  Let $G$ be a group where $|G|= p$, a prime number. Then the only
  subgroups of $G$ are itself and $\{e\}$.
\end{corollary}

\begin{corollary}[Prime order group implies cyclic]
  Let $G$ be a group where $|G|= p$, a prime number. Then $G$ is
  cyclic and any element $g\in G$, $g\ne e$, can act as a generator.
\end{corollary}


\begin{corollary}[Index is multiplicative]
  Let $S \subgp T \subgp G$ be subgroups of a finite group
  $G$. Then
  \[
  [G:S] = [G:T]\cdot [T:S].
  \]
\end{corollary}


\begin{corollary}[Fermat's little theorem]\index{Fermat's little theorem}
  Let $p\in \N$ be a prime number. Given $a\in\Z_p$,
  \[
  a^{p-1}\equiv 1 \pmod{p}.
  \]
\end{corollary}


\begin{corollary}[Euler's little theorem]\index{Euler's little theorem}
  Let $n\in \N$. Given $a\in\Z_n$,
  \[
  a^{\eulerphi(n)}\equiv 1 \pmod{n}.
  \]
\end{corollary}



\begin{exercise}
  Let $G$ be a group with subgroups $H$ and $K$. If $|H|=7$ and $|K| =
  9$, what are the possible values for $|H\cap K|$?
\end{exercise}


\begin{exercise}
  Let $G$ be a group with subgroups $H$ and $K$. If $|H|=15$ and $|K|
  = 24$, what are the possible values for $|H\cap K|$?
\end{exercise}


\begin{exercise}
  Let $G$ be a group with subgroups $H$ and $K$. If $|H|=12$ and $|K|
  = 24$, what are the possible values for $|H\cap K|$?
\end{exercise}

\begin{exercise}
  Suppose that $K$ is a proper subgroup of $H$, and $H$ is a proper
  subgroup of $G$. If $|K| =12$ and $|G| = 108$, what are the possible
  orders of $H$?
\end{exercise}


\section{Converse to Lagrange?}


The converse to Lagrange's theorem would be:
\begin{quote}
  If $G$ is a group and $n$ divides $|G|$, then $G$ has a subgroup of
  order $n$.
\end{quote}

You may wonder, is the converse of Lagrange's theorem true? The answer
is ``no.''

\begin{example}[Counterexample to converse of Lagrange]\index{tetrahedral group}\index{alternating group}\index{symmetries!of the regular tetrahedron}
  Consider a group $G\iso A_4\iso \mathcal{T}_{12}$. The order of $G$
  is $12$, yet $G$ has no subgroup of order $6$.

  Note, at this point we have not proved that no such subgroup exists,
  we have merely stated this.
\end{example}

Now that we know there is no converse to Lagrange's theorem, we will
state several partial converses.


\begin{exercise}
  Let $G$ be a cyclic of order $n$, then if $d|n$, $G$ has a subgroup
  of order $d$.
\end{exercise}




The theorem below provides a partial converse. We will not prove
Sylow's theorem in this course, but it is good to know that such a
theorem exists.

\begin{theorem}[Sylow's theorem]
  Let $p$ be a prime number. If $p^n$ divides $|G|$, then for $k =
  1,\dots, n$ there exists subgroups $S_p, S_{p^2},\dots,S_{p^n}$ such
  that
  \[
  |S_{p^k}| = p^k.
  \]
  The subgroup $S_{p^n}$ is called a \textbf{Sylow
    $\boldsymbol{p}$-subgroup}\index{Sylow p-subgroup@Sylow $p$-subgroup}
\end{theorem}

\begin{corollary}[Prime order subgroups]
  Let $G$ be a group and $p$ be a prime number that divides
  $|G|$. Then $G$ has a subgroup of order $p$.
\end{corollary}

\begin{exercise}
  Find a Sylow $p$-subgroup of $S_3$, $S_4$, and $S_5$.
\end{exercise}





\section{Lagrange's question}

The formal definition of a group came well after the time of
Lagrange. This being the case, it is somewhat interesting that a
fundamental theorem in group theory bears his name. How did this
happen? Let's give a little history lesson. Given a quadratic equation
\[
ax^2 + bx + c = 0
\]
we've (as humans) have know the \textit{quadratic
  formula}\index{quadratic formula} for nearly $4000$ years. Here it
is
\[
x = \frac{-b \pm \sqrt{b^2-4ac}}{2a}
\]
Later Ferro and Tartaglia solved the cubic formula around $1500$ CE. Given,
\[
ax^3 + bx^2 + cx + d = 0
\]
we can find a solution with
\begin{gather}
\begin{align*}
  x &= \sqrt[3]{\left(\frac{-b^3}{27a^3}+\frac{bc}{6a^2}-\frac{d}{2a}\right)+\sqrt{\left(\frac{-b^3}{27a^3}+\frac{bc}{6a^2}-\frac{d}{2a}\right)^2 + \left(\frac{c}{3a}-\frac{b^2}{9a^2}\right)^3}} \\
    &+ \sqrt[3]{\left(\frac{-b^3}{27a^3}+\frac{bc}{6a^2}-\frac{d}{2a}\right)+\sqrt{\left(\frac{-b^3}{27a^3}+\frac{bc}{6a^2}-\frac{d}{2a}\right)^2 - \left(\frac{c}{3a}-\frac{b^2}{9a^2}\right)^3}} -\frac{b}{3a}.
\end{align*}
\end{gather}
This formula was popularized by Cardano in the mid $1500$s.


At this time, Cardano's student, Ferrari joined the quest. He was able
to defeat Tartaglia in a ``mathematical duel'' and also discovered the
quartic formula.  Given
\[
ax^4 + bx^3 + cx^2 + dx + e = 0
\]
we can find solutions with
\begin{gather}
  \begin{align*} %% Copied from https://github.com/curtisbright/quartic/
    x &= \frac{1}{12a}\Big(-3b\pm\big(\sqrt{3\big(3b^2-8ac+2a\root3\of{4\big(2c^3-9bcd+27ad^2+27b^2e-72ace+\sqrt{(2c^3-9bcd+27ad^2+27b^2e-72ace)^2-4(c^2-3bd+12ae)^3}\big)}+2a\root3\of{4\big(2c^3-9bcd+27ad^2+27b^2e-72ace-\sqrt{(2c^3-9bcd+27ad^2+27b^2e-72ace)^2-4(c^2-3bd+12ae)^3}\big)}\big)}\\
    &\pm\sqrt{3\big(3b^2-8ac+2a\big({-1+\sqrt{-3}\over2}\big)\root3\of{4\big(2c^3-9bcd+27ad^2+27b^2e-72ace+\sqrt{(2c^3-9bcd+27ad^2+27b^2e-72ace)^2-4(c^2-3bd+12ae)^3}\big)}+2a\big({-1-\sqrt{-3}\over2}\big)\root3\of{4\big(2c^3-9bcd+27ad^2+27b^2e-72ace-\sqrt{(2c^3-9bcd+27ad^2+27b^2e-72ace)^2-4(c^2-3bd+12ae)^3}\big)}\big)}\big)\\
    &\pm\sgn\big(\big(\sgn\big(-b^3+4abc-8a^2d\big)-{1\over2}\big)\big(\sgn\big(\max((2c^3-9bcd+27ad^2+27b^2e-72ace)^2-4(c^2-3bd+12ae)^3,\min(3b^2-8ac,3b^4+16a^2c^2+16a^2bd-16ab^2c-64a^3e))\big)-{1\over2}\big)\big)\sqrt{3\big(3b^2-8ac+2a\big({-1-\sqrt{-3}\over2}\big)\root3\of{4\big(2c^3-9bcd+27ad^2+27b^2e-72ace+\sqrt{(2c^3-9bcd+27ad^2+27b^2e-72ace)^2-4(c^2-3bd+12ae)^3}\big)}+2a\big({-1+\sqrt{-3}\over2}\big)\root3\of{4\big(2c^3-9bcd+27ad^2+27b^2e-72ace-\sqrt{(2c^3-9bcd+27ad^2+27b^2e-72ace)^2-4(c^2-3bd+12ae)^3}\big)}\big)}\Big).
  \end{align*}
\end{gather}

Two hundred years pass, and Lagrange enters the scene. Lagrange wanted
further formulas that would solve equations of degree $5$, $6$, and so
on. As part of this pursuit, Lagrange sought the answer to the
following question:


\begin{quote}
  Consider a polynomial of $n$ variables, if one permutes the
  variables, will the number of different polynomials found always
  divide $n!$?
\end{quote}

So, for example, if $F(x_1,x_2,x_3) = x+ y+z$, then every permutation
of the variables will be equal to $F(x_1,x_2,x_3)$, and $1|3$.

On the other hand, if $F(x_1,x_2,x_3) = x_1$, then we find $3$
different polynomials, and $3|3$.

Of course, $F$ need not be linear, and could be quite
complicated. Check out our next lemma.



\begin{lemma}[Permutations that fix $\boldsymbol F$ form a subgroup]
  Let $F(x_1,\dots,x_n)$ be a polynomial of $n$ variables.
  \[
  H = \{\epsilon\in S_n: F(x_{\epsilon(1)},\dots, x_{\epsilon(n)}) = F(x_1,\dots,x_n)\},
  \]
  then $H\subgp S_n$
  \begin{proof}
    First note that $e\in H$ and that $H\subset S_n$. Now note that if
    $\sigma\in H$, $\sigma^{-1}\in H$. Also, $\epsilon\circ \sigma^{-1}\in
    H$. Thus by the subgroup criterion, Theorem~\ref{T:sc}, $H\subgp
    S_n$.
  \end{proof}
\end{lemma}


\begin{lemma}[Cosets and polynomials]\label{L:cap}
  Let $F(x_1,\dots,x_n)$ be a polynomial of $n$ variables and consider
  the subgroup of $S_n$
  \[
  H = \{\epsilon\in S_n: F(x_{\epsilon(1)},\dots, x_{\epsilon(n)}) = F(x_1,\dots,x_n)\}.
  \]
  We have that 
  \[
  F(x_{\sigma(1)},\dots, x_{\sigma(n)}) = F(x_{\tau(1)},\dots, x_{\tau(n)})
  \]
  if and only if there exists $\pi \in S_n$ such that $\sigma,\tau\in\pi H$.
  \begin{proof}
    $(\Rightarrow)$ Suppose that
    \[
    F(x_{\sigma(1)},\dots, x_{\sigma(n)}) = F(x_{\tau(1)},\dots, x_{\tau(n)})
    \]
    Now, note that if $\epsilon \in H$,
    \[
    F(x_{\sigma\circ\epsilon(1)},\dots, x_{\sigma\circ\epsilon(n)}) =
    F(x_{\tau\circ\epsilon(1)},\dots,
    x_{\tau\circ\epsilon(n)}),
    \]
    meaning that $\sigma H \cap \tau H \ne \emptyset$. By
    Theorem~\ref{T:CPG}, $S_n$ is partitioned by the left cosets of
    $H$. This means that every permutation of $S_n$ is found in some
    left coset, and that cosets are either equal or disjoint. This
    means that both $\sigma$ and $\tau$ are in the same left coset of
    $H$. Hence there exists $\pi \sigma,\tau\in\pi H\in S_n$ such that
    $\sigma,\tau\in\pi H$.

    $(\Leftarrow)$ Suppose that $\sigma,\tau\in \pi H$. Then
    there exist $\epsilon_1,\epsilon_2\in H$ such that
    \begin{align*}
      F(x_{\sigma(1)},\dots, x_{\sigma(n)}) &=
    F(x_{\pi\circ\epsilon_1(1)},\dots,
    x_{\pi\circ\epsilon_1(n)}), \\
      F(x_{\tau(1)},\dots, x_{\tau(n)}) &=
    F(x_{\pi\circ\epsilon_2(1)},\dots,
    x_{\pi\circ\epsilon_2(n)}).
    \end{align*}
    Since the right-hand sides of both of these equations are equal,
    we have completed the proof of the lemma.
  \end{proof}
\end{lemma}

\begin{theorem}[A resolution to the question]
  Let $F(x_1,\dots,x_n)$ be a polynomial of $n$ variables and consider
  \[
  H = \{\sigma\in S_n: F(x_{\sigma(1)},\dots, x_{\sigma(n)}) = F(x_1,\dots,x_n)\}.
  \]
  The number of distinct polynomials found by permuting the
  variables of $F$ will divide $n!$.
    \begin{proof}
      From Lemma~\ref{L:cap}, we see that $[S_n:H]$ is the number of
      distinct polynomials found by permuting the variables. By
      Lagrange's theorem, Theorem~\ref{T:lagrange}, the index of a
      subgroup divides the order of the group. Since the order of
      $S_n$ is $n!$, we have proved the theorem.
      \end{proof}
\end{theorem}


However, mathematics is hard. Even with this answer Lagrange was
unable to find a formula for a degree $5$ equation, making (perhaps in
his eyes) his work a \textbf{failure}. But Lagrange, despite failing
in this one regard, \textbf{was not a failure.} He laid the
foundations for mathematics that he would not live to see.




For some interesting extra reading check out:
\begin{itemize}
\item \link[\textit{A history of Lagrange's theorem on groups},
  R.\ Roth, Mathematics Magazine. April (2001), 99--198.]{https://www.jstor.org/stable/2690624}.
\end{itemize}

\end{document}
