\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{Quotient groups}

\begin{document}
\begin{abstract}
  We introduce quotient groups.
\end{abstract}
\maketitle



We are now going to give a construction that will allow us to make new
groups from old groups. Here's the idea: Given a group $G$ with
subgroup $N\subgp G$, we will make a group on the set of left cosets
of $N$. This group will be called the \textit{quotient group} of $G$
modulo $N$, and will be denoted $G/N$.  As a first step, we must look
at a special set of subgroups called \textit{normal subgroups}.




\section{Normal subgroups}




\begin{definition}
  Let $(G,\star)$ be a group. A subgroup $N\subgp G$ is a \dfn{normal subgroup}
  if
  \[
  a\star N = N\star a,
  \]
  meaning every left coset is equal to the right coset involving the
  same group element. In this case we write $N\normal G$. \index{normal@$\normal$}\index{N<G@$N\normal G$}
\end{definition}

\begin{corollary}[Abelian implies normal]
  If $G$ is an Abelian group and $N\subgp G$, then $N\normal G$.
\end{corollary}





\begin{exercise}
  Prove that every subgroup of index two is normal.
\end{exercise}

\begin{exercise}
  Prove that every subgroup of $Q_8$ is normal. Note, this is
  interesting, as $Q_8$ is not Abelian.
\end{exercise}



\begin{example}[$\boldsymbol{\<r\>\normal D_3}$]\index{symmetries!of the equilateral triangle}
  If we arrange a Cayley table for $D_3$ appropriately, we can ``see''
  the subgroup $\<r\>$ is normal in $D_3$.
    \[
    \renewcommand{\arraystretch}{1.6}
    \begin{array}{c!{\vline width 2pt}cccccc}
      (D_3,\circ)& e                              & r                              & r^2                            & f                             & r f                                   & r^2 f  \\  \Xhline{2pt}
      e          & e\cellcolor{blue!12!white}     & r\cellcolor{blue!12!white}    & r^2\cellcolor{blue!12!white}   & f \cellcolor{red!12!white}    & r f \cellcolor{blue!33!red!24!white}   & r^2f \cellcolor{blue!66!red!24!white} \\  
      r                                  & r\cellcolor{blue!12!white}    & r^2\cellcolor{blue!12!white}   & e \cellcolor{blue!12!white}    & rf\cellcolor{red!12!white} & r^2f \cellcolor{blue!33!red!24!white}    & f \cellcolor{blue!66!red!24!white}   \\  
      r^2                                 & r^2\cellcolor{blue!12!white}   & e\cellcolor{blue!12!white}     & r\cellcolor{blue!12!white}    & r^2 f \cellcolor{red!12!white}   & f \cellcolor{blue!33!red!24!white} & rf \cellcolor{blue!66!red!24!white}   \\  
      f        & f \cellcolor{red!12!white}    & r^2f \cellcolor{red!12!white}   & r f \cellcolor{red!12!white} &     &   &   \\  
      r f       & rf \cellcolor{blue!33!red!24!white}   & f \cellcolor{blue!33!red!24!white} & r^2f \cellcolor{blue!33!red!24!white}   &    &      &     \\  
      r^2 f       & r^2 f\cellcolor{blue!66!red!24!white} & rf \cellcolor{blue!66!red!24!white}    & f \cellcolor{blue!66!red!24!white}  &     &     &      \\  
    \end{array}
    \]
    The left cosets of $\<r\>$ are below $\<r\>$ and the right cosets
    of $\<r\>$ are to the right of $\<r\>$. Here we explicitly list the left cosets
    \begin{align*}
      \<r\> &= \{e, r, r^2\},\\
      f\<r\> &= \{f, rf, r^2f\}.
    \end{align*}
    These are equal to the right cosets,
     \begin{align*}
      \<r\> &= \{e, r, r^2\},\\
      \<r\>f &= \{f, rf, r^2f\}.
    \end{align*}
    Hence $\<r\>\normal D_3$.
\end{example}

\begin{example}[$\boldsymbol{\<f\>\not\normal D_3}$]\index{symmetries!of the equilateral triangle}
  If we arrange a Cayley table for $D_3$ appropriately, we can ``see''
  the subgroup $\<f\>$ is \textbf{not} normal in $D_3$.
    \[
    \renewcommand{\arraystretch}{1.6}
    \begin{array}{c!{\vline width 2pt}cccccc}
      (D_3,\circ)& e                         & f                              & rf                            & r^2f                             & r                                    & r^2  \\  \Xhline{2pt}
      e          & e\cellcolor{blue!12!white}     & f\cellcolor{blue!12!white}    & rf\cellcolor{red!12!white}   & r^2f \cellcolor{blue!33!red!24!white}     & r  \cellcolor{blue!66!red!24!white}   & r^2 \cellcolor{blue!36!white} \\  
      f                & f\cellcolor{blue!12!white}    & e\cellcolor{blue!12!white}   & r^2 \cellcolor{red!12!white}    & f\cellcolor{blue!33!red!24!white} & r^2f \cellcolor{blue!66!red!24!white}    & rf \cellcolor{blue!36!white}   \\  
      rf                & rf\cellcolor{red!12!white}   & r\cellcolor{red!12!white}     &    &    &  &     \\  
      r^2f       & r^2f \cellcolor{blue!33!red!24!white}   & r^2 \cellcolor{blue!33!red!24!white}   &   &      &   &   \\  
      r      & r \cellcolor{blue!66!red!24!white}   & rf \cellcolor{blue!66!red!24!white} &    &   &      &     \\  
      r^2       & r^2 \cellcolor{blue!36!white} & r^2f \cellcolor{blue!36!white}    &   &    &     &      \\  
    \end{array}
    \]
    The left cosets of $\<f\>$ are below $\<f\>$ and the right cosets
    of $\<f\>$ are to the right of $\<f\>$. Explicitly, the left
    cosets are
    \begin{align*}
      \<f\> &= \{e, f\},\\
      r\<f\> &= \{r, rf\},\\
      r^2\<f\> &= \{r^2, r^2f\},
    \end{align*}
    and the right cosets are
    \begin{align*}
      \<f\> &= \{e, f\},\\
      \<f\>r &= \{r, r^2f\},\\
      \<f\> r^2&= \{r^2, rf\}.
    \end{align*}
    Hence $\<f\>\not\normal D_3$.
\end{example}






\begin{lemma}[Normal subgroup criterion]\label{L:nsc}\index{normal subgroup criterion}
  A subgroup $N$ of a group $(G,\star)$ is normal if and only if for
  all $n\in N$ and all $g\in G$,
  \[
  g^{-1}\star n\star g\in N.
  \]
  \begin{proof}
    $(\Rightarrow)$ Suppose that $N\normal G$. In this case, for all
    $g\in G$
    \[
    g\star N = N\star g.
    \]
    This means that for all $n\in N$, there exist $n'\in N$ such
    that
    \[
    g\star n = n'\star g.
    \]
    since $G$ is a group, we may multiply both sides by $g^{-1}$ on
    the left to obtain
    \[
    g^{-1} \star n' \star g = n\in N.
    \]
    This completes this direction of the proof.

    $(\Leftarrow)$ Now suppose for all $n\in N$ and $g\in G$ 
    \[
    g^{-1}\star n\star g \in N.
    \]
    This means there is a $n'\in N$ such that
    \begin{align*}
      g^{-1}\star n\star g  &= n'\\
      n \star g &= g\star n'.
    \end{align*}
    Thus, $g\star N =N\star g$ for all $g\in G$, and we see $N$ is
    normal.
  \end{proof}
\end{lemma}




\begin{exercise}
  Let $G$ be a group with subgroups $L$, $M$, and $N$. If
  \begin{align*}
    L &\normal M,\\
    N &\normal G,
  \end{align*}
  then, $NL\normal NM$.
\end{exercise}

\begin{exercise}
  Show that if $G$ is a group, then the center (see
  Exercise~\ref{E:center}) $Z(G)$ is always normal in $G$.

  If you haven't already, you should also show that $Z(G)$ is subgroup
  of $G$.
\end{exercise}



\section{Quotient groups}

A quotient group is a group where the elements are left cosets of a
normal subgroup. Of course to make a group, we need an associative
operation.  This next lemma is inspired by Lemma~\ref{L:mawd} and
Lemma~\ref{L:mmwd}, and in fact, it proves both results.

\begin{lemma}[Coset product is well-defined]\index{well-defined}\label{L:cpwd}
  Let $N$ be a subgroup of a group $(G,\star)$ with $a,b\in G$.  The
  product on cosets
  \[
  (a\star N) \star (b\star N ) = (a\star b)\star N
  \]
  is well-defined if and only if $N$ is a normal subgroup of $G$.
  \begin{proof}
    $(\Rightarrow)$ Suppose that if
    \begin{align*}
      a\star N &= a'\star N\\
      b\star N &= b'\star N,
    \end{align*}
    then $(a\star b) \star N = (a'\star b')\star N$. We must show that
    $N$ is a normal subgroup of $G$. We claim that if $n\in N$,
    \[
    g \star N = (n\star g)\star N.
    \]
    To see this, first note
    \[
    e \star N = n \star N.
    \]
    By our assumption,
    \begin{align*}
    (e\star N) \star (g\star N) &= (e\star g)\star N \\
    &= g\star N
    \end{align*}
    and
    \[
    (n\star N) \star (g\star N) = (n\star g)\star N.
    \]
    Hence
    \[
    g\star N = (n\star g)\star N.
    \]
    Now multiplying both sides by $g^{-1}$ on the left, we have
    \begin{align*}
      (g^{-1}\star g)\star N &= (g^{-1} \star n\star g)\star N,\\
      N &= (g^{-1} \star n\star g)\star N.
    \end{align*}
    but this means that
    \[
    g^{-1} \star n\star g \in N
    \]
    for all $n\in N$ and $g\in G$. By the normal subgroup criterion,
    Lemma~\ref{L:nsc}, $N$ is normal in $G$.

    
    $(\Leftarrow)$ Suppose that $N\normal G$. We must show that if
    \begin{align*}
      a\star N &= a'\star N\\
      b\star N &= b'\star N,
    \end{align*}
    then $(a\star b) \star N = (a'\star b')\star N$.


    Since $N$ is normal, we may write
    \begin{align*}
    (a\star b) \star N &= a\star(b\star N) & &\text{(associativity)}\\
    &= a\star ( b'\star N ) & &\text{($b\star N = b'\star N$)} \\
    &= a\star (N\star b') & &\text{($N\normal G$)}\\
    &= (a\star N) \star b' & &\text{(associativity)}\\
    &= (a'\star N) \star b' & &\text{($a\star N = a'\star N$)} \\
    &= a' \star (N\star b') & &\text{(associativity)}\\
    &= a'\star (b'\star N) & &\text{($N\normal G$)}\\
    &= (a'\star b')\star N & &\text{(associativity)}.
    \end{align*}
    This is our desired result.
  \end{proof}
\end{lemma}

\begin{corollary}[Modular addition is well-defined]
  Addition on $\Z_n$ is well-defined.
\end{corollary}


\begin{corollary}[Modular multiplication is well-defined]
  Multiplication on $\Z_n$ is well-defined.
\end{corollary}


\begin{theorem}[Quotient groups]\label{T:quotient}\index{quotient group}
  Given a group $(G,\star)$ and a normal subgroup $N$, the set of left
  cosets of $N$ form a group under coset product. This group is
  denoted by $G/N$, and is pronounced ``$G$ modulo $N$.''
  \begin{proof}
    To start, note that the set of left cosets of $N$ is closed under
    the coset product. Next, note that the coset product is
    associative. Finally, the identity element is the coset $N$, and
    given $a\star N$, $a^{-1}\star N$ is the inverse. Hence $G/N$ is a
    group.
  \end{proof}
\end{theorem}

Let's catch our breath and talk about notation. If the group in
question is $(G,\star)$ with identity $e$ and $N\normal G$, then $G/N$
is a group with elements
\[
G/N=\{e\star N, a_1\star N, a_2 \star N, \dots\}
\]
where $a_i\in G$. In this case, the identity is $e\star N = N$ and the
inverse of $a\star N$ is $a^{-1}\star N$.



If the group in question is $(G,\cdot)$ with identity $1$ and
$N\normal G$, then $G/N$ is a group with elements
\[
G/N=\{1\cdot N, a_1\cdot N, a_2 \cdot N, \dots\}
\]
where $a_i\in G$. In this case, the identity is $1\cdot N = N$ and the
inverse of $a\cdot N$ is $a^{-1}\cdot N$.


Finally, if the group in question is $(G,+)$ with identity $0$ and
$N\normal G$, then $G/N$ is a group with elements
\[
G/N=\{0+N, a_1+ N, a_2 + N, \dots\}
\]
where $a_i\in G$. In this case, the identity is $0+ N = N$ and the
inverse of $a+ N$ is $-a +  N$.






\begin{exercise}
  Let $G$ be a group and $N\normal G$. Find the order of $G/N$.
\end{exercise}


\begin{exercise}
  Prove that every quotient group of a cyclic group is cyclic.
\end{exercise}


\begin{exercise}
  Write out the elements of $\Z/2\Z$ and use a Cayley table to show
  that this quotient group is isomorphic to the cyclic group $\Z_2$.
\end{exercise}


\begin{exercise}
  Write out the elements of $D_3/\<r\>$ and use a Cayley table to show
  that this quotient group is isomorphic to the cyclic group $\Z_2$.
\end{exercise}


\begin{exercise}
  Write out the elements of $\Z/3\Z$ and use a Cayley table to show
  that this quotient group is isomorphic to the cyclic group $\Z_3$.
\end{exercise}


\begin{exercise}
  Write out the elements of $\Z/5\Z$ and use a Cayley table to show
  that this quotient group is isomorphic to the cyclic group $\Z_5$.
\end{exercise}


\begin{example}[Solutions to trigonometric equations]
  Suppose you want to solve $\sin(\theta) = 1/2$. In this case,
  $\theta \in \R/2\pi\Z$.
\end{example}

\begin{exercise}
  Describe the quotient group $\Q/\Z$.
\end{exercise}

\begin{example}[Antiderivatives]
  Let $G= C^\infty(\R,\R)$ the set of continuous functions from $\R$
  to $\R$ whose derivatives are continuous. This set is a group under
  addition. Given a function $f\in C^\infty(\R,\R)$, the function
  \[
  \int f(x) \d x + C
  \]
  is an element of the quotient group $C^\infty(\R,\R)/\R$.
\end{example}
\end{document}
