\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{Subgroups}

\begin{document}
\begin{abstract}
  Subgroups are subsets of groups that are also groups. 
\end{abstract}
\maketitle


\begin{definition}
  Let $(G,\star)$ be a group, a subset $S\subset G$ is a \dfn{subgroup} if
  \begin{enumerate}
  \item $S \ne \emptyset$,
  \item $a,b\in S \Rightarrow a\star b\in S$,
  \item $a\in S \Rightarrow a^{-1}\in S$.
  \end{enumerate}
  We denote subgroups by writing $S\subgp G$.\index{S<G@$S\subgp G$}
\end{definition}


\begin{exercise}\index{nZ@$n\Z$}
  Let $n\Z = \{n\cdot x : x\in \Z\}$. Prove that $n\Z\subgp \Z$.
\end{exercise}


\begin{exercise}
  Let $(G,\star)$ be a group and $S$ be a subgroup of $G$. Prove that
  $(S,\star)$ is a group.
\end{exercise}

\begin{exercise}
  Prove that if $(G,\star)$ is a group and $S\subset G$ is a \textbf{finite}
  subset of $G$ such that
  \begin{enumerate}
  \item $S \ne \emptyset$,
  \item $a,b\in S \Rightarrow a\star b\in S$,
  \end{enumerate}
  then $S$ is a subgroup of $G$.
\end{exercise}


\begin{example}[Lattice diagram for $\boldsymbol{\pmb\Z_6}$]
  We can visualize the subgroup structure of a group with a lattice
  diagram. Here is the lattice diagram for $\Z_6$:
  \[
   \begin{tikzcd}
    & \Z_6 \ar[ld,-]  \ar[rdd,-] &       \\
    \<2\> \ar[rdd,-] &  &\\
   & &  \< 3\> \ar[ld,-]        \\   
    & \<0\> &
   \end{tikzcd}
   \]
  Typically we draw a lattice diagram with the largest subgroups at
  the top. The lines mean that there is containment between the groups.
\end{example}

\begin{exercise}
  Find the order of each subgroup of $\Z_6$.
\end{exercise}

\begin{example}[Lattice diagram for $\boldsymbol{\pmb\Z_8}$]
  Here is the subgroup lattice diagram for $\Z_8$:
  \[
  \begin{tikzcd}
    \Z_8  \ar[d,-] \\
    \<2\> \ar[d,-] \\
    \<4\> \ar[d,-] \\   
    \<0\> 
  \end{tikzcd}
  \]
\end{example}

\begin{exercise}
  Find the order of each subgroup of $\Z_8$.
\end{exercise}

\begin{exercise}
  Draw lattice diagrams for $\Z_n$ for $n=2,3\dots, 12$.
\end{exercise}

\begin{example}[Lattice diagram for $\boldsymbol{D_3}$]\index{symmetries!of the equilateral triangle}
  Here is the subgroup lattice diagram for $D_3$:
  \[
  \begin{tikzcd}
    & D_3 \ar[rdd,-]\ar[rrdd,-]\ar[rrrdd,-]  \ar[ld,-] & & &      \\
    \<r\> \ar[rdd,-]       &       &  & &  \\
    &       &  \< f\> \ar[ld,-]   &  \< rf\> \ar[lld,-]       &  \< r^2f\> \ar[llld,-]        \\   
    & \<e\> & & &
  \end{tikzcd}
  \]
\end{example}

\begin{exercise}
  Find the order of each subgroup of $D_3$.
\end{exercise}


\begin{example}[Lattice diagram for $\boldsymbol{D_4}$]
  Here is the subgroup lattice diagram for $D_4$:
  \[
  \begin{tikzcd}
    &    & D_4 \ar[rd,-] \ar[d,-] \ar[ld,-] &       \\
    &\<r^2,f\>  \ar[rd,-] \ar[d,-] \ar[ld,-]& \<r \>  \ar[d,-]     & \<r^2,rf\> \ar[ld,-] \ar[d,-] \ar[rd,-]\\
    \<f\>\ar[rrd,-]& \<r^2f\> \ar[rd,-]&\< r^2\> \ar[d,-]  &   \<rf\> \ar[ld,-]& \<r^3f\>\ar[lld,-]     \\   
&     & \<e\> &  &
  \end{tikzcd}
  \]  
\end{example}

\begin{exercise}
  Find the order of each subgroup of $D_4$.
\end{exercise}



\begin{example}[Lattice diagram for $\boldsymbol{Q_8}$]\index{quaternion group}
  Here is the subgroup lattice diagram for $Q_8$:
  \[
  \begin{tikzcd}
    & Q_8 \ar[rd,-] \ar[d,-] \ar[ld,-] &       \\
    \<i\>  \ar[rd,-] & \<j\>  \ar[d,-]     & \<k\> \ar[ld,-]\\
    &\< -1\> \ar[d,-]&        \\   
    & \<1\> &
  \end{tikzcd}
  \]
\end{example}


\begin{exercise}
  Find the order of each subgroup of $Q_8$.
\end{exercise}



\begin{example}[Nonisomorphic groups with similar lattice diagrams]
  Note, different groups may have the same lattice diagram structure.
  Again, consider the lattice for $\Z_6$:
  \[
  \begin{tikzcd}
    & \Z_6 \ar[rd,-]  \ar[ldd,-] &       \\
    &       & \<2\> \ar[ldd,-]\\
    \< 3\> \ar[rd,-]&       &       \\   
    & \<0\> &
  \end{tikzcd}
  \]
  Now check out the lattice diagram for $\Z_{15}$:
  \[
  \begin{tikzcd}
    & \Z_{15} \ar[rd,-]  \ar[ldd,-] &       \\
    &       & \<3\> \ar[ldd,-]\\
    \< 5\> \ar[rd,-]&       &       \\   
    & \<0\> &
  \end{tikzcd}
  \]
\end{example}






Let's show the lattice diagram for one more group.

\begin{example}[Lattice diagram for $\boldsymbol{A_4}$]\index{alternating group}
  Setting $\sigma = (1 \ 2 \ 3)$ and $\tau = (2 \ 3 \ 4)$, here is the
  subgroup lattice diagram for $A_4$:
  \[
  \begin{tikzpicture}
  \node[scale=.6] at (0,0) {\begin{tikzcd}
                   &   & & A_4   \ar[lld,-] \ar[rdd,-] \ar[rrdd,-]  \ar[rrrdd,-]  \ar[rrrrdd,-]   & &  & &        \\
                   & \<\sigma\tau, \sigma^2\tau^2\> \ar[ldd,-] \ar[dd,-] \ar[rdd,-] &  &              & &  & &  \\
                   &                                  &            &   & \<\sigma\> \ar[ldd,-]& \<\tau\> \ar[lldd,-] & \<\ \tau\sigma^2\> \ar[llldd,-]& \< \sigma \tau\sigma^2 \>\ar[lllldd,-]       \\
   \<\sigma\tau\>\ar[drrr,-]  & \<\sigma^2\tau^2\> \ar[drr,-] & \<\sigma^2\tau\sigma^2\> \ar[dr,-]& & &  & &    \\
                   &                               &   & \< e\>       & & & &  
  \end{tikzcd}};
  \end{tikzpicture}
  \]
\end{example}


\begin{exercise}
  Find the order of each subgroup of $A_4$.
\end{exercise}

\begin{exercise}
  Draw a subgroup lattice diagram for $\mathcal{T}_{12}$.
\end{exercise}


\begin{theorem}[Subgroup criterion]\label{T:sc}\index{subgroup criterion}
  Let $G$ be a group, a nonempty $S\subset G$ is a subgroup if and
  only if
  \[
  a,b\in S \quad \Rightarrow \quad ab^{-1}\in S.
  \]
  \begin{proof}
    $(\Rightarrow)$ Follows from the definition of a subgroup.

    $(\Leftarrow)$ Suppose for all $a,b\in S$, we know that
    $ab^{-1}\in S$. We must show that $ab\in S$ and that every element
    has an inverse.

    Suppose that $a\in S$. This means $aa^{-1} = e\in S$.  Now since
    $e,b\in S$, $eb^{-1}\in S$, so $b^{-1}\in S$.  Finally, since
    $a,b^{-1}\in S$, $ab\in S$. We see that $S\subgp G$.
  \end{proof}
\end{theorem}


\begin{exercise}
  Prove that $SO(n)$ is a subgroup of $O(n)$.
\end{exercise}


\begin{exercise}\label{E:center}
  Let $G$ be a group, define the \dfn{center} of $G$ as
  \[
  Z(G):=\{g\in G: \text{$gx=xg$ for all $x\in G$}\}.
  \]
  Prove that $Z(G)$ is a subgroup of $G$.
\end{exercise}


\begin{exercise}
  Let $G$ be a group, define the \dfn{normalizer} of $S$ in $G$ as
  \[
  N_G(S):=\{g\in G: \text{$gSg^{-1} = S$}\},
  \]
  where $gSg^{-1}:=\{g sg^{-1}:s\in S\}$. Prove that $N_G(S)$ is a
  subgroup of $G$.
\end{exercise}



\begin{exercise}
  Let $S$ and $T$ be subgroups of $(G,\star)$. Define
  \[
  S\star T := \{ s\star t: \text{$s\in S$ and $t\in T$}\}.
  \]
  Prove that $S\star T\subgp G$ if and only if $S\star T = T\star S$.
\end{exercise}



%% \begin{exercise}
%%   Consider the polynomial $x_1+x_2-x_3$. Prove that the set
%%   permutations on $\{x_1,x_2,x_3\}$ that fi
%% \end{exercise}



\begin{exercise}
  Let $H_1,H_2,\dots,H_n$ be subgroups of a group $G$. Prove that
  \[
  \bigcap_{i=1}^n H_i \subgp G.
  \]
\end{exercise}


%% This last exercise helps us reformulate what a generator is.

%% \begin{definition}
%%   Let $S$ be a subset of $G$. The group \dfn{generated} by $S$ is the
%%   smallest subgroup of $G$ that contains $S$. In particular
%%   \[
%%   \< S \> = \bigcap_{S\subset  H} H \subgp G.
%%   \]
%% \end{definition}


\begin{exercise}
  Prove that $A_n\subgp S_n$ is generated by all $3$-cycles in $S_n$.
  \begin{hint}
    Since the intersection of subgroups is a subgroup, and the
    subgroup generated by a set is the intersection of all subgroups,
    we just need to see the smallest set generated by all $3$-cycles. 
  \end{hint}
\end{exercise}

\begin{exercise}
  Find the subgroup of $\mathcal{T}_{12}$ generated by $\< rp,
  r^2p^2\>$. Make a Cayley table for this subgroup. Which common
  group is isomorphic to the subgroup $\< rp, r^2p^2\>$?
\end{exercise}





\section{Subgroups of cyclic groups}

\begin{theorem}[Classification of subgroups of the integers]
  The subgroups of $(\Z,+)$ are exactly $n\Z$ where $n = 0,1,\dots$.
  \begin{sketch}
    Use the well ordering principle to show that any subgroup $H\subgp
    G$ has a least positive element. Then use the division theorem.
  \end{sketch}
\end{theorem}

\begin{exercise}
  Describe the subgroup $n\Z\cap m\Z \subset \Z$ in terms of its
  cyclic generator.
\end{exercise}


\begin{exercise}
  Describe the subgroup of $\Z$ generated by $m$ and $n$ in terms of
  its cyclic generator.
\end{exercise}

\begin{exercise}
  Find a subgroup of $\Z\times\Z$ that is not of the form $m\Z \times
  n\Z$.
\end{exercise}


\begin{exercise}
  Find all subgroups of $\Z_9$, and identify each by a generator. Draw
  a lattice diagram to show off your work.
\end{exercise}


\begin{exercise}
  Find all subgroups of $\Z_{10}$, and identify each by a generator. Draw
  a lattice diagram to show off your work.
\end{exercise}

\begin{exercise}
  Find all subgroups of $\Z_{11}$, and identify each by a generator. Draw
  a lattice diagram to show off your work.
\end{exercise}

\begin{exercise}
  Find all subgroups of $\Z_{12}$, and identify each by a generator.  Draw
  a lattice diagram to show off your work.
\end{exercise}

  

%% \section{Back to the fifteen puzzle}



%% I feel that we would be remiss if our conversation on fifteen puzzles
%% didn't include an algorithm for solving them.


%% \begin{theorem}[Algorithm for solving fifteen puzzles]
%%   Algorithm For solving 15 puzzle
%%   %http://www.waynesthisandthat.com/15puzzle.htm

%%   Basic moves
%%   \[
%%   \arraycolsep=7pt\def\arraystretch{1.6}
%%   \begin{array}{!{\vline width 2pt}c|c|c|c!{\vline width 2pt}}
%%     \Xhline{2pt}
%%     1  & 2 & 3  & X_4 \\ \hline
%%     X_5  & X & X_7  & 4 \\ \hline
%%     X_9  & X_{10} & X_{11}  & \cellcolor{black} \\ \hline
%%     X  & X & X  & X \\
%%     \Xhline{2pt}
%%   \end{array}
%%   \]
%%   \[
%%   (7 \ 8)(8 \ 4)(4 \ 3)(3 \ 2) (2 \ 1)(1\ 5)(5\ 9)(9 \ 10)(10\ 11)(11 \ 12) 
%%   \left(\begin{smallmatrix}
%%     1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16\\
%%     1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 14 & 16
%%   \end{smallmatrix}\right)
%%   \]
  
%%   Basic strategy
%% \end{theorem}



%% \begin{exercise}
%%   Here is a devilish arrangement of a fifteen puzzle that I found in
%%   \textit{Selected topics in mathematics} by E.\ Spitznagel, Jr.
%%   \[
%%   \arraycolsep=7pt\def\arraystretch{1.6}
%%   \begin{array}{!{\vline width 2pt}c|c|c|c!{\vline width 2pt}}
%%     \Xhline{2pt}
%%     R  & A  & T  & E \\ \hline
%%     Y  & O  & U  & R \\ \hline
%%     M  & I & N & D \\ \hline
%%     P & A & L & \cellcolor{black}\\
%%     \Xhline{2pt}
%%   \end{array}
%%   \]
%% \end{exercise}





\end{document}
