\documentclass{ximera}

\author{Bart Snapp}

\input{../preamble.tex}

\title{Cyclotomic polynomials}

\begin{document}
\begin{abstract}
  We introduce cyclotomic, or ``circle-dividing'' polynomials.
\end{abstract}
\maketitle

\begin{theorem}[DeMoivre's Theorem]\index{DeMoivre's Theorem} 
Let $\alpha$ be any nonzero complex number. Then the equation
\[
x^n - \alpha = 0
\]
has exactly $n$ distinct roots, all of which are complex numbers.
\begin{proof} 
We will give a direct proof of DeMoivre's Theorem \textit{without}
appealing to the Fundamental Theorem of Algebra.
\begin{enumerate}
\item Prove that $e^{i \theta} = \cos(\theta) + i \sin(\theta)$.
\item Prove that any complex number $a + bi$ can be expressed as $R
  e^{i\theta}$ where $R\in \R$ and $\theta \in [0,2\pi)$.
\item Explain why if $\alpha = R(\cos(\theta) +i \sin(\theta))$, then 
\[
\sqrt[n]{R} \cdot  e^{\frac{i(\theta + 2k\pi)}{n}}
\]
is a root of $x^n - \alpha$.
\item Plot the roots on the unit circle, how many are there?
\end{enumerate}
Explain how we have proved the theorem.
\end{proof}
\end{theorem}


\begin{definition}\index{root of unity}\index{zetan@$\zeta_n$} 
For $n\in \N$, let 
\[
\zeta_n = e^{2\pi i/n}
\]
In this case we call $\zeta_n^k$ an \textbf{$\boldsymbol{n}$th-root of unity}, where $k\in
\N$.
\end{definition}

\begin{exercise} 
Prove that $\zeta_n^k\bar{\zeta_n^k} = 1$, where $\bar{\zeta_n^k}$ is
the complex conjugate of $\zeta_n^k$. Explain what this means
geometrically.
\end{exercise}

\begin{exercise}\label{E:KN} Prove that the set 
\[
K_n = \{\zeta_n^k = e^{2k\pi i/n}: k\in \N\}
\]
is a cyclic group under multiplication. Further prove that $K_n \iso
\Z_n$ as groups.
\end{exercise}



\begin{exercise}
If $n = 2m$, prove that 
\[
x^n -1 = (x-1)(x+1)q_1(x) \cdots q_{m-1}(x)
\]
where $q_i(x)$ are distinct irreducible polynomials in $\R[x]$. Hint, see (\ref{E:RQ}).
\end{exercise}

\begin{exercise}
If $n = 2m+1$, prove that 
\[
x^n -1 = (x-1)q_1(x) \cdots q_{m}(x)
\]
where $q_i(x)$ are distinct irreducible polynomials in $\R[x]$. Hint, see (\ref{E:RQ}).
\end{exercise}


\begin{definition}\index{primitive root of unity} 
A number $\zeta$ is called a \textbf{primitive} $n$th root of unity if
\[
\zeta^n = 1 \qquad\text{and}\qquad \zeta^{m} \ne 1
\]
for all $m< n$ where $m,n\in\N$.
\end{definition}


\begin{theorem} 
A number $\zeta_n^k$, where $1\le k\le n$, is a primitive $n$th root
of unity if and only if $(k,n) = 1$.

\begin{proof} 
$(\Rightarrow)$ Seek a contradiction supposing that $\zeta_n^k$ is a
  primitive $n$th root of unity and that $(k,n) \ne 1$.

$(\Leftarrow)$ Seek a contradiction supposing $(k,n) = 1$ and
  $\zeta_n^k$ is not a primitive $n$th root of unity. Use the fact
  that the order of $\zeta_n$ in $K_n$ is $n$ and Euclid's Lemma
  (\ref{E:EL2}).\index{Euclid's Lemma}
\end{proof}
\end{theorem}



\begin{exercise}\label{E:KP}
Prove that given a prime $p\nmid n$, the primitive $n$th roots of
unity are given by
\[
\{\zeta_n^{kp}: 1\le k \le n\}.
\]
\end{exercise}


\begin{exercise}\index{Euler phi-function@Euler $\phi$-function}\index{phi@$\phi$}
Prove that there are exactly $\phi(n)$ primitive roots of unity in
$\C$.
\end{exercise}



\begin{definition}\index{cyclotomic polynomial}
The $n$th \dfn{cyclotomic polynomial} is defined as:
\[
\Phi_n(x) = \prod_{\substack{(k,n)=1 \\ k<n}}(x-\zeta_n^k)
\]
\end{definition}

\begin{exercise}
  What is the degree of $\Phi_n(x)$?
\end{exercise}

\begin{exercise}
  Carefully explain each step shown below:
  \begin{align*}
    x^n - 1 &= \prod_{k=1}^n (x-\zeta_n^k) \\
    &=\prod_{d|n}\prod_{\substack{(k,d) = 1\\ \zeta_d^k \in K_n}}(x-\zeta_d^k) \\
    &=\prod_{d|n} \Phi_d(x).
  \end{align*}
\end{exercise}

\begin{corollary}
  The upshot of the above exercise is that 
  \[
  \Phi_n(x) = \frac{x^n-1}{\prod_{d|n} \Phi_d(x)}
  \]
\end{corollary}

\begin{exercise}
  Given a prime $p$, compute $\Phi_p(x)$. 
\end{exercise}

\begin{exercise}
  Compute $\Phi_n$ for $n = 1, \dots, 10$.
\end{exercise}

\begin{exercise}
  Prove that 
  \[
  n = \sum_{d|n} \phi(d).
  \]
\end{exercise}


\begin{theorem} 
  The $n$th cyclotomic polynomial $\Phi_n(x)$ is a monic polynomial in
  $\Z[x]$ for all $n\in \N$.
  
  \begin{proof} Proceed by induction on $n$.
    \begin{enumerate}
    \item Check the case when $n = 1$. 
    \item Now suppose our statement holds for all values less than $n$.
    \item Noting that 
      \[
      x^n - 1 = \prod_{d|n}\Phi_d(x) = \Phi_n(x)\cdot \prod_{\substack{d|n\\ d \ne n}}\Phi_d(x).
      \]
      Hence $\prod_{\substack{d|n\\ d \ne n}}\Phi_d(x)$ divides $x^n-1$ in
      $\Q(\zeta_n)[x]$. Explain why this is also true in $\Q[x]$. Big hint,
      see the Division Theorem for polynomials, see
      (\ref{T:divThmPoly}). Use this theorem in $\Q(\zeta_n)[x]$ and $\Q[x]$
      utilizing uniqueness!
    \end{enumerate}
    Explain how to finish the proof. Hint, use (\ref{D:cont}).
  \end{proof}
\end{theorem}

\begin{theorem} 
  The $n$th cyclotomic polynomial $\Phi_n(x)$ is irreducible in $\Q[x]$
  for all $n\in \N$.
  \begin{proof} 
    Seeking a contradiction, suppose that $\Phi_n(x)$ is not irreducible.
    \begin{enumerate}
    \item Explain why we may assume that $\Phi_n(x)$ is reducible in
      $\Z[x]$.
    \item Explain why we may assume that $\Phi_n(x) = f(x)\cdot g(x)$
      where $f(x)$ and $g(x)$ are monic polynomials in $\Z[x]$ with
      \[
      \deg(f) < \phi(n) \qquad \text{and}\qquad \deg(g) < \phi(n).
      \]
      Hint, use (\ref{D:cont}).
\item Explain why we may assume that $f$ is a minimal polynomial for
  $\zeta_n$.
    \end{enumerate}
    Now we claim that given a prime $p \nmid n$, $\zeta_n^p$ is also a
    root of $f(x)$. Further seek another contradiction, and suppose that
    $\zeta_n^p$ is not a root of $f(x)$.
    \begin{enumerate}
    \item Explain why $\zeta_n^p$ is a root of $g(x)$.
    \item Explain why $f(x) | g(x^p)$ and write $g(x^p) = f(x) \cdot
      h(x)$, where $h(x)\in \Q[x]$.
    \item Explain why we may conclude that $h(x)$ is a monic polynomial
      $\Z[x]$, use (\ref{D:cont}).
    \item Reduce the equation $g(x^p) = f(x) \cdot h(x)$ modulo $p$,
      writing the image of the polynomials $f$, $g$, and $h$ as $\bar{f}$,
      $\bar{g}$, and $\bar{h}$:
      \[
      \bar{g}(x^p) = \bar{f}(x) \cdot \bar{h}(x)
      \]\index{Freshman Binomial Theorem}\index{Fermat's Little Theorem}
    \item Explain how to use the Freshman Binomial Theorem (\ref{E:FBT}),
      or Fermat's Little Theorem (\ref{T:FLT}), to conclude that
      \[
      \bar{g}(x)^p = \bar{f}(x) \cdot \bar{h}(x)
      \]
    \item Explain how the fact that $\Z_p[x]$ is a UFD allows us to
      conclude that $\bar{f}$ and $\bar{g}$ have a common factor.
    \item Explain why we must conclude that $x^n-1$ has a multiple root
      modulo $p$. Further explain why $\bar{f}$ and $\bar{g}$ both have
  $x$ as a factor, see (\ref{E:mr}). Explain why this is a
      contradiction.
    \end{enumerate}
    Thus we conclude that $\zeta^p_n$ is a root of $f(x)$. Since this
    argument will work for any prime $p$ with $p\nmid n$, we can show that
    every primitive $n$th root of unity is a root of $f$, see (\ref{E:KP}). This will
    show that $\Phi_n(x) = f$, and is hence irreducible.
  \end{proof}
\end{theorem}




\end{document}
