\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{Fields}

\begin{document}
\begin{abstract}
  We introduce fields.
\end{abstract}
\maketitle

In this section we will introduce fields, which are objects that are a
bit like two groups at once.


\begin{definition}
  A \dfn{field} is a set $K$ with two operations that we will denote
  by $+$ and $\cdot$, and call addition and multiplication, such that
  the following properties hold:
  \begin{description}
  \item[Abelian group under addition] $(K,+)$ is an Abelian
    group with identity element denoted by $0$.
  \item[Abelian group under multiplication] $(K-\{0\},\cdot)$ is
    an Abelian group with identity element denoted by $1$.
  \item[Distributivity] The operation $\cdot$ distributes over $+$,
    meaning that if $a,b,c\in K$,
    \[
    a\cdot (b+ c) = a\cdot b+ a\cdot c \quad\text{and}\quad (a+ b)\cdot c  = a\cdot c+ b\cdot c.
    \]
  \end{description}
\end{definition}

\begin{exercise} 
  Which of the following are fields under the canonical operations?
  Select all that apply.
  \begin{selectAll}
    \choice[correct]{$\Q$}
    \choice[correct]{$\R$}
    \choice[correct]{$\C$}
    \choice{$\Z$}
    \choice[correct]{$\Z_p$ where $p$ is a prime number}
    \choice{$\Z_m^\times$}
  \end{selectAll}
\end{exercise}




\begin{exercise}
  Note, in the definition of a field, the elements ``$0$'' and ``$1$''
  are not necessarily the integers who look and sound the same. In
  particular, nowhere in the definition of a field does it state what
  how $0$ acts on field elements by field multiplication. Prove or
  disprove: $a\cdot 0 = 0$.
\end{exercise}

\begin{exercise}
  Let $K$ be a field. Prove that $(-1)\cdot a = (-a)$.
\end{exercise}





\begin{exercise}
  Let $K$ be a field. Prove that ${\left(ab^{-1}\right)}\hspace{0pt}^{-1} = a^{-1}
  b$. Note, here you are using the definition of a field to show
  \[
  \frac{1}{a/b} = \frac{b}{a}.
  \]
\end{exercise}


\begin{exercise} 
  Let $a$ and $b$ be elements of a field $K$. Prove that if $a\cdot b
  = 0$, then $a$ is zero or $b$ is zero. 
\end{exercise}




\begin{exercise} 
  When is $\Z_n$ definitely not a field?
\end{exercise}

\begin{exercise}\label{E:FF4}
  Describe a field with four elements.
  \begin{hint}
    Trial and error with a Cayley table for addition and a Cayley
    table for multiplication is not a bad way to start.
  \end{hint}
\end{exercise}


\begin{exercise}\label{E:FF6} Can you find a field with six elements?
\end{exercise}


\begin{exercise}
  Let $K$ be a field. Prove that $K\times K$ is never a field under
  componenetwise addition and componenetwise multiplication.
\end{exercise}


\begin{exercise} 
Let $K$ be a field. Find some quality inherent in $K$ that is
necessary and sufficient to make the set of ordered pairs in $K\times
K$ a field when ``addition'' is defined by $\diamond$ and
``multiplication'' is defined by $\star$:
\[
(a,b)\diamond (c,d) = (a+c, b+d)\qquad\text{and}\qquad (a,b)\star (c,d) =
(ac-bd,ad+bc).
\]
Hint: What if $K = \R$? What if $K = \C$? Keep on going!
\end{exercise}



\section{Homomorphisms of fields}


\begin{definition}\index{homomorphism!of fields}\index{field homomorphism}\index{isomorphism!of fields}\index{field isomorphism}
  Let $K$ and $L$ be fields. A function
  \[
  \phi:K\to L
  \]
  is a \dfn{homomorphism} of fields if
  \begin{align*}
    \phi(a+b) &= \phi(a)+\phi(b)\\
    \phi(a\cdot b) &= \phi(a)\cdot \phi(b)\\
    \phi(1_K) &= 1_L,
  \end{align*}
  where the operations on the left-hand side of the equations might
  not be the operations on right-hand side of the equations. A
  bijective homomorphism of fields is called a \dfn{field isomorphism}.
\end{definition}

\begin{exercise}
  Let $\phi:K\to L$ be a homomorphism of fields. Prove that $\phi(0_K)
  = 0_L$. Can you prove that $\phi(1_K) = 1_L$ strictly from the first
  two equations that define a field homomorphism? Why or why not?
\end{exercise}


\begin{definition}\index{kernel!field homomorphism}
  Let $\phi: K\to L$ be a homomorphism of fields. The \textbf{kernel}
  of $\phi$ is
  \[
  \ker(\phi) := \{a\in K:\phi(a) = 0_L\}\subset K.
  \]
  The \textbf{image} of $\phi$ is\index{image!field homomorphism}
  \[
  \im(\phi) = \{\phi(a):a\in K\}\subset L.
  \]
\end{definition}


\begin{exercise}
  Let $\phi:K\to L$ be a homomorphism of fields. Prove that $\phi$ is
  injective if and only if $\ker(\phi) = \{0\}$.
\end{exercise}

\begin{exercise}
  Let $\phi:K\to L$ be a homomorphism of fields. Prove that $\phi$ is
  injective.
\end{exercise}

\begin{exercise}
  Prove that every surjective homomorphism of fields is a bijective
  homomorphism of fields.
\end{exercise}







\section{The characteristic of a field}


Fields can be dichotomized by their \textit{characteristic}.

\begin{definition}
The \dfn{characteristic} of a field $K$ is the least positive
integer $n$ such that
\[
\underbrace{1 + 1 + \cdots +1}_{\text{$n$ times}} =0.
\]
If there is no such positive integer, then we say that $K$ has
characteristic $0$.
\end{definition}


\begin{exercise}
  Prove that if a field has characteristic $0$, then it contains a
  subgroup isomorphic to $\Z$.
\end{exercise}

\begin{exercise}
  Prove that if a field has characteristic $p$, then it contains a
  subgroup isomorphic to $\Z_p$.
\end{exercise}



\begin{exercise} 
Prove that if a field has finite characteristic, then its
characteristic is a prime number.
\end{exercise}

\begin{exercise}\index{freshman binomial theorem}\label{E:FBT}
Prove that in a field of positive characteristic $p$, the so-called
\textit{freshman binomial theorem} holds:
\[
(a+b)^p = a^p + b^p
\]
\end{exercise}



When working in fields of characteristic $p$, we have a new
homomorphism.

\begin{example}[The Frobenius homomorphism]\index{Frobenius homomorphism}
  Let $K$ be a field of characteristic $p$. Prove that
  \begin{align*}
    \phi:K &\to K\\
    a &\mapsto a^p
  \end{align*}
  is a homomorphism of fields. Is the Frobenius homomorphism surjective?
\end{example}





This simple idea of characteristic tells us something about fields.

\begin{definition}
  A $K$ is a \dfn{subfield} of a field $K$ if
  \begin{enumerate}
  \item $k\subset K$,
  \item $k$ has the same identity elements and operations as $K$.
  \end{enumerate}
\end{definition}


\begin{theorem}[Base fields of positive characteristic]
  If a field $K$ has positive characteristic, then $K$ contains a
  subfield $Z$ such that $Z\iso \Z_p$.
  \begin{sketch}
    Set
    \[
    Z =\left\{\sum_{i=1}^n 1:n\in\{1,2,\dots,p\}\right\}.
    \]
    Prove that $Z$ is a group of order $p$, and hence is isomorphic to
    $\Z_p$. Since $\Z_p$ is also a field, we are done.
  \end{sketch}
\end{theorem}





\begin{theorem}[Base fields of characteristic zero]
  If a field $K$ has positive characteristic, then $K$ contains a
  subfield $Q$ such that $Q\iso \Q$.
  \begin{sketch}
    Set
    \[
    Z =\left\{\sum_{i=1}^n \pm 1:n\in \N\cup\{0\}\right\}.
    \]
    Prove that $(Z,+)$ is an infinite cyclic group and hence is
    isomorphic to $(\Z,+)$.

    Now set
    \[
    Q = \{a^{-1}b : a,b\in Z\}.
    \]
    We claim that $Q$ is a field. First note that $Q$ is a subgroup of
    $K$ under field addition. Suppose $\alpha,\beta\in Q$. We'll
    show that $\alpha-\beta\in Q$. Let $\alpha = a^{-1}b$ and $\beta=
    c^{-1}d$. Write
    \[
    a^{-1}b - c^{-1} d = (ac)^{-1}(bc- ad)\in Q.
    \]
    Now we must show that $Q-\{0\}$ is a subgroup of $K$ under
    multiplication. Again let $\alpha = a^{-1}b$ and $\beta=
    c^{-1}d$. We'll show that $\alpha^{-1}\beta \in Q$. Write
    \[
    a^{-1}b (c^{-1}d)^{-1} = (ad)^{-1} bc\in Q.
    \]
    Since $Q\subset K$, multiplication distributes over addition.

    Now we will prove that $\Q\iso Q$. Define a map
    \begin{align*}
      \Q &\to Q\\
      b/a &\mapsto a^{-1} b.
    \end{align*}
    We must show that this map is \index{well-defined}well-defined.
    Suppose that
    \[
    \frac{b}{a} = \frac{d}{c},
    \]
    this happens if and only if
    \[
    ad = bc\quad\Leftrightarrow a^{-1} b = c^{-1} d.
    \]
    Thus our map is well-defined. Since our map is clearly a
    surjective homomorphism fields, it is an isomorphism of fields.
  \end{sketch}
\end{theorem}

\begin{corollary}[Number fields]
  Every subfield of $\C$ contains $\Q$. Subfields of $\C$ are often
  called \dfn{number fields}.
\end{corollary}












\end{document}
