\documentclass{ximera}

\input{../../preamble.tex}

\author{Bart Snapp}

\title{Irreducible polynomials}

\begin{document}
\begin{abstract}
  We discuss how to identify irreducible polynomials.
\end{abstract}
\maketitle

From our previous work, we know that $K[x]/(n(x))$ is a field if and
only if $n(x)$ is an irreducible polynomial. Let's figure out some
strategies to determine if a polynomial is irreducible. In particular,
we are interested in showing that polynomials in $\Q[x]$ are
irreducible.

\section{Degree two or three}

\begin{lemma}[Irreducibility for degree two and three]\label{L:d23}
  Let $K$ be a field and $f(x)\in K[x]$ where $2 \le\deg(f(x))\le 3$. In
  this case, $f(x)$ is irreducible if and only if $f(x)$ has no roots
  in $K$.
  \begin{sketch}
    If $2 \le \deg(f(x))\le 3$ and $f(x)$ factors, then it must have a
    factor of degree $1$.
  \end{sketch}
\end{lemma}






\section{Changing coefficients}

So far, we've been thinking about polynomials whose coefficients are
field elements. This doesn't have to be the case. Define
\[
\Z[x] = \{\text{polynomials in terms of $x$ with coefficients in
  $\Z$}\}.
\]

When the polynomial has integer coefficients, there is an easy way to
find rational roots.

\begin{lemma}[Rational roots test]\label{L:rr}
  Consider $f(x) \in \Z[x]$,
  \[
  f(x) = c_nx^n + c_{n-1}x^{n-1} + \cdots + c_1 x + c_0.
  \]
  If $\frac{r}{s}\in\Q$ is in lowest terms and $\frac{r}{s}$ is a root
  of $f(x)$, then
  \[
  r | c_0 \quad\text{and}\quad s | c_n.
  \]
  \begin{sketch}
    Evaluate $f\left(\frac{r}{s}\right)$ and clear denominators.
  \end{sketch}
\end{lemma}

Even if the degree of the polynomial you are working with is higher
than two or three, the rational roots test allows you to easily find
linear factors.


\begin{lemma}[Reduction to characteristic $\boldsymbol p$]\label{L:RCP}
  Given a polynomial in $\Z_p[x]$, if $f(x)$ is irreducible in
  $\Z_p[x]$ for some prime $p$, then $f(x)$ is irreducible in $\Z[x]$.
  \begin{sketch}
    Prove the contrapositive of the statement.
  \end{sketch}
\end{lemma}


\begin{lemma}[Gauss' lemma]\index{Gauss' lemma}\label{L:G}
  Given a polynomial $f(x)\in\Z[x]$, if $f(x)$ is irreducible in
  $\Z[x]$, then $f(x)$ is irreducible in $\Q[x]$.
  \begin{proof}
    We'll prove the contrapositive of this statement. Suppose $f(x) =
    g(x)\cdot h(x)$ where $g(x),h(x)\in\Q[x]$. Let $d$ be a common
    denominator for the coefficients of $g(x)$ and $h(x)$. Hence
    \[
    d\cdot f(x) = g'(x)\cdot h'(x)
    \]
    where $g'(x)\in \Z[x]$ and $h'(x) \in\Z[x]$. We may factor
    \[
    d = p_1,\dots, p_n.
    \]
    Since $p_1$ divides $d$, $p_1$ divides the coefficients of
    $g'(x)\cdot h'(x)$.


    We claim that $p_1$ divides all the coefficients of $g'(x)$ or
    $p_1$ divides all the coefficients of $h'(x)$.  Let $a_i$
    represent the coefficients of $g'(x)$ and $b_j$ represent the
    coefficients of $h'(x)$. Seeking a contradiction, let:
    \begin{itemize}
      \item $a_n$ be the first coefficient of $g'(x)$ that is not
        divisible by $p_1$,
      \item $b_m$ be the first coefficient of $h'(x)$ that is not
        divisible by $p_1$.
    \end{itemize}
    Writing out the coefficient of $x_{m+n}$ of $d\cdot f(x)$, we have
    \[
    \underbrace{a_0b_{m+n}  + \cdots + a_{n-1}b_{m+1}}_{\text{divisible by $p_1$}} + a_nb_m + \underbrace{a_{n+1}b_{m-1} + \cdots + a_{m+n} b_0.}_{\text{divisible by $p_1$}}
    \]
    Since this entire expression is divisible by $p_1$, we conclude
    that $a_nb_m$ is divisible by $p_1$. Now by Euclid's lemma,
    Corollary~\ref{C:EL2}, this means $p_1|a_n$ or $p_1|b_m$. Hence $p_1$
    divides all the coefficients of $g'(x)$ or $p_1$ divides all the
    coefficients of $h'(x)$. This means that
    \[
    g'(x)/p \in \Z[x] \quad\text{or}\quad h'(x)/p \in \Z[x]
    \]
    Repeat for every prime factor of $d$ to see that $f(x)$ factors in
    $\Z[x]$.
  \end{proof}
\end{lemma}


Now we have the following:
\[
\text{Irreducible in $\Z_p[x]$} \Rightarrow \text{Irreducible in $\Z[x]$}  \Rightarrow \text{Irreducible in $\Q[x]$}
\]


\section{Eisenstein's criterion}



\begin{theorem}[Eisenstein's criterion]\index{Eisenstein's criterion}\label{T:ec}
  Given a polynomial $f(x) \in\Z[x]$
  \[
  f(x) = c_nx^n + c_{n-1}x^{n-1} + \cdots + c_1 x + c_0,
  \]
  if there exists a prime $p$ such that
  \begin{enumerate}
  \item $p|c_i$ for all $i \ne n$,
  \item $p\nmid c_n$, and
  \item $p^2\nmid c_0$,
  \end{enumerate}
  then $f(x)$ is irreducible in $\Z[x]$.
  \begin{proof}
    Seeking a contradiction, suppose that $f(x)$ factors in $\Z[x]$
    and there exists a prime $p$ such that
    \begin{enumerate}
    \item $p|c_i$ for $i \ne n$,
    \item $p\nmid c_n$,
  \item $p^2\nmid c_0$.
    \end{enumerate}
    Write
    \[
    f(x) = g(x) \cdot h(x)
    \]
    where
    \begin{align*}
      g(x) &= a_rx^r + \cdots + a_1 x+ a_0\\
      h(x) &= b_sx^s + \cdots + b_1 x+ b_0,
    \end{align*}
    and $r + s = n = \deg(f(x))$. From our work above we know that:
    \[
    \text{Factors in $\Z[x]$} \Rightarrow \text{Factors in $\Z_p[x]$} 
    \]
    This means that if we reduce the \textbf{coefficients} of the
    polynomials $f$, $g$, and $h$ modulo $p$ and write
    \begin{align*}
      \bar{f} \equiv f \ \text{in $\Z_p[x]$},\\
      \bar{g} \equiv g \ \text{in $\Z_p[x]$},\\
      \bar{h} \equiv h \ \text{in $\Z_p[x]$}.
    \end{align*}
    By our setup
    \[
    \bar{f}(x) =  \bar{a}_r\bar{b}_sx^{r+s} \in \Z_p[x].
    \]
    Now we have an equation in $\Z_p[x]$,
    \[
    \bar{a}_r\bar{b}_sx^{r+s}=  \bar{g}(x)\cdot \bar{h}(x).
    \]
    Polynomials are equal as elements if and only if their
    coefficients are equal. On the left-hand side above, we have a
    polynomial whose only nonzero coefficient is at the $(r+s)$-degree
    term. Since we know $\Z_p$ is a field, and $\Z_p[x]$ has unique
    factorization, Corollary~\ref{C:UFP}, we deduce that neither
    $\bar{g}$ nor $\bar{h}$, can have a linear term (a constant
    term). This means that $p$ divides the constant term of both $g$
    and $h$. But this means that $p^2$ divides the constant term of
    $f$, a contradiction.
  \end{proof}
\end{theorem}

Some folks say that a polynomial is \textbf{Eisenstein} if it
satisfies the conditions of Eisenstein's criterion. Eisenstein's
criterion is often combined with the following exercise:
\begin{exercise}\label{E:et}
  Prove that if $f(ax+b)$, where $a\ne 0$ and $b\ne 0$ is irreducible,
  then $f(x)$ is irreducible.
\end{exercise}
So now, if you want to show a polynomial $f$ is irreducible, one
strategy is to see if $f(ax+b)$ is Eisenstein for some values of $a$
and $b$. Usually you try $a=b=1$.


\begin{exercise}
  Prove that $x^4 + 10x + 5$ is irreducible in $\Q[x]$.
\end{exercise}



\begin{exercise}
  Prove that $x^4 + 1$ is irreducible in $\Q[x]$.
  \begin{hint}
    Set $x = 1+y$.
  \end{hint}
\end{exercise}


\begin{exercise}
  Determine whether the following polynomials are irreducible. For those
  are are reducible, give their factorizations into irreducible
  polynomials. For those that are irreducible, give a proof explaining
  your conclusion.
  \begin{enumerate}
  \item $x^2 + x +1$ in $\Z_2[x]$.
  \item $x^3 + x + 1$ in $\Z_3[x]$.
  \item $x^4+1$ in $\Z_5[x]$.
  \item $x^2 + x + 4$ in $\Z_{11}[x]$.
  \item $x^5 + 9x^4 + 12x^2 + 6$ in  $\Q[x]$.
  \item $17 x^3 - 7x^2 + 34x + 1$ in $\Q[x]$.
  \item $x^4 + 10x + 1$ in $\Q[x]$.
  \item $x^4 + 10x^2 + 1$ in $\Q[x]$. This is known as a
    \dfn{Swinnerton-Dyer polynomial}.
  \item $4x^3 -3x -1/2$ in $\Q[x]$.\index{4x^3-3x-1/2@$4x^3-3x-1/2$}
  \end{enumerate}
  \begin{hint}
    Note, each of these may require different techniques. Some will
    probably require `brute-force.'
  \end{hint}
\end{exercise}





For some interesting extra reading check out:
\begin{itemize}
\item \link[\textit{Counting irreducible polynomials over finite fields using the inclusion-exclusion principle},
  S.\ Chebolu and J.\ Min\'a\v c]{https://doi.org/10.4169/math.mag.84.5.369}.

\end{itemize}



\end{document}
