\documentclass{ximera}

\input{../preamble.tex}

\title{The method of Mr.\ Gauss}

\author{Bart Snapp}

\begin{document}
\begin{abstract}
  We prove that certain regular polygons are constructible.
\end{abstract}
\maketitle

When Gauss was around $16$ years old, he put his massive brain to the
question of which regular $n$-gons are constructible with compass and
straightedge.


We too shall put our massive brains to this task. Key to understanding
this will be the maxim that we first encountered so long ago:
\begin{quote}
  To understand an object in mathematics, you should understand the
  functions on the object.
\end{quote}
Recall that $\aut(X)$\index{automorphism} is the set of bijective maps
from $X$ to $X$. When $G$ is a group, we \textit{\bfseries insist}
that $\aut(G)$ is the set of group isomorphisms from $G$ to $G$. When
$K$ is a field, we \textit{\textbf{insist}} that $\aut(K)$ we
\textit{\textbf{insist}} that $\aut(K)$ is the set of field
isomorphisms from $K$ to $K$. Of critical importance, in each case
above: $\aut(X)$, $\aut(G)$, and $\aut(K)$ form groups.



As it turns out, the group $\aut(K)$ will be ``too large'' for our
purposes. Instead, we want a set of field isomorphisms that
\textbf{fix} a ground field.


\begin{definition}\index{automorphism}\index{field automorphism}
  Let $K\subset L$ be a field extension and define the
  \dfn{automorphism group}\textbf{ of $\boldsymbol L$ over
    $\boldsymbol K$}, denoted
  \[
  \aut_K(L)\index{autKL@$\aut_K(L)$}
  \]
  to be the set of field automorphisms of $L$ that fix elements of
  $K$. This means for all $\sigma \in \aut_K(L)$:
  \begin{itemize}
  \item $\sigma: L\to L$ is an isomorphism of fields.
  \item If $a\in K$, then $\sigma(a) = a$.
  \item If $\zeta\in (L - K)$, then $\sigma(\zeta) \in (L-K)$.
  \end{itemize}
  Sometimes people write $\aut(L/K)$\index{aut(LK)@$\aut(L/K)$} for
  this as well.
\end{definition}

\begin{example}[Complex conjugation]
  Let $\phi:\C\to\C$ be the homomorphism of fields defined by
  \begin{align*}
    \phi : \C &\to \C\\
    a+bi &\mapsto a - bi.
  \end{align*}
  In this case, $\phi \in \aut_\R(\C)$, as $\phi$ is the identity on
  $\R$.
\end{example}


\begin{exercise}
  Prove that $\aut_K(L)$ forms a group under composition.
\end{exercise}


At this point, it is common to introduce new notation for a specific
instance of $\aut_K(L)$. We don't really need this notation, but we would
be remiss if we did not mention it.


Given a tower of field extensions $K\subset L$, we call $\aut_K(L)$
the \dfn{Galois group of $\boldsymbol L$ over $\boldsymbol K$},
denoted
  \[
  \gal_K(L)\index{galKL@$\gal_K(L)$} \quad\text{or}\quad\gal_K(f)
  \]
  if for some degree $n$ polynomial $f(x)\in K[x]$ with roots $\alpha_1,\dots, \alpha_n$:
  \begin{itemize}
  \item The roots of $f$ are distinct, if $i\ne j$ then, $\alpha_i \ne
    \alpha_j$.
  \item $L = K[\alpha_1,\dots,\alpha_n]$.
  \end{itemize}
It might seem strange to introduce new notation for something that
could be expressed in old notation. This author feels your pain. This
notation is often used when $L$ is not just any field extension of
$K$, but one with this special context; the field extensions we are
interested in come from finding minimal field extensions where a given
polynomial factors completely into \textit{distinct} linear factors.
This notation allows us to ``pinpoint'' the field extension where a
given polynomial ``splits'' into \textbf{distinct} linear factors.

Finally note, in the definition of $\gal_K(f)$, at no point did we
insist that $f$ be irreducible, in fact, finding Galois groups of
(nonirreducible) polynomials is an interesting problem.


Galois groups are named after the mathematician \link[\'Evariste
  Galois]{https://en.wikipedia.org/wiki/Evariste\_Galois} (pronounced
\textit{gal-wah}). He lived a short and tragic life. He was shot dead
in a meaningless duel at the age of 20.

In retrospect, this was an act of suicide on Galois part. The genius
of Galois' work was overlooked and unnoticed in his life.  Much of this
was simply due to bad luck. Galois was lost in a world that did not
seem to value him.  It's sad that when we feel hopeless our brains try
to kill us with our thoughts and actions; especially when there are
\textbf{always other options}\link[]{https://ccs.osu.edu/}. Had Galois
made it just a little further in his life, I think he would have seen
many changes for the better. Now back to math!




%% \begin{example}[Galois group of a quadratic extension]
%%   If $\Q(\alpha)$ is a degree two extension of $\Q$, then
%%   \[
%%   \aut_\Q(\Q(\alpha)) \iso \Z_2^\times \iso {e}
%%   \]
%%   \begin{proof}
%%     We already know that $\Z_2^\times\iso {e}$. We will show that
%%     \[
%%     \aut_{\Q}(\Q(\alpha))\iso\Z_2^\times.
%%     \]
%%     Let
%%     \[
%%     ax^2 +bx + c
%%     \]
%%     be an irreducible quadratic polynomial, meaning $a \ne 0$, in
%%     $\Q[x]$. It's roots are
%%     \begin{align*}
%%     \alpha &= \frac{-b+\sqrt{b^2-4ac}}{2a} \\
%%     \beta &= \frac{-b-\sqrt{b^2-4ac}}{2a}.
%%     \end{align*}
%%     As an aside, note that
%%     \[
%%     \Q(\alpha) = \Q(\beta) 
%%     \]
%%     as
%%     \[
%%     \beta = \frac{-b-(b+2a\alpha)}{2}\in\Q(\alpha).
%%     \]
%%     Now by Lemma~\ref{L:fevs} and Theorem~\ref{T:dae}, $\Q(\alpha)$ is a
%%     $\Q$-vector space with basis
%%     \[
%%     \{1,\alpha\}.
%%     \]
%%     However, 
%%     \[
%%     \{\alpha,\alpha^2\}
%%     \]
%%     is also a basis, as $\dim_\Q(\Q(\alpha))=2$ and there are two
%%     linearly independent elements, behold
%%     \begin{align*}
%%       c_0 \alpha + c_1 \alpha^2 &= 0\\
%%       \alpha(c_0 + c_1 \alpha) &= 0.
%%     \end{align*}
%%     Since $\alpha\ne 0$, we see
%%     \[
%%     c_0 + c_1 \alpha = 0,
%%     \]
%%     and so $c_0 = c_1 = 0$.  This means that every element of
%%     $\Q(\alpha)$ can be expressed uniquely as
%%     \[
%%     c_0\alpha + c_1 \alpha^2 
%%     \]
%%     where $c_i\in\Q$. If $\sigma\in\aut_\Q(\Q(\alpha))$, then
%%     \begin{align*}
%%       \sigma(c_0 + c_1 \alpha) &= \sigma(c_0) + \sigma(c_1 \alpha)\\
%%       &= c_0 + c_1 \sigma(\alpha)\\
%%       &= c_0 + c_1 \sigma(\alpha).
%%     \end{align*}
%%     This means that an automorphisms of fields $\sigma \in
%%     \aut_\Q(\Q(\alpha))$ is uniquely determined by where it maps
%%     $\alpha$. In this case, there  are two choices
%%     \begin{align*}
%%       e :\alpha &\mapsto \alpha &  \sigma :\alpha &\mapsto \alpha^2\\
%%          \alpha^2 &\mapsto \alpha^2, &          \alpha^2 &\mapsto \alpha.
%%     \end{align*}
%%     Since $\aut_\Q(\Q(\alpha)) = \{e,\sigma\}$ as defined above, we
%%     see $\aut_\Q(\Q(\alpha))\iso\Z_3^\times$.
%%   \end{proof}
%% \end{example}
  
Finding Galois groups (automorphism groups) in general is hard,
because they depend not only on the degree of the minimal polynomial,
but also on the explicit coefficients of the polynomial. However,
looking for Galois groups of $n$th roots of unity is a much easier
problem.

\begin{example}[$\boldsymbol{\aut_{\pmb \Q}\pmb(\pmb\Q\pmb(\zeta_3\pmb)\pmb)}$]
  Let $\zeta_3$ be a primitive third-root of
  unity. Now we have an isomorphism of groups:
  \[
  \aut_{\Q}(\Q(\zeta_3))\iso\Z_3^\times\iso \Z_2
  \]
  \begin{proof}
    We already know that $\Z_3^\times\iso \Z_2$ by
    Theorem~\ref{T:mgc}. We will show that
    \[
    \aut_{\Q}(\Q(\zeta_3))\iso\Z_3^\times.
    \]
    First note $\Q(\zeta_3)$ is an algebraic extension of degree $2$
    since $\zeta_3$ is a root of
    \[
    \Phi_3(x) = x^2 + x + 1. 
    \]
    When we substitute $x = y+ 1$, we find
    \[
    \Phi_3(x) = y^2 + 3y + 3
    \]
    which is irreducible by Eisenstein's criterion,
    Theorem~\ref{T:ec}. Thus $\Phi_3(x)$ is the minimal polynomial for
    $\zeta_3$. By Lemma~\ref{L:fevs} and Theorem~\ref{T:dae},
    $\Q(\zeta_3)$ is a $\Q$-vector space with basis
    \[
    \{1,\zeta_3\}.
    \]
    However, 
    \[
    \{\zeta_3,\zeta_3^2\}
    \]
    is also a basis, as $\dim_\Q(\Q(\zeta_3))=2$ and this is a minimal
    spanning set since
    \[
    \zeta_3 + \zeta_3^2 = -1.
    \]
    This means that every element of $\Q(\zeta_3)$ can be expressed
    uniquely as
    \[
    c_0\zeta_3 + c_1 \zeta_3^2
    \]
    where $c_i\in\Q$. If $\sigma\in\aut_\Q(\Q(\zeta_3))$, then
    \begin{align*}
      \sigma(c_0\zeta_3 + c_1 \zeta_3^2) &= \sigma(c_0\zeta_3) + \sigma(c_1 \zeta_3^2)\\
      &= c_0 \sigma(\zeta_3) + c_1 \sigma(\zeta_3^2)\\
      &= c_0 \sigma(\zeta_3) + c_1 \sigma(\zeta_3)^2.
    \end{align*}
    This means that an automorphisms of fields $\sigma \in
    \aut_\Q(\Q(\zeta_3))$ is uniquely determined by where it maps
    $\zeta_3$. In this case, there are two choices
    \begin{align*}
      e :\zeta_3 &\mapsto \zeta_3 &  \sigma :\zeta_3 &\mapsto \zeta_3^2\\
         \zeta_3^2 &\mapsto \zeta_3^2, &          \zeta_3^2 &\mapsto \zeta_3.
    \end{align*}
    Since $\aut_\Q(\Q(\zeta_3)) = \{e,\sigma\}$ as defined above, we
    write the following isomorphism of groups
    \begin{align*}
      \phi: \aut_\Q(\Q(\zeta_3)) &\to \Z_3^\times\\
      e &\mapsto 1\\
      \sigma &\mapsto 2.
    \end{align*}
    Hence, $\aut_\Q(\Q(\zeta_3))\iso\Z_3^\times$.
  \end{proof}
\end{example}


\begin{example}[$\boldsymbol{\aut_{\pmb\Q}\pmb(\pmb\Q\pmb(\zeta_5\pmb)\pmb)}$]
  Let $\zeta_5$ be a primitive fifth-root of
  unity. Now we have an isomorphism of groups:
  \[
  \aut_{\Q}(\Q(\zeta_5))\iso\Z_5^\times\iso \Z_4
  \]
  \begin{proof}
    We already know that $\Z_5^\times\iso \Z_4$ by
    Theorem~\ref{T:mgc}. We will show that
    \[
    \aut_{\Q}(\Q(\zeta_5))\iso\Z_5^\times.
    \]
    First note $\Q(\zeta_5)$ is an algebraic extension of degree $4$
    since the minimal polynomial for $\zeta_5$ is
    \[
    \Phi_5(x) = x^4+x^3+x^2 + x + 1.
    \]
    When we substitute $x = y+ 1$, we find
    \[
    \Phi_5(x) = y^{4} +5y^3+10y^2+10y+5
    \]
    which is irreducible by Eisenstein's criterion,
    Theorem~\ref{T:ec}. Thus $\Phi_5(x)$ is the minimal polynomial for
    $\zeta_5$.


    By Lemma~\ref{L:fevs} and Theorem~\ref{T:dae},
    $\Q(\zeta_5)$ is a $\Q$-vector space with basis
    \[
    \{1,\zeta_5,\zeta_5^2,\zeta_5^3\}.
    \]
    However,
    \[
    \{\zeta_5,\zeta_5^2,\zeta_5^3,\zeta_5^4\}
    \]
    is also a basis, as $\dim_\Q(\Q(\zeta_5))=4$ and this is a minimal
    spanning set since
    \[
    \zeta_5 + \zeta_5^2  + \zeta_5^3 + \zeta_5^4 = -1.
    \]
    This means that every element of $\Q(\zeta_5)$ can be expressed
    uniquely as
    \[
    c_0\zeta_5 + c_1\zeta_5^2  + c_2\zeta_5^3 + c_3\zeta_5^4 
    \]
    where $c_i\in\Q$. If $\sigma\in\aut_\Q(\Q(\zeta_5))$, then
    \begin{align*}
      \sigma(c_0\zeta_5 + c_1\zeta_5^2  + c_2\zeta_5^3 + c_3\zeta_5^4) &= \sigma(c_0\zeta_5) + \sigma(c_1 \zeta_5^2)+ \sigma(c_2 \zeta_5^3)+ \sigma(c_3 \zeta_5^4)\\
      &= c_0 \sigma(\zeta_5) + c_1 \sigma(\zeta_5^2) + c_2 \sigma(\zeta_5^3) + c_3 \sigma(\zeta_5^4)\\
      &= c_0 \sigma(\zeta_5) + c_1 \sigma(\zeta_5)^2 + c_2 \sigma(\zeta_5)^3 + c_3 \sigma(\zeta_5)^4.
    \end{align*}
    This means that an automorphisms of fields $\sigma \in
    \aut_\Q(\Q(\zeta_5))$ is uniquely determined by where it maps
    $\zeta_5$. In this case, there are four choices
    \begin{align*}
      e :\zeta_5 &\mapsto \zeta_5 &  \sigma :\zeta_5 &\mapsto \zeta_5^2\\
      \sigma^2 :\zeta_5 &\mapsto \zeta_5^4 &  \sigma^3 :\zeta_5 &\mapsto \zeta_5^3.
    \end{align*}
    Since
    \[
    \aut_\Q(\Q(\zeta_5)) = \{e,\sigma,\sigma^2,\sigma^3,\sigma^4\}
    \]
    as defined above, we
    write the following isomorphism of groups
    \begin{align*}
      \phi: \aut_\Q(\Q(\zeta_5)) &\to \Z_5^\times\\
      e &\mapsto 1\\
      \sigma &\mapsto 2\\
      \sigma^2 &\mapsto 4\\
      \sigma^3 &\mapsto 3.\\
    \end{align*}
    Hence, $\aut_\Q(\Q(\zeta_5))\iso\Z_5^\times$.
  \end{proof}
\end{example}

\begin{exercise}
  Let $p$ be a prime number. Prove that $\aut_\Q(\Q(\zeta_p))\iso
  \Z_p^\times$.
\end{exercise}

\begin{exercise}\index{automorphism}
  If $G$ is a group, define $\aut(G)$ to be the set of group
  automorphisms (bijective group homomorphisms from $G$ to
  itself). Prove that
  \[
  \aut(\Z_p) \iso \Z_p^\times.
  \]
\end{exercise}



Now it is time for a dirty lemma that we will use over and over again.


\begin{lemma}[Fixed elements are in the ground field]\label{L:fgf}
  Suppose $K\subset L$ is an algebraic extension of degree $n$
  with a basis
  \[
  \{\beta_1,\beta_2,\beta_3,\cdots,\beta_{n-1},\beta_n\}
  \]
  for $L$ as a $K$-vector space and
  \[
  \beta_1 + \beta_2 + \dots + \beta_n \in K.
  \]
  Additionally, suppose that $\aut_{K}(L)$ is cyclic on
  $\{\beta_1,\dots,\beta_n\}$, and that $\sigma\in\aut_{K}(L)$ is a
  generator. If $\alpha\in L$ with $\sigma(\alpha) = \alpha$, then
  $\alpha\in K$.
  \begin{proof}
    Since
    \[
    \{\beta_1,\beta_2,\beta_3,\cdots,\beta_{n-1},\beta_n\}
    \]
    is a basis for $K(\beta_1,\dots,\beta_n)$ as a $K$-vector space, we may  write
    \[
    \alpha = c_1 \beta_1 + c_2 \beta_2 + \cdots + c_n\beta_{n},
    \]
    where each $c_i\in K$. Applying $\sigma$ we find
    \[
    \sigma(\alpha) = c_1 \sigma(\beta_1) + c_2 \sigma(\beta_2) + \cdots + c_n\sigma(\beta_{n-1}).
    \]
    However, since the $\beta_i$ form a basis and $\sigma$ is a
    generator of the cyclic group $\aut_{K}(L)$, $\sigma$ permutes the
    basis vectors full-cycle, and hence
    \[
    c_1 = c_2 = \cdots = c_{n}\in K.
    \]
    Setting $c= c_1$, write
    \[
    \alpha = c \cdot (\beta_1 + \beta_2 + \cdots + \beta_{n}) \in K.
    \]
    This completes the proof.
  \end{proof}
\end{lemma}





\begin{example}[Regular $\boldsymbol{5}$-gon is constructible]
  We'll prove that $5$ equally spaced points on the unit circle are
  constructible 
  \[
  \begin{tikzpicture}  
    \begin{axis}[  
        xmin=-1.2,  
        xmax=1.2,  
        ymin=-1.2,  
        ymax=1.2,  
        axis lines=center,
        ticks=none,
        xlabel=$\Re$,  
        ylabel=$\Im$,  
        every axis y label/.style={at=(current axis.above origin),anchor=south},  
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
        axis equal image,clip mode=individual
      ]
      \draw[dashed] (axis cs: 0,0) circle[radius=.935in];
      
      \draw[fill=black] (axis cs: {1},{0}) circle[radius=.03in];
      \node[anchor=south west] at (axis cs:{1},{0}) {\scriptsize$1$};
      
      \draw[fill=black] (axis cs: {cos(360/5)},{sin(360/5)}) circle[radius=.03in];
      \node[anchor=south west] at (axis cs:{cos(360/5)},{sin(360/5)}) {\scriptsize$\zeta_5$};

      \draw[fill=black] (axis cs: {cos(2*360/5)},{sin(2*360/5)}) circle[radius=.03in];
      \node[anchor=east] at (axis cs:{cos(2*360/5)},{sin(2*360/5)}) {\scriptsize$\zeta_5^2$};

      \draw[fill=black] (axis cs: {cos(3*360/5)},{sin(3*360/5)}) circle[radius=.03in];
      \node[anchor=east] at (axis cs:{cos(3*360/5)},{sin(3*360/5)}) {\scriptsize$\zeta_5^3$};

      \draw[fill=black] (axis cs: {cos(4*360/5)},{sin(4*360/5)}) circle[radius=.03in];
      \node[anchor=north west] at (axis cs:{cos(4*360/5)},{sin(4*360/5)}) {\scriptsize$\zeta_5^4$};
    \end{axis}
  \end{tikzpicture}  
  \]
  by proving that the $5$th roots of unity are constructible.  We
  start with the polynomial
  \begin{align*}
    x^{5} -1 &= (x-1) \Phi_{5}(x)\\
    &= (x-1)(x^{4} + x^{3} + x^2 + x+1).
  \end{align*}
  By substituting $x=y+1$ into $x^{4} + x^3 +x^2 + x+1$, we find
  \[
  y^{4} +5y^3+10y^2+10y+5
  \]
  Since this polynomial satisfies Eisenstein's criterion,
  Theorem~\ref{T:ec}, we see $\Phi_5(x)$ is irreducible in $\Z[x]$. By
  Gauss' lemma, Lemma~\ref{L:G}, we see $\Phi_5(x)$ is irreducible in
  $\Q[x]$. Thus by Theorem~\ref{T:dae}, 
  \[
  \begin{tikzcd}
    \Q(\zeta_{5})\\
    \Q\ar[u, -, "4"]
  \end{tikzcd}
  \]
  hence we see that the regular $5$-gon may be construtible, but to
  conclude this, we must show that we can find a tower of degree two
  extensions, starting with $\Q$ and finishing with $\Q(\zeta_{5})$.
  Consider $\aut_{\Q}(\Q(\zeta_{5}))$. In this case
  \[
  \aut_{\Q}(\Q(\zeta_{5}))\iso \Z_{5}^\times
  \]
  and is cyclic. There is no easy method other than trial-and-error to
  find the generator of this group, as this is the discrete log
  problem.\index{discrete log problem} However, though
  trial-and-error, we find the following generator:
  \begin{align*}
    \sigma:\Q(\zeta_{5}) &\to\Q(\zeta_{5})\\
    \zeta_{5} &\mapsto \zeta_{5}^2.
  \end{align*}
  We may now use this cyclic generator to produce subgroups of
  $\aut_{\Q}(\Q(\zeta_{5})) = \< \sigma \>$. Let's draw a subgroup lattice:
  \[
    \begin{tikzcd}
    \Z_{4}\iso\aut_\Q(\Q(\zeta_{5})) = \< \sigma \>\ar[d,-]\\
    \< \sigma^2\> \ar[d,-]\\
    \<\sigma^{4}=e\> 
  \end{tikzcd}
  \]
  Let's examine $\<\sigma^2\>\subgp \aut_\Q(\Q(\zeta_{5}))$,
  \[
  \<\sigma^2\> = \{ \sigma^2, \sigma^{4} = e\}.
  \]
  Each of these maps permute powers of $\zeta_{5}$ and fix elements
  of $\Q$. Quite explicitly,
  \begin{align*}
    e: \zeta_5 &\mapsto \zeta_5\\
    \sigma^2: \zeta_{5} &\mapsto \zeta_{5}^4
  \end{align*}
  There are two field isomorphisms above, and hence there are two more
  field isomorphisms in $\aut_\Q(\Q(\zeta_{5})) - \< \sigma^2\>$, and
  these are all different since the cosets of a subgroup partition the
  group, Theorem~\ref{T:CPG}. Thus if we set
  \begin{align*}
    \alpha &= \zeta_{5}   + \zeta_{5}^4,\\
    \beta &= \zeta_{5}^2 + \zeta_{5}^{3},
  \end{align*}
  the elements $\{\alpha,\beta\}$ form a basis for $\Q[\alpha,\beta]$
  over $\Q$, as $\alpha$ and $\beta$ are linearly independent and they
  span $\Q[\alpha,\beta]$. In this case, $\sigma(\alpha) = \beta$ and
  $\sigma(\beta) = \alpha$. This means
  \begin{align*}
    \sigma(\alpha+\beta) &= \alpha + \beta,\\
    \sigma(\alpha\beta) &= \alpha\beta,
  \end{align*}
  which by Lemma~\ref{L:fgf} shows that
  $\alpha+\beta,\alpha\beta\in\Q$ and so we see $\Q[\alpha,\beta]
  \subset \Q(\alpha)$. However, $\alpha$ and $\beta$ are both roots of
  the polynomial
  \[
  x^2 - (\alpha +\beta) x + \alpha\beta\in\Q[x]
  \]
  and so we see $\Q[\alpha,\beta]= \Q(\alpha)$ and $[\Q(\alpha):\Q] =
  2$:
  \[
  \begin{tikzcd}
    \Q(\alpha) \\ \Q \ar[u,-,"2"]
  \end{tikzcd}
  \]
  Since we know that $[\Q(\zeta_{5}):\Q] = 4$, and
  \[
  \Q(\alpha)\subset \Q(\zeta_{5})
  \]
  with $[\Q(\alpha):\Q]=2$, by Lemma~\ref{L:dm},
  \[
    [\Q(\zeta_{5}):\Q(\alpha)]=2.
  \]
  Since we have a tower of field extensions, each of degree two over
  the previous field, we see that the $5$-gon is constructible by
  compass and straightedge.
\end{example}

\begin{example}[Regular $\boldsymbol{17}$-gon is constructible]
  We'll prove that $17$ equally spaced points on the unit circle are
  constructible 
  \[
  \begin{tikzpicture}  
    \begin{axis}[  
        xmin=-1.2,  
        xmax=1.2,  
        ymin=-1.2,  
        ymax=1.2,  
        axis lines=center,
        ticks=none,
        xlabel=$\Re$,  
        ylabel=$\Im$,  
        every axis y label/.style={at=(current axis.above origin),anchor=south},  
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
        axis equal image,clip mode=individual
      ]
      \draw[dashed] (axis cs: 0,0) circle[radius=.935in];
      
      \draw[fill=black] (axis cs: {1},{0}) circle[radius=.03in];
      \node[anchor=south west] at (axis cs:{1},{0}) {\scriptsize$1$};
      
      \draw[fill=black] (axis cs: {cos(1*360/17)},{sin(1*360/17)}) circle[radius=.03in];
      \node[anchor=south west] at (axis cs:{cos(1*360/17)},{sin(1*360/17)}) {\scriptsize$\zeta_{17}$};

      \draw[fill=black] (axis cs: {cos(2*360/17)},{sin(2*360/17)}) circle[radius=.03in];
      \node[anchor=south west] at (axis cs:{cos(2*360/17)},{sin(2*360/17)}) {\scriptsize$\zeta_{17}^2$};

      \draw[fill=black] (axis cs: {cos(3*360/17)},{sin(3*360/17)}) circle[radius=.03in];
      \node[anchor=south west] at (axis cs:{cos(3*360/17)},{sin(3*360/17)}) {\scriptsize$\zeta_{17}^3$};

      \draw[fill=black] (axis cs: {cos(4*360/17)},{sin(4*360/17)}) circle[radius=.03in];
      \node[anchor=south west] at (axis cs:{cos(4*360/17)},{sin(4*360/17)}) {\scriptsize$\zeta_{17}^4$};

      \draw[fill=black] (axis cs: {cos(5*360/17)},{sin(5*360/17)}) circle[radius=.03in];
      \node[anchor=south east] at (axis cs:{cos(5*360/17)},{sin(5*360/17)}) {\scriptsize$\zeta_{17}^5$};

      \draw[fill=black] (axis cs: {cos(6*360/17)},{sin(6*360/17)}) circle[radius=.03in];
      \node[anchor=south east] at (axis cs:{cos(6*360/17)},{sin(6*360/17)}) {\scriptsize$\zeta_{17}^6$};

      \draw[fill=black] (axis cs: {cos(7*360/17)},{sin(7*360/17)}) circle[radius=.03in];
      \node[anchor=south east] at (axis cs:{cos(7*360/17)},{sin(7*360/17)}) {\scriptsize$\zeta_{17}^7$};

      \draw[fill=black] (axis cs: {cos(8*360/17)},{sin(8*360/17)}) circle[radius=.03in];
      \node[anchor=south east] at (axis cs:{cos(8*360/17)},{sin(8*360/17)}) {\scriptsize$\zeta_{17}^8$};

      \draw[fill=black] (axis cs: {cos(9*360/17)},{sin(9*360/17)}) circle[radius=.03in];
      \node[anchor=north east] at (axis cs:{cos(9*360/17)},{sin(9*360/17)}) {\scriptsize$\zeta_{17}^9$};

      \draw[fill=black] (axis cs: {cos(10*360/17)},{sin(10*360/17)}) circle[radius=.03in];
      \node[anchor=north east] at (axis cs:{cos(10*360/17)},{sin(10*360/17)}) {\scriptsize$\zeta_{17}^{10}$};

      \draw[fill=black] (axis cs: {cos(11*360/17)},{sin(11*360/17)}) circle[radius=.03in];
      \node[anchor=north east] at (axis cs:{cos(11*360/17)},{sin(11*360/17)}) {\scriptsize$\zeta_{17}^{11}$};

      \draw[fill=black] (axis cs: {cos(12*360/17)},{sin(12*360/17)}) circle[radius=.03in];
      \node[anchor=north east] at (axis cs:{cos(12*360/17)},{sin(12*360/17)}) {\scriptsize$\zeta_{17}^{12}$};

      \draw[fill=black] (axis cs: {cos(13*360/17)},{sin(13*360/17)}) circle[radius=.03in];
      \node[anchor=north west] at (axis cs:{cos(13*360/17)},{sin(13*360/17)}) {\scriptsize$\zeta_{17}^{13}$};

      \draw[fill=black] (axis cs: {cos(14*360/17)},{sin(14*360/17)}) circle[radius=.03in];
      \node[anchor=north west] at (axis cs:{cos(14*360/17)},{sin(14*360/17)}) {\scriptsize$\zeta_{17}^{14}$};

      \draw[fill=black] (axis cs: {cos(15*360/17)},{sin(15*360/17)}) circle[radius=.03in];
      \node[anchor=north west] at (axis cs:{cos(15*360/17)},{sin(15*360/17)}) {\scriptsize$\zeta_{17}^{15}$};

      \draw[fill=black] (axis cs: {cos(16*360/17)},{sin(16*360/17)}) circle[radius=.03in];
      \node[anchor=north west] at (axis cs:{cos(16*360/17)},{sin(16*360/17)}) {\scriptsize$\zeta_{17}^{16}$};
      
    \end{axis}
  \end{tikzpicture}  
  \]
  by proving that the $17$th roots of unity are constructible.
  We start with the polynomial
  \begin{align*}
    x^{17} -1 &= (x-1) \Phi_{17}(x)\\
    &= (x-1)(x^{16} + x^{15} + \cdots + x+1).
  \end{align*}
  By substituting $x=y+1$ into $x^{16} + x^{15} + \cdots + x+1$, we find
  \begin{align*}
  y^{16} &+17 y^{15}+136 y^{14}+680 y^{13}+2380 y^{12} \\
  &+6188 y^{11}+12376 y^{10}+19448 y^9+24310 y^8+24310 y^7\\
  &+19448 y^6+12376 y^5+6188 y^4+2380 y^3+680 y^2+136 y+17.
  \end{align*}
  Since this polynomial satisfies Eisenstein's criterion,
  Theorem~\ref{T:ec}, we see $\Phi_{17}(x)$ is irreducible in
  $\Z[x]$. By Gauss' lemma, Lemma~\ref{L:G}, we see that $\Phi_{17}(x)$ is irreducible in
  $\Q[x]$. Thus by Theorem~\ref{T:dae}, 
  \[
  \begin{tikzcd}
    \Q(\zeta_{17})\\
    \Q\ar[u, -, "16"]
  \end{tikzcd}
  \]
  hence we see that the regular $17$-gon may be construtible, but to
  conclude this, we must show that we can find a tower of degree two
  extensions, starting with $\Q$ and finishing with $\Q(\zeta_{17})$.
  Consider $\aut_{\Q}(\Q(\zeta_{17}))$. In this case
  \[
  \aut_{\Q}(\Q(\zeta_{17}))\iso \Z_{17}^\times
  \]
  and is cyclic. There is no easy method other than trial-and-error to
  find the generator of this group as this is the discrete log
  problem\index{discrete log problem}. Nevertheless we, like the
  heros before us, persist and find a generator
  \begin{align*}
  \sigma:\Q(\zeta_{17}) &\to \Q(\zeta_{17})\\
  \zeta_{17} &\mapsto \zeta_{17}^3.
  \end{align*}
  Using the cyclic generator $\sigma$, we can produce subgroups of
  $\aut_\Q(\Q(\zeta_{17})) = \< \sigma \>$. Let's draw a subgroup lattice:
  lattice:
  \[
    \begin{tikzcd}
    \Z_{16}\iso\aut_\Q(\Q(\zeta_{17})) = \< \sigma \>\ar[d,-]\\
    \< \sigma^2\> \ar[d,-]\\
    \<\sigma^4\> \ar[d,-]\\
    \<\sigma^8\> \ar[d,-]\\
    \<\sigma^{16}=e\> 
  \end{tikzcd}
  \]
  Let's examine $\<\sigma^2\>\subgp \aut_\Q(\Q(\zeta_{17}))$,
  \[
  \<\sigma^2\> = \{ \sigma^2, \sigma^4,\sigma^6,\sigma^8,\sigma^{10},\sigma^{12},\sigma^{14},\sigma^{16} = e\}.
  \]
  Each of these maps permute powers of $\zeta_{17}$ and fix elements
  of $\Q$. Quite explicitly,
  \begin{align*}
    \sigma^2: \zeta_{17} &\mapsto \zeta_{17}^9    & \sigma^{10}: \zeta_{17} &\mapsto \zeta_{17}^{8}\\
    \sigma^4: \zeta_{17} &\mapsto \zeta_{17}^{13} & \sigma^{12}: \zeta_{17} &\mapsto \zeta_{17}^{4}\\
    \sigma^6: \zeta_{17} &\mapsto \zeta_{17}^{15} & \sigma^{14}: \zeta_{17} &\mapsto \zeta_{17}^{2}\\
    \sigma^8: \zeta_{17} &\mapsto \zeta_{17}^{16} & \sigma^{16}: \zeta_{17} &\mapsto \zeta_{17}.
  \end{align*}
  There are eight field isomorphisms above, and hence there are eight
  more field isomorphisms in $\aut_\Q(\Q(\zeta_{17})) - \<
  \sigma^2\>$, and these are all different since the cosets of a
  subgroup partition the group, Theorem~\ref{T:CPG}. Thus if we set 
  \begin{align*}
    \alpha &= \zeta_{17}   + \zeta_{17}^9 + \zeta_{17}^{13} + \zeta_{17}^{15} + \zeta_{17}^{16} + \zeta_{17}^{8} + \zeta_{17}^{4} + \zeta_{17}^{2},\\
    \beta &= \zeta_{17}^3 + \zeta_{17}^{10} + \zeta_{17}^{5} + \zeta_{17}^{11} + \zeta_{17}^{14} + \zeta_{17}^{7} + \zeta_{17}^{12} + \zeta_{17}^{6},
  \end{align*} 
  the elements $\{\alpha,\beta\}$ form a basis for $\Q[\alpha,\beta]$
  over $\Q$, as $\alpha$ and $\beta$ are linearly independent and they
  span $\Q[\alpha,\beta]$. Moreover, by construction, $\sigma(\alpha)
  = \beta$ and $\sigma(\beta) = \alpha$. This means
  \begin{align*}
    \sigma(\alpha+\beta) &= \alpha + \beta,\\
    \sigma(\alpha\beta) &= \alpha\beta,
  \end{align*}
  which by Lemma~\ref{L:fgf} shows that
  $\alpha+\beta,\alpha\beta\in\Q$ and so we see
  $\Q[\alpha,\beta]\subset \Q(\alpha)$. However, $\alpha$ and $\beta$
  are both roots of the polynomial
  \[
  x^2 - (\alpha +\beta) x + \alpha\beta\in\Q[x]
  \]
  and so we see $\Q[\alpha,\beta]= \Q(\alpha)$ and $[\Q(\alpha):\Q] =
  2$:
  \[
  \begin{tikzcd}
    \Q(\alpha) \\ \Q \ar[u,-,"2"]
  \end{tikzcd}
  \]
  Now look at $\<\sigma^4\>\subgp \< \sigma^2\>$,
  \[
  \<\sigma^4\> = \{ \sigma^4, \sigma^8,\sigma^{12},\sigma^{14},\sigma^{16} = e\}.
  \]
  Let's look at these field isomorphisms:
  \begin{align*}
    \sigma^4: \zeta_{17} &\mapsto \zeta_{17}^{13}    & \sigma^{12}: \zeta_{17} &\mapsto \zeta_{17}^{4}\\
    \sigma^8: \zeta_{17} &\mapsto \zeta_{17}^{16} & \sigma^{16}: \zeta_{17} &\mapsto \zeta_{17}.
  \end{align*}
  There are four field isomorphisms above, and hence there are four
  other field isomorphisms in $\<\sigma^2\> - \< \sigma^4\>$. Since 
  these are all different since the cosets of a subgroup partition the
  group, Theorem~\ref{T:CPG}. Thus if we set
  \begin{align*}
    \gamma &= \zeta_{17}   + \zeta_{17}^{13} + \zeta_{17}^{16} + \zeta_{17}^{4},\\
    \delta &= \zeta_{17}^9 + \zeta_{17}^{15} + \zeta_{17}^{8} + \zeta_{17}^{2},
  \end{align*}
  the elements $\{\gamma,\delta\}$ form a basis for
  $\Q[\alpha,\gamma,\delta]$ over $\Q(\alpha)$ as $\gamma$ and
  $\delta$ are linearly independent, and they span
  $\Q[\alpha,\gamma,\delta]$.  In this case, $\sigma^2(\gamma)= \delta$ and
  $\sigma^2(\delta) = \gamma$. This means
  \begin{align*}
    \sigma^2(\gamma+\delta) &= \gamma + \delta,\\
    \sigma^2(\gamma\delta) &= \gamma\delta,
  \end{align*}
  which by Lemma~\ref{L:fgf} shows that
  $\gamma+\delta,\gamma\delta\in\Q(\alpha)$. Again, both $\gamma$ and
  $\delta$ are roots of the polynomial
  \[
  x^2 - (\gamma +\delta) x + \gamma\delta\in\Q(\alpha)[x]
  \]
  and so we see $\Q[\alpha,\gamma,\delta] = \Q(\alpha,\gamma)$ and
  $[\Q(\alpha,\gamma):\Q(\alpha)] = 2$:
  \[
  \begin{tikzcd}
    \Q(\alpha,\gamma)\\
    \Q(\alpha) \ar[u,-,"2"]\\
    \Q \ar[u,-,"2"]
  \end{tikzcd}
  \]
  Finally look at $\<\sigma^8\>\subgp \<\sigma^4\>$,
  \[
  \<\sigma^8\> = \{ \sigma^8,\sigma^{16} = e\}.
  \]
  There are two field isomorphisms above, and hence there are two
  other field isomorphisms in $\<\sigma^4\> - \< \sigma^8\>$, and
  these are all different since the cosets of a subgroup partition the
  group, Theorem~\ref{T:CPG}. Thus if we set
  \begin{align*}
    \epsilon &= \zeta_{17} + \zeta_{17}^{16},\\
    \eta &= \zeta_{17}^{13} + \zeta_{17}^4.
  \end{align*}
  The elements $\{\epsilon,\eta\}$ form a basis for
  $\Q[\alpha,\gamma,\epsilon,\eta]$ over $\Q(\alpha,\gamma)$ as
  $\epsilon$ and $\eta$ are linearly independent, and they span
  $\Q[\alpha,\gamma,\epsilon,\eta]$.  In this case,
  $\sigma^4(\epsilon) = \eta$ and $\sigma^4(\eta) = \epsilon$. This
  means that
  \begin{align*}
    \sigma^4(\epsilon+\eta) &= \epsilon + \eta\\
    \sigma^4(\epsilon\eta) &= \epsilon\eta,
  \end{align*}
  which by Lemma~\ref{L:fgf} shows that
  $\epsilon+\eta,\epsilon\eta\in\Q(\alpha,\gamma)$. Again, both
  $\epsilon$ and $\eta$ are roots of the polynomial
  \[
  x^2 - (\epsilon +\eta) x + \epsilon\eta\in\Q(\alpha,\gamma)[x]
  \]
  and so we see $\Q[\alpha,\gamma,\epsilon,\eta]=
  \Q(\alpha,\gamma,\epsilon)$ and
  $[\Q(\alpha,\gamma,\epsilon):\Q(\alpha,\gamma)] = 2$:
  \[
  \begin{tikzcd}
    \Q(\alpha,\gamma,\epsilon)\\
    \Q(\alpha,\gamma)\ar[u,-,"2"]\\
    \Q(\alpha) \ar[u,-,"2"]\\
    \Q \ar[u,-,"2"]
  \end{tikzcd}
  \]
  Since we know that $[\Q(\zeta_{17}):\Q] = 16$, and
  \[
  \Q(\alpha,\gamma,\epsilon)\subset \Q(\zeta_{17})
  \]
  with $[\Q(\alpha,\gamma,\epsilon):\Q]=8$, by Lemma~\ref{L:dm},
  \[
    [\Q(\zeta_{17}):\Q(\alpha,\gamma,\epsilon)]=2.
  \]
  Since we have a tower of field extensions, each of degree two over
  the previous field, we see that the $17$-gon is constructible by
  compass and straightedge.
\end{example}



\begin{theorem}[Constructible regular $\boldsymbol p$-gons]
  If $p$ is a prime number, it is possible to construct a regular
  $p$-gon if $p= 2^n-1$ for some $n\in \N$.
  \begin{proof}
    Suppose you wish to construct a regular $p$-gon where $p$ is a
    prime number. This is equivalent to constructing the roots of
    $x^p-1$. Write:
    \begin{align*}
      x^{p} -1 &= (x-1) \Phi_{p}(x)\\
      &= (x-1)(x^{p-1} + x^{p-2} + \cdots + x+1).
    \end{align*}
    The polynomail $\Phi_{p}(x)$ can be shown to be irrducible by
    setting $x = y+1$ and using Eisenstein's criterion,
    Theorem~\ref{T:ec}, we see $\Phi_{p}(x)$ is irreducible in
    $\Z[x]$. By Gauss' lemma, Lemma~\ref{L:G}, see $\Phi_{p}(x)$ is
    irreducible in $\Q[x]$. Thus by Theorem~\ref{T:dae}, 
    \[
    \begin{tikzcd}
      \Q(\zeta_{p})\\
      \Q\ar[u, -, "p-1"]
    \end{tikzcd}
    \]
    Hence the regular $p$-gon is constructible only if $p-1 = 2^n$
    where $n\in \N$, by Theorem~\ref{T:dc}. However, this is not
    sufficient for a regular $p$-gon to be constructible. We must show
    that there is a tower of field extensions 
    \[
    \begin{tikzcd}
      K_n=\Q(\zeta_{p})\\
      K_{n-1}\ar[u, -, "2"]\\
      \vdots\ar[u, -, "2"]\\
      K_1\ar[u, -, "2"]\\
      K_0=\Q\ar[u, -, "2"]
    \end{tikzcd}
    \]
    where at each step, $[K_i:K_{i+1}]=2$. To do this, we will look at
    $\aut_\Q(\Q(\zeta_p))$. In this case
    \[
    \aut_\Q(\Q(\zeta_p))\iso\Z_p^\times \iso \Z_{p-1} = \Z_{2^n}
    \]
    with the right-most isomorphism holding by
    Corollary~\ref{C:mzpc}. Let $\sigma$ be a generator for the cyclic
    group $\aut_\Q(\Q(\zeta_p))$. In this case, $\sigma$ cyclicly permutes the set
    \[
    \{\zeta_p,\zeta_p^2,\dots,\zeta_p^{p-1}\},
    \]
    and $\sigma^2$ permutes exactly half of this set. Thus we can set
    $\alpha$ to be the sum of the $\zeta_p^i$ that are permuted by
    $\sigma^2$ and $\beta$ to be the sum of the $\zeta_p^i$ that are
    not permuted by $\sigma^2$.  Note that the elements
    $\{\alpha,\beta\}$ form a basis for $\Q[\alpha,\beta]$ over $\Q$,
    as $\alpha$ and $\beta$ are linearly independent and they span
    $\Q[\alpha,\beta]$. Moreover, by construction, $\sigma(\alpha) =
    \beta$ and $\sigma(\beta) = \alpha$. This means
    \begin{align*}
      \sigma(\alpha+\beta) &= \alpha + \beta,\\
      \sigma(\alpha\beta) &= \alpha\beta,
    \end{align*}
    which by Lemma~\ref{L:fgf} shows that
    $\alpha+\beta,\alpha\beta\in\Q$. However, $\alpha$ and $\beta$ are
    both roots of the polynomial
    \[
    x^2 - (\alpha +\beta) x + \alpha\beta\in\Q[x]
    \]
    which is irreducible in $\Q[x]$, since neither $\alpha$ nor $\beta$
    are rational. Hence $\Q[\alpha,\beta]= \Q(\alpha)$ and
    $[\Q(\alpha):\Q] = 2$.


    Now we repeat this process for $\zeta_p^i$ that appear as summands
    of $\alpha$ and use Lemma~\ref{L:fgf} to ensure that our minimal
    polynomial has coefficients in the lower field. At each step, our
    list of $\zeta_p^i$, is cut in half, and hence after $2^n-1$
    steps, we will have a tower of degree two field extensions as
    required.  Hence we can construct a regular $p$-gon with compass
    and straightedge alone when $p=2^n+1$.
  \end{proof}
\end{theorem}

In our work above to show certain $n$-gons are constructible, it was
fairly easy to show that we had a field extension of degree $2^n$. The
work came in showing that there was a tower of field extensions where
each subextension was of degree $2$.

You may wonder, was all that work necessary? The answer lies in
\textit{Galois theory.} Don't be disheartened that we leave on a
cliffhanger. This means that there is more to learn. Remember,
there's always more, and that is wonderful.

\end{document}
