\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{Groups up to order eight}

\begin{document}
\begin{abstract}
  We classify groups up to order eight.
\end{abstract}
\maketitle

This section may be too demanding of the average reader. However, you
are no average reader! You are an intreped young mathematician!

We, and by we, I kind of mean \textit{you}, are going to classify all
groups up to order eight. This means that for each order, you will
find a set of groups such that any group of that order is isomorphic
to a group in your list.

\begin{definition}
  Two groups $G$ and $H$ are said to be of the same \dfn{isomorphism
    class} if $G\iso H$. A group $R$ is a \dfn{representative of an
    isomorphism class} if $R$ is isomorphic to the groups in the
  class.
\end{definition}


To start your goal is to fill-in the boxes in the theorem and lemmas
below.

\begin{theorem}[Classification of groups up to order eight]
  There are \raisebox{-1ex}{\framebox(30,20){}} isomorphism classes of groups up to
  order eight. The specfic list of isomorphism classes will be listed
  below by order.
\end{theorem}


\section{Orders one}

\begin{lemma}[Groups of order one]
  The group of order one is isomorphic to:
  \[
  \framebox(300,100){}
  \]
  \begin{sketch}
    There's only one! (I claim this is a hilarous pun.)
  \end{sketch}
\end{lemma}

\section{Order two, three, five, and seven}

We will prove a stronger result.

\begin{lemma}[Groups of prime order]
  Groups of prime order are isomorphic to:
  \[
  \framebox(300,100){}
  \]
  \begin{sketch}
    Prove these groups are always cyclic.
  \end{sketch}
\end{lemma}

\begin{corollary}[Groups of prime order less than eight]
  Groups of prime order less than eight are isomorphic to one of:
  \[
  \framebox(300,100){}
  \]
\end{corollary}

\section{Order four}

\begin{lemma}[Groups of order four]
  Groups of order four are isomorphic to:
  \[
  \framebox(300,100){}
  \]
  \begin{sketch}
    Look at possible orders of elements.
  \end{sketch}
\end{lemma}


\section{Order six}

\begin{lemma}[Groups of order six]
  Groups of order six are isomorphic to:
  \[
  \framebox(300,100){}
  \]
  \begin{sketch}
    Look at possible orders of elements and think about the different
    ways you could have a group that is not Abelian.
  \end{sketch}
\end{lemma}





\section{Order eight}



\begin{lemma}[Groups of order eight]
  Groups of order eight are isomorphic to:
  \[
  \framebox(300,100){}
  \]
  \begin{sketch}
    Look at possible orders of elements and think about the different
    ways you could have a group that is not Abelian.
  \end{sketch}
\end{lemma}




\end{document}
