\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{Dihedral groups}

\begin{document}
\begin{abstract}
  We introduce dihedral groups.
\end{abstract}
\maketitle

Earlier we mentioned that the set of all bijections from a set $X$ to
$X$ is a group under composition. We will now prove this.


\begin{theorem}[Automorphisms form a group]\label{E:aut0}
  Given a set $X$, prove that the collection of all bijections from
  $X$ to itself is a group under function composition.  This group is
  called the \dfn{automorphism}\textbf{ group} of $X$, and is denoted
  \[
  \aut(X).
  \]
  \begin{sketch}
    Just check the definition of a group. First show $\aut(X)$ is
    closed under $\circ$. Next explain why the group operation is
    associative. Next find the identity element. Finally, explain why
    every element has an inverse.
  \end{sketch}
\end{theorem}


If you find $\aut(X)$ difficult to think about, don't worry. We'll
take small steps to understand this.




To start, imagine an equilateral triangle sitting
on an empty $2$D-plane:
\[
\begin{tikzpicture}
  \pic {tikzngon={3}};
\end{tikzpicture}
\]
We are thinking about the points that lie on the lines that contain
the edges of the triangle.  Call this set of points, this
\dfn{geometric set}, $T$. In symbols we could write
\[
T = \{(x,y):\text{the point $(x,y)$ is on the triangle}\}
\]
Unfortunately the group $\aut(T)$ seems much to hard (at least at this
point, to this author) to tackle. Instead, let me tell you something I
tell myself in hard times:

\begin{quote}
  For every problem I cannot solve, there's an easier problem that I
  cannot solve.
\end{quote}



So here's my proposal for an easier problem. Let's study a subset of
$\aut(T)$. Let's roll it way back to middle school. We'll study the
\textit{symmetries} of $T$, the symmetries of $T$ will be a subset of
$\aut(T)$. An informal definition of symmetry might be:
\begin{quote}
  Symmetry is immunity to a possible change.
\end{quote}

With this in mind we will think of a special class of groups.

\begin{definition}
  Consider the symmetries of the regular $n$-gon. This set of
  symmetries forms a group under functional composition called a
  \dfn{dihedral group}. We denote the set of symmetries of the regular
  $n$-gon as $D_n$.
\end{definition}

The easiest $n$-gon in two-dimensional space is the triangle. Hence,
let's try to understand $D_3$.  In this case,
\[
D_3\subset \aut(T).
\]
There are six symmetries of the equilateral triangle. Let's list
them. There are $3$ flips (reflections across lines) $f_1$, $f_2$,
$f_3$:
\[
\begin{tikzpicture}
  \tikzmath{\rrr=1.7;}
  \pic {tikzngon={3}};
  \draw[fill] ({\rrr*sin(0)},{\rrr*cos(0)}) circle (1mm) node[right] {$3$};
  \draw[fill] ({\rrr*sin(360/3)},{\rrr*cos(360/3)}) circle (1mm) node[right] {$2$};
  \draw[fill] ({\rrr*sin(2*360/3)},{\rrr*cos(2*360/3)}) circle (1mm) node[left] {$1$};

  \pic at (-4,-4) {tikzngon={3}};
  \draw[dashed] ({-4+sqrt(3)*\rrr/4},{\rrr/4-4}) --  ({-\rrr*sin(360/3)-4},{\rrr*cos(360/3)-4});
  \draw[fill] ({\rrr*sin(0)-4},{\rrr*cos(0)-4}) circle (1mm) node[right] {$2$};
  \draw[fill] ({\rrr*sin(360/3)-4},{\rrr*cos(360/3)-4}) circle (1mm) node[right] {$3$};
  \draw[fill] ({\rrr*sin(2*360/3)-4},{\rrr*cos(2*360/3)-4}) circle (1mm) node[left] {$1$};
  
  \pic at (0,-4) {tikzngon={3}};
  \draw[dashed] (0,{\rrr-4}) -- (0,{\rrr*cos(360/3)-4});
  \draw[fill] ({\rrr*sin(0)},{\rrr*cos(0)-4}) circle (1mm) node[right] {$3$};
  \draw[fill] ({\rrr*sin(360/3)},{\rrr*cos(360/3)-4}) circle (1mm) node[right] {$1$};
  \draw[fill] ({\rrr*sin(2*360/3)},{\rrr*cos(2*360/3)-4}) circle (1mm) node[left] {$2$};

  \pic at (4,-4) {tikzngon={3}};
  \draw[dashed] ({4-sqrt(3)*\rrr/4},{\rrr/4-4}) --  ({\rrr*sin(360/3)+4},{\rrr*cos(360/3)-4});
  \draw[fill] ({\rrr*sin(0)+4},{\rrr*cos(0)-4}) circle (1mm) node[right] {$1$};
  \draw[fill] ({\rrr*sin(360/3)+4},{\rrr*cos(360/3)-4}) circle (1mm) node[right] {$2$};
  \draw[fill] ({\rrr*sin(2*360/3)+4},{\rrr*cos(2*360/3)-4}) circle (1mm) node[left] {$3$};

  
  \draw[|->] (0,-1.3) -- (0,-1.9); 
  \draw[|->] (-2,-1.5) -- (-3,-2.5);
  \draw[|->] (2,-1.5) -- (3,-2.5);

  \node[left] at (0,-1.6) {$f_2$};
  \node[above left] at (-2.5,-2) {$f_1$};
  \node[above right] at (2.5,-2) {$f_3$};
\end{tikzpicture}
\]
Note, the flip $f_1$ always flips across a line from the lower
left-hand vertex of the triangle to the opposite side; the flip $f_2$
always flips across a line from the top vertex of the triangle to the
opposite side; and $f_3$ always flips across a line from the lower
right-hand vertex of the triangle to the opposite side.


There are two counterclockwise rotations, $r_{120}$, $r_{240}$:
\[
\begin{tikzpicture}
  \tikzmath{\rrr=1.7;}
  \pic at (0,0) {tikzngon={3}};
  \draw[fill] ({\rrr*sin(0)},{\rrr*cos(0)}) circle (1mm) node[right] {$3$};
  \draw[fill] ({\rrr*sin(360/3)},{\rrr*cos(360/3)}) circle (1mm) node[right] {$2$};
  \draw[fill] ({\rrr*sin(2*360/3)},{\rrr*cos(2*360/3)}) circle (1mm) node[left] {$1$};


  \node[above] at (-2,.5) {$r_{120}$};
  \node[above] at (2,.5) {$r_{240}$};

  \pic at (-4,0) {tikzngon={3}};
  \draw[fill] ({\rrr*sin(0)-4},{\rrr*cos(0)}) circle (1mm) node[right] {$2$};
  \draw[fill] ({\rrr*sin(360/3)-4},{\rrr*cos(360/3)}) circle (1mm) node[right] {$1$};
  \draw[fill] ({\rrr*sin(2*360/3)-4},{\rrr*cos(2*360/3)}) circle (1mm) node[left] {$3$};
  \draw[dashed,->] (-3.5,0) arc (0:120:.5);
  
  \pic at (4,0) {tikzngon={3}};
  \draw[fill] ({\rrr*sin(0)+4},{\rrr*cos(0)}) circle (1mm) node[right] {$1$};
  \draw[fill] ({\rrr*sin(360/3)+4},{\rrr*cos(360/3)}) circle (1mm) node[right] {$3$};
  \draw[fill] ({\rrr*sin(2*360/3)+4},{\rrr*cos(2*360/3)}) circle (1mm) node[left] {$2$};
  \draw[dashed,->] (4.5,0) arc (0:240:.5);
  
  \draw[|->] (-1,.5) -- (-3,.5);
  \draw[|->] (1,.5) -- (3,.5);  

  
\end{tikzpicture}
\]
and an identity symmetry, the ``do nothing'' symmetry, let's call it
$e$.

\begin{example}[The dihedral group of order six]
  The six symmetries
  \[
  \{e,r_{120},r_{240},f_1,f_2,f_3\}
  \]
  of the equilateral triangle form a group under function
  composition. This group is known as the \dfn{dihedral group}\textbf{
    of order six}, the \textbf{symmetries of the equilateral triangle},
  and is also called (and denoted) $D_3$.\index{symmetries!of the equilateral triangle}
  \begin{proof}
    Let's check the conditions for this to be a group. We know that
    the operation is associative, since functional composition is
    associative, Lemma~\ref{L:funCompAss}.

    There is an identity element, $e$. Every element we've seen so far
    has an inverse, namely the rotations are inverses of each other,
    and the flips are their own inverse.
    
    We must show that this set is closed under function composition.
    To see this we will make a Cayley table. We will express all the
    symmetries in terms of $e$, $r=r_{120}$, $f=f_2$, with the
    stipulation that when the rotation and flip are applied, we always
    apply the flip first. Hence
    \[
    r_{240} = r^2, \quad  f_1 = r^2\circ f, \quad  f_3 = r\circ f.
    \]
    In other words, $\{r,f\}$ generate $D_3$.  In our Cayley table
    below, let's omit the $\circ$, so we will write something like
    $f\circ r$ as simply $fr$.
    \[
    \renewcommand{\arraystretch}{1.6}
    \begin{array}{c!{\vline width 2pt}cccccc}
      (D_3,\circ)& e     & r\cellcolor{blue!12!white}     & r^2\cellcolor{blue!24!white}   & f \cellcolor{red!12!white}    & r f \cellcolor{blue!33!red!24!white}  & r^2 f\cellcolor{blue!66!red!24!white} \\  \Xhline{2pt}
      e          & e     & r\cellcolor{blue!12!white}    & r^2\cellcolor{blue!24!white}   & f \cellcolor{red!12!white}    & r f \cellcolor{blue!33!red!24!white}   & r^2f \cellcolor{blue!66!red!24!white} \\  
      r\cellcolor{blue!12!white}         & r\cellcolor{blue!12!white}    & r^2\cellcolor{blue!24!white}   & e     & rf\cellcolor{blue!33!red!24!white} & r^2f \cellcolor{blue!66!red!24!white}    & f \cellcolor{red!12!white}   \\  
      r^2\cellcolor{blue!24!white}        & r^2\cellcolor{blue!24!white}   & e     & r\cellcolor{blue!12!white}    & r^2 f \cellcolor{blue!66!red!24!white}   & f \cellcolor{red!12!white} & rf \cellcolor{blue!33!red!24!white}    \\  
      f \cellcolor{red!12!white}         & f \cellcolor{red!12!white}    & r^2f \cellcolor{blue!66!red!24!white}   & r f \cellcolor{blue!33!red!24!white} & e     & r^2\cellcolor{blue!24!white}    & r\cellcolor{blue!12!white}   \\  
      r f \cellcolor{blue!33!red!24!white}        & rf \cellcolor{blue!33!red!24!white}   & f \cellcolor{red!12!white} & r^2f \cellcolor{blue!66!red!24!white}    & r\cellcolor{blue!12!white}   & e     & r^2\cellcolor{blue!24!white}    \\  
      r^2 f\cellcolor{blue!66!red!24!white}      & r^2 f\cellcolor{blue!66!red!24!white} & rf \cellcolor{blue!33!red!24!white}    & f \cellcolor{red!12!white}   & r^2\cellcolor{blue!24!white}    & r\cellcolor{blue!12!white}   & e     \\  
    \end{array}
    \]
    Note, from the table above we can see that $D_3$ is not Abelian.
  \end{proof}
\end{example}


\begin{exercise}
  Use Cayley tables to show that the group $F$ (Exercise~\ref{E:invg})
  is isomorphic to the group $D_3$.
\end{exercise}



If a set has symmetry, then that symmetry is a function that doesn't
change the set. In our case, the symmetries map $T$ to itself, and
they maintain the distance between points.  Automorphisms that
maintain the distance between points are called
\textit{isometries}.

\begin{definition}
  Let $X\subset \R^n$. An \dfn{isometry} of $X$ is a function $f:X\to X$ such
  that the distance between points on $X$ is preserved by $f$. This means
  that if $a,b\in X$, and $\nu$ is the vector from $a$ to $b$ then
  \[
  |\nu| = |f(\nu)|,
  \]
  where $|\nu|$ is the magnitude of vector $\nu$ and
  $|f(\nu)|$ is the magnitude of vector $f(\nu)$.
\end{definition}

\begin{exercise}
  Prove that if $X$ is a geometric set, a surjective isometry $f:X\to
  X$ will always be bijective.
\end{exercise}


Isometries of an $n$-gon are exactly the symmetries of the
$n$-gon. These symmetries (functions that are isometries on the
$n$-gon that map the $n$-gon to itself) will always form a group. This
type of a group is called a \textit{dihedral} group.

\begin{definition}
  A \dfn{dihedral group} a group of symmetries of the regular
  $n$-gon. We denote this group as $D_n$.
\end{definition}



\begin{exercise}
  Make a Cayley table for $D_4$, the symmetries of the regular
  $4$-gon. Write the elements so that $f$ is applied before $r$, that is
  $r^nf$:
  \[
  \begin{tikzpicture} %% N-GON
    \pic {tikzegon={4}};
  \end{tikzpicture}
  \]
\end{exercise}


\begin{exercise}
  Make a Cayley table for $D_5$, the symmetries of the regular
  $5$-gon. Write the elements so that $f$ is applied before $r$, that is
  $r^nf$:
  \[
  \begin{tikzpicture} %% N-GON
    \pic {tikzngon={5}};
  \end{tikzpicture}
  \]
\end{exercise}


\begin{exercise}
  Make a Cayley table for $D_6$, the symmetries of the regular
  $6$-gon. Write the elements so that $f$ is applied before $r$, that
  is $r^nf$:
  \[
  \begin{tikzpicture} %% N-GON
    \pic {tikzegon={6}};
  \end{tikzpicture}
  \]
\end{exercise}

\begin{exercise}
  Give an argument explaining why in $2$D, every flip or rotation can
  be written in terms of one flip, and compositions of rotations.
  \begin{hint}
    Apply the flip first. 
  \end{hint}
\end{exercise}

\begin{exercise}
  Prove that $D_n$ is never Abelian.
\end{exercise}


\begin{exercise}\label{E:D2n}
  Let $\mathcal{S}$ be a geometric set for a regular $n$-gon.
  \begin{enumerate}
  \item How many rotations are there in $\aut(\mathcal{S})$? 
  \item How many flips (reflections across lines) are there in
    $\aut(\mathcal{S})$?
  \item Find the order of $D_n$. 
  \end{enumerate}
  In each case, prove your answer is correct.
\end{exercise}



\begin{definition}
  Let $(G,\star)$ be a group, the \dfn{order of an element} $g\in G$ is
  \[
  \o(g) := \min\left\{n: \bigstar_{i=1}^n g = e\right\}\in\N.
  \]
  If the group has an operation like multiplication, $(G,\cdot)$, we will write
  \[
  \o(g) := \min\left\{n: \prod_{i=1}^n g = 1\right\}\in\N.
  \]
  If the group has an operation like addition, $(G,+)$, we will write
  \[
  \o(g) := \min\left\{n: \sum_{i=1}^n g = 0\right\}\in\N.
  \]
\end{definition}

\begin{exercise}
   Find the order of every element of $D_3$. Organize this information
   into a table, \textbf{make wild conjectures.}
\end{exercise}



\begin{exercise}
   Find the order of every element of $D_4$. Organize this information
   into a table, \textbf{make wild conjectures.}
\end{exercise}






\section{Symmetries of the tetrahedron}


We can also study the symmetries of objects in higher
dimensions. Working in $3$D should prove challenging enough, so let's
do that. It's a fact, there are exactly five regular solids, called
\link[\textit{Platonic solids}]{https://en.wikipedia.org/wiki/Platonic\_solid}.

\begin{definition}
  The \textbf{symmetries of the five Platonic solids} will be denoted
  by:
  \begin{enumerate}
  \item $\mathcal{T}_{12}$ for the regular tetrahedron.
  \item $\mathcal{C}_{24}$ for the cube.
  \item $\mathcal{O}_{24}$ for the regular octahedron.
  \item $\mathcal{D}_{60}$ for the regular dodecahedron.
  \item $\mathcal{I}_{60}$ for the regular icosahedron.
  \end{enumerate}
\end{definition}

\begin{exercise}
  How many faces do each of the regular polyhedra have?
\end{exercise}


Now let's study the symmetries of the tetrahedron in $3$ dimensions.
As it turns out, these symmetries will form a group (though not a
dihedral group). This group will have some interesting properties.


A regular \dfn{tetrahedron} is a triangular based pyramid where each
side is an equilateral triangle.

\[
\begin{tikzpicture}
  \draw[ultra thick,rounded corners=.05mm] (0,0) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (3,1) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (0,0) -- (3,1);
  %% \draw[fill] (3,1) circle (1mm) node[above right] {$1$};
  %% \draw[fill] (0,0) circle (1mm) node[left] {$2$};
  %% \draw[fill] (3,0) circle (1mm) node[right] {$3$};
  %% \draw[fill] (1.5,2.8) circle (1mm) node[above] {$4$};
\end{tikzpicture}
\]
Let's think about the symmetries of this tetrahedron. The first one
will be a counterclockwise rotation of the base:
\[
\begin{tikzpicture}
  \draw[ultra thick,rounded corners=.05mm] (0,0) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (3,1) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (0,0) -- (3,1);
  \draw[fill] (3,1) circle (1mm) node[above right] {$1$};
  \draw[fill] (0,0) circle (1mm) node[left] {$2$};
  \draw[fill] (3,0) circle (1mm) node[right] {$3$};
  \draw[fill] (1.5,2.8) circle (1mm) node[above] {$4$};

  \draw[|->] (3,2) -- (5,2);
  \node[above] at (4,2) {$r$};
  
  \draw[ultra thick,rounded corners=.05mm] (5,0) -- (8,0) -- (6.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (8,1) -- (8,0) -- (6.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (5,0) -- (8,1);
  \draw[fill] (8,1) circle (1mm) node[above right] {$3$};
  \draw[fill] (5,0) circle (1mm) node[left] {$1$};
  \draw[fill] (8,0) circle (1mm) node[right] {$2$};
  \draw[fill] (6.5,2.8) circle (1mm) node[above] {$4$};
\end{tikzpicture}
\]
The next one will be a counterclockwise pivot around the far vertex. 
\[
\begin{tikzpicture}
  \draw[ultra thick,rounded corners=.05mm] (0,0) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (3,1) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (0,0) -- (3,1);
  \draw[fill] (3,1) circle (1mm) node[above right] {$1$};
  \draw[fill] (0,0) circle (1mm) node[left] {$2$};
  \draw[fill] (3,0) circle (1mm) node[right] {$3$};
  \draw[fill] (1.5,2.8) circle (1mm) node[above] {$4$};

  \draw[|->] (3,2) -- (5,2);
  \node[above] at (4,2) {$p$};
  
  \draw[ultra thick,rounded corners=.05mm] (5,0) -- (8,0) -- (6.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (8,1) -- (8,0) -- (6.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (5,0) -- (8,1);
  \draw[fill] (8,1) circle (1mm) node[above right] {$1$};
  \draw[fill] (5,0) circle (1mm) node[left] {$4$};
  \draw[fill] (8,0) circle (1mm) node[right] {$2$};
  \draw[fill] (6.5,2.8) circle (1mm) node[above] {$3$};
\end{tikzpicture}
\]
From these two functions, we can write all the (orientation
preserving) symmetries of the tetrahedron:
\[
\{\underbrace{e,r,r^2}_{\text{$4$ on top}}, \underbrace{p,rp,r^2p}_{\text{$3$ on top}}, \underbrace{p^2,rp^2,r^2p^2}_{\text{$2$ on top}},\underbrace{pr^2,rpr^2,r^2pr^2}_{\text{$1$ on top}}\}
\]
Thus there are $12$ orientation preserving symmetries of the
tetrahedron.

\begin{example}[The tetrahedral group]
  The twelve symmetries
  \[
  \{e,r,r^2,p,rp,r^2p,p^2,rp^2,r^2p^2,pr^2,rpr^2,r^2pr^2\}
  \]
  of the regular tetrahedron form a group under function
  composition. This group is known as the \dfn{tetrahedral group}, the
  \textbf{symmetries of the regular tetrahedron}, and is also called (and
  denoted) $\mathcal{T}_{12}$.\index{symmetries!of the regular tetrahedron}
  \begin{proof}
    Let's check the conditions for this to be a group. We know that
    the operation is associative, since functional composition is
    associative, Lemma~\ref{L:funCompAss}.

    There is an identity element, $e$. Every element we've seen so far
    has an inverse, namely $r^3 = e$ and $p^3= e$. Any composition of
    $r$ and $p$ will have an inverse.
    
    We must show that this set is closed under function composition.
    To see this we will make a Cayley table. We will express all the
    symmetries in terms of
    \[
    \{e,r,r^2,p,rp,r^2p,p^2,rp^2,r^2p^2,pr^2,rpr^2,r^2pr^2\}
    \]
    In other words, $\{r,p\}$ generate $\mathcal{T}_{12}$.  
    \begin{gather*}
    \renewcommand{\arraystretch}{1.6}
      \begin{array}{c!{\vline width 2pt}cccccccccccc}
      (\mathcal{T}_{12},\circ)& e\cellcolor{red!6!white}     & r\cellcolor{red!12!white}     & r^2\cellcolor{red!24!white}   & p \cellcolor{yellow!6!white}    & rp \cellcolor{yellow!12!white}  & r^2 p\cellcolor{yellow!24!white} & p^2\cellcolor{green!6!white}    &  r p^2\cellcolor{green!12!white}  &  r^2 p^2\cellcolor{green!24!white} & pr^2 \cellcolor{blue!6!white} & rpr^2 \cellcolor{blue!12!white}& r^2 p r^2\cellcolor{blue!24!white}\\  \Xhline{2pt}
      e \cellcolor{red!6!white}         & e\cellcolor{red!6!white}         & r\cellcolor{red!12!white}           & r^2\cellcolor{red!24!white}   & p \cellcolor{yellow!6!white}    & rp \cellcolor{yellow!12!white}  & r^2 p\cellcolor{yellow!24!white} & p^2\cellcolor{green!6!white}    &  r p^2\cellcolor{green!12!white}  &  r^2 p^2\cellcolor{green!24!white} & pr^2 \cellcolor{blue!6!white} & rpr^2 \cellcolor{blue!12!white}& r^2 p r^2\cellcolor{blue!24!white}\\ 
      r\cellcolor{red!12!white}         & r\cellcolor{red!12!white}        & r^2\cellcolor{red!24!white}            & e\cellcolor{red!6!white}     & rp\cellcolor{yellow!12!white} & r^2p \cellcolor{yellow!24!white}    & p \cellcolor{yellow!6!white}   & rp^2 \cellcolor{green!12!white}  & r^2p^2 \cellcolor{green!24!white} & p^2 \cellcolor{green!6!white} & r pr^2 \cellcolor{blue!12!white} & r^2 pr^2 \cellcolor{blue!24!white} & pr^2 \cellcolor{blue!6!white}\\  
      r^2\cellcolor{red!24!white}       & r^2\cellcolor{red!24!white}      & e\cellcolor{red!6!white}           & r\cellcolor{red!12!white}    & r^2 p \cellcolor{yellow!24!white}   & p \cellcolor{yellow!6!white} & rp \cellcolor{yellow!12!white}   &  r^2p^2 \cellcolor{green!24!white}  & p^2 \cellcolor{green!6!white} & rp^2\cellcolor{green!12!white}  & r^2pr^2\cellcolor{blue!24!white} & pr^2\cellcolor{blue!6!white} & rpr^2\cellcolor{blue!12!white}\\  
      p \cellcolor{yellow!6!white}      & p \cellcolor{yellow!6!white}     & r^2p^2 \cellcolor{green!24!white}   & pr^2 \cellcolor{blue!6!white} & p^2\cellcolor{green!6!white}    & r^2\cellcolor{red!24!white}    & r^2pr^2\cellcolor{blue!24!white}  & e \cellcolor{red!6!white} & r^2p \cellcolor{yellow!24!white} &  rpr^2\cellcolor{blue!12!white}  &  rp \cellcolor{yellow!12!white}  &  r \cellcolor{red!12!white}  &  rp^2 \cellcolor{green!12!white}   \\
      r p \cellcolor{yellow!12!white}   & rp \cellcolor{yellow!12!white}   & p^2 \cellcolor{green!6!white}   & rpr^2 \cellcolor{blue!12!white}    & rp^2\cellcolor{green!12!white}   & e\cellcolor{red!6!white}     & pr^2\cellcolor{blue!6!white}  & r\cellcolor{red!12!white}  &  p \cellcolor{yellow!6!white}  & r^2pr^2\cellcolor{blue!24!white} & r^2p \cellcolor{yellow!24!white}  &  r^2 \cellcolor{red!24!white}  & r^2p^2 \cellcolor{green!24!white}  \\  
      r^2 p\cellcolor{yellow!24!white}  & r^2 p\cellcolor{yellow!24!white} & rp^2 \cellcolor{green!12!white}    & r^2 pr^2\cellcolor{blue!24!white}   & r^2p^2\cellcolor{green!24!white}    & r\cellcolor{red!12!white}   & rpr^2\cellcolor{blue!12!white} & r^2\cellcolor{red!24!white} &  rp \cellcolor{yellow!12!white}  &  pr^2 \cellcolor{blue!6!white}  & p \cellcolor{yellow!6!white}  &  e \cellcolor{red!6!white}  & p^2 \cellcolor{green!6!white}    \\
      p^2\cellcolor{green!6!white}      & p^2\cellcolor{green!6!white} & rpr^2 \cellcolor{blue!12!white} & rp \cellcolor{yellow!12!white} & e\cellcolor{red!6!white} &  pr^2 \cellcolor{blue!6!white}  &  rp^2 \cellcolor{green!12!white} &  p \cellcolor{yellow!6!white}  & r^2pr^2 \cellcolor{blue!24!white}  & r \cellcolor{red!12!white}   & r^2\cellcolor{red!24!white} &  r^2p^2 \cellcolor{green!24!white}  & r^2p \cellcolor{yellow!24!white}\\
      rp^2\cellcolor{green!12!white}    & rp^2\cellcolor{green!12!white} & r^2pr^2\cellcolor{blue!24!white} & r^2p \cellcolor{yellow!24!white} & r\cellcolor{red!12!white}& rpr^2 \cellcolor{blue!12!white}  & r^2p^2 \cellcolor{green!24!white} &  rp \cellcolor{yellow!12!white}  & pr^2 \cellcolor{blue!6!white}  & r^2 \cellcolor{red!24!white}   & e\cellcolor{red!6!white} &  p^2 \cellcolor{green!6!white}  & p \cellcolor{yellow!6!white}\\
      r^2p^2\cellcolor{green!24!white}  & r^2p^2\cellcolor{green!24!white}& pr^2 \cellcolor{blue!6!white} & p\cellcolor{yellow!6!white} & r^2\cellcolor{red!24!white}&  r^2pr^2 \cellcolor{blue!24!white}  &  p^2 \cellcolor{green!6!white} &  r^2p \cellcolor{yellow!24!white}  & rpr^2 \cellcolor{blue!12!white}  & e \cellcolor{red!6!white}   & r\cellcolor{red!12!white} &  rp^2 \cellcolor{green!12!white}  & rp \cellcolor{yellow!12!white}\\
      pr^2\cellcolor{blue!6!white}      & pr^2\cellcolor{blue!6!white} & p\cellcolor{yellow!6!white} & r^2p^2 \cellcolor{green!24!white} & r^2pr^2\cellcolor{blue!24!white}& p^2\cellcolor{green!6!white}& r^2\cellcolor{red!24!white} & rpr^2\cellcolor{blue!12!white} & e\cellcolor{red!6!white}  &  r^2p \cellcolor{yellow!24!white}  & rp^2 \cellcolor{green!12!white}  & rp \cellcolor{yellow!12!white}  & r \cellcolor{red!12!white}\\
      rpr^2\cellcolor{blue!12!white}    & rpr^2\cellcolor{blue!12!white} & rp\cellcolor{yellow!12!white} & p^2\cellcolor{green!6!white} & pr^2\cellcolor{blue!6!white} & rp^2\cellcolor{green!12!white}& e\cellcolor{red!6!white} & r^2pr^2\cellcolor{blue!24!white} & r\cellcolor{red!12!white}  &  p \cellcolor{yellow!6!white}  & r^2p^2 \cellcolor{green!24!white}  &  r^2p \cellcolor{yellow!24!white}  & r^2 \cellcolor{red!24!white}\\
      r^2pr^2\cellcolor{blue!24!white}  & r^2pr^2\cellcolor{blue!24!white}& r^2p\cellcolor{yellow!24!white} & rp^2\cellcolor{green!12!white} & rpr^2\cellcolor{blue!12!white} & r^2p^2\cellcolor{green!24!white}& r\cellcolor{red!12!white} & pr^2\cellcolor{blue!6!white} & r^2\cellcolor{red!24!white}  &  rp \cellcolor{yellow!12!white}  & p^2 \cellcolor{green!6!white}  &  p \cellcolor{yellow!6!white}  & e \cellcolor{red!6!white}\\
    \end{array}
    \end{gather*}
    Note, from the table above we can see that $\mathcal{T}_{12}$ is not Abelian.
  \end{proof}
\end{example}



\begin{exercise}
  Find the order of each element of $\mathcal{T}_{12}$. Organize this
  information into a table, \textbf{make wild conjectures.}
\end{exercise}


\begin{exercise}
  Consider the map:
  \[
  \begin{tikzpicture}
  \draw[ultra thick,rounded corners=.05mm] (0,0) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (3,1) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (0,0) -- (3,1);
  \draw[fill] (3,1) circle (1mm) node[above right] {$1$};
  \draw[fill] (0,0) circle (1mm) node[left] {$2$};
  \draw[fill] (3,0) circle (1mm) node[right] {$3$};
  \draw[fill] (1.5,2.8) circle (1mm) node[above] {$4$};

  \draw[|->] (3,2) -- (5,2);
  \node[above] at (4,2) {$f$};
  
  \draw[ultra thick,rounded corners=.05mm] (5,0) -- (8,0) -- (6.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (8,1) -- (8,0) -- (6.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (5,0) -- (8,1);
  \draw[fill] (8,1) circle (1mm) node[above right] {$4$};
  \draw[fill] (5,0) circle (1mm) node[left] {$1$};
  \draw[fill] (8,0) circle (1mm) node[right] {$3$};
  \draw[fill] (6.5,2.8) circle (1mm) node[above] {$2$};
  \end{tikzpicture}
  \]
  Is this map in $\mathcal{T}_{12}$? If so find the element. If not, explain why
  not.
\end{exercise}


\begin{exercise}
  Consider the map:
  \[
  \begin{tikzpicture}
  \draw[ultra thick,rounded corners=.05mm] (0,0) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (3,1) -- (3,0) -- (1.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (0,0) -- (3,1);
  \draw[fill] (3,1) circle (1mm) node[above right] {$1$};
  \draw[fill] (0,0) circle (1mm) node[left] {$2$};
  \draw[fill] (3,0) circle (1mm) node[right] {$3$};
  \draw[fill] (1.5,2.8) circle (1mm) node[above] {$4$};

  \draw[|->] (3,2) -- (5,2);
  \node[above] at (4,2) {$f$};
  
  \draw[ultra thick,rounded corners=.05mm] (5,0) -- (8,0) -- (6.5,2.8) -- cycle;
  \draw[ultra thick,rounded corners=.05mm] (8,1) -- (8,0) -- (6.5,2.8) -- cycle;
  \draw[ultra thick,dashed] (5,0) -- (8,1);
  \draw[fill] (8,1) circle (1mm) node[above right] {$1$};
  \draw[fill] (5,0) circle (1mm) node[left] {$3$};
  \draw[fill] (8,0) circle (1mm) node[right] {$2$};
  \draw[fill] (6.5,2.8) circle (1mm) node[above] {$4$};
  \end{tikzpicture}
  \]
  Is this map in $\mathcal{T}_{12}$? If so find the element. If not, explain why
  not.
\end{exercise}

\begin{exercise}
  Both the groups $D_6$ and $\mathcal{T}_{12}$ have $12$ elements. Is
  $D_6$ isomorphic to $\mathcal{T}_{12}$?
\end{exercise}



\begin{exercise}
  List all elements of $\mathcal{T}_{12}$ that leave vertex $4$
  fixed. Show that these form a group.
\end{exercise}

\begin{exercise}
  List all elements of $\mathcal{T}_{12}$ that leave vertex $3$
  fixed. Show that these form a group.
\end{exercise}

\begin{exercise}\label{E:symcube}
  Let $\mathcal{C}_{24}$ be the group of order preserving symmetries
  of the cube. Prove that $|\mathcal{C}_{24}| = 24$.
\end{exercise}

\begin{exercise}\label{E:symoct}
  Let $\mathcal{O}_{24}$ be the group of order preserving symmetries
  of the regular octahedron. Prove that $|\mathcal{O}_{24}|= 24$.
\end{exercise}

\begin{exercise}\label{E:symdoe}
  Let $\mathcal{D}_{60}$ be the group of order preserving symmetries
  of the regular dodecahedron. Prove that $|\mathcal{D}_{60}|=60$.
\end{exercise}

\begin{exercise}\label{E:symicos}
  Let $\mathcal{I}_{60}$ be the group of order preserving symmetries
  of the regular icosahedron. Prove that $|\mathcal{I}_{60}|=60$.
\end{exercise}

\begin{exercise}
  Give a geometric argument explaining \textbf{why}:
  \begin{align*}
    |\mathcal{C}_{24}| &= |\mathcal{O}_{24}|\\
    |\mathcal{D}_{60}| &= |\mathcal{I}_{60}|
  \end{align*}
  Further, explain why $|\mathcal{T}_{12}|$ has no ``partner'' solid,
  like the others do.
\end{exercise}









\begin{question}
  Symmetries of shapes can be complicated. What happens if we only
  study rotations in 2D?
\end{question}




\end{document}
