\documentclass{ximera}

\input{../preamble.tex}

\title{Generators and direct sums of cyclic groups}

\begin{document}
\begin{abstract}
  We further explore cyclic groups.
\end{abstract}
\maketitle

In this section, we will investigate when an element is a generator of
a cyclic group, and groups of the form $\Z_m\times \Z_n$ where
\[
\Z_m\times \Z_n :=\{(a,b):\text{$a\in\Z_m$ and $b\in\Z_n$}\}
\]
and addition is done componentwise. This construction is called the
\dfn{direct sum} of $\Z_m$ and $\Z_n$.




\section{Number theory background}

\begin{definition} \index{greatest common divisor}\index{GCD}
An integer $g$ is called a \textbf{greatest common divisor} of two
integer $a$ and $b$ provided that
\begin{enumerate}
\item $g| a$ and $g | b$.
\item If $d$ is an element where $d| a$ and $d | b$, then $d| g$.
\end{enumerate}
\end{definition}

\begin{exercise} 
Given $n\in \Z$, what is $\gcd(n,0)$? Prove your conclusion.
\end{exercise}

\begin{exercise} 
Prove that if $n$ is any integer, then $\gcd(a + n\cdot b, b) = \gcd(a, b)$.
\end{exercise}




\begin{theorem}[Euclid's lemma, version 1]\label{E:EL1}
  Given nonzero $a,b\in\Z$, we will prove that $g=(a,b)$ is the
  \textit{smallest} positive integer such that
  \[
  g = a\cdot m +b\cdot n
  \]
  for some integers $m$ and $n$.
  \begin{enumerate}
  \item Let $\mathcal S = \{x\in \N: x = a\cdot m +b\cdot n\text{ for } m,n\in
    \Z\}$. Prove that $\mathcal S$ has a least element, call it $d$.
  \item Prove that $d| x$ for all $x\in \mathcal S$.
  \item Prove that $d| a$ and $d| b$, explain why $1 \le d \le g$.
  \item Recall that $d = a\cdot m +b\cdot n$, and prove that $g |
    d$. Explain why we must conclude that $d = g$.
  \end{enumerate}
\end{theorem}



\begin{corollary}[Euclid's lemma, version 2]\label{E:EL2} 
  Let $a$, $b$, and $c$ be nonzero integers. Suppose that 
  \[
  a |bc\qquad\text{and}\qquad (a,b)=1.
  \]
  Prove that $a|c$. 
\end{corollary}



\begin{corollary}[Irreducible elements are prime]\label{E:EL}\index{prime element}\index{irreducible element} 
  We will prove that in the integers, all irreducible elements are
  prime, see (\ref{E:IRP}). Let $p$ be irreducible and let $a,b\in \Z$
  such that $p|ab$.
  \begin{enumerate}
  \item Suppose that $p\nmid a$, explain why $(a,p) =1$.
  \item Write $1 = am + pn$, multiply both sides by $b$.
  \item Can you finish it from here?
  \end{enumerate}
\end{corollary}



The statements of (\ref{E:EL1}), (\ref{E:EL2}), and (\ref{E:EL}) are
all sometimes referred to as \textbf{Euclid's Lemma}.



\begin{exercise} Prove that if $\gcd(m,n) = 1$, then 
\[
in \equiv jn \pmod m \qquad \Rightarrow\qquad i\equiv j\pmod m.
\]
\end{exercise}




\begin{theorem}[Euclidean Algorithm] 
We can easily compute the GCD of two numbers using the following
algorithm:
\vspace{2in}

\end{theorem}
\noindent The above space has intentionally been left blank for you to
fill in.


\begin{exercise} Study the following calculations:
\begin{align*}
22 &= \boldsymbol{6}\cdot 3 + \boldsymbol{4}\\ 6 &= \boldsymbol{4}
\cdot 1 + \fbox{$\boldsymbol{2}$}\\ 4 &= 2 \cdot 2 + 0 \qquad
\therefore \gcd(22,6) = 2
\end{align*}

\begin{align*}
33 &= \boldsymbol{24}\cdot 1 + \boldsymbol{9}\\
24 &= \boldsymbol{9} \cdot 2 + \boldsymbol{6}\\
9 &= \boldsymbol{6} \cdot 1 + \fbox{$\boldsymbol{3}$}\\
6 &= 3 \cdot 2 + 0 \qquad \therefore \gcd(33,24) = 3 
\end{align*}

\begin{align*}
42 &= \boldsymbol{16}\cdot 2 + \boldsymbol{10}\\
16 &= \boldsymbol{10} \cdot 1 + \boldsymbol{6}\\
10 &= \boldsymbol{6} \cdot 1 + \boldsymbol{4}\\
6 &= \boldsymbol{4} \cdot 1 + \fbox{$\boldsymbol{2}$}\\
4 &= 2 \cdot 2 + 0 \qquad \therefore \gcd(42,16) = 2 
\end{align*}

Explain how the above algorithm works and write it under the Euclidean
Algorithm above.
\end{exercise}

\begin{exercise} 
We will prove that when working with integers, the Euclidean Algorithm
will always produce the GCD of two numbers.
\begin{enumerate}
\item Prove that the remainders found in the Euclidean Algorithm form a
  decreasing sequence.
\item Prove that this sequence must terminate with a final remainder
  of zero.
\item Proceed by induction on the number of steps in the Euclidean
  Algorithm. If there are two steps:
\begin{align*}
a &= b\cdot q_1 + g \\
b &= g\cdot q_2 +0
\end{align*}
Prove that any divisor of both $a$ and $b$ is necessarily a divisor
of $g$. Explain why this proves that $g = (a,b)$.
\item Now suppose that any time we have $n+1$ equations:
\begin{align*}
a &= b\cdot q_1 + r_1 \\
b &= r_1\cdot q_2 + r_2 \\
  &\hspace{.5em}\vdots\\
r_{n-2} &= r_{n-1}\cdot q_{n} + r_n\\
r_{n-1} &= r_{n}\cdot q_{n+1} + 0
\end{align*}
that $r_n = (a,b)$. Prove that when we have $n+2$ equations, $r_{n+1}
= (a,b)$.
\item Explain how we have proved that when working with integers, the
  Euclidean Algorithm will always produce the GCD of two numbers.
\end{enumerate}
\end{exercise}


\begin{exercise} 
Prove that if $x = am +bn$ and $d$ is a common divisor of $a$ and $b$,
then $d|x$. What does this say about the GCD of $a$ and $b$?
\end{exercise}



\section{Generators of cyclic groups}


MV TO numberTheory.tex




In this section we will show that an $g\in \Z_n$ generates $\Z_n$ if
and only if $\gcd(g,n) = 1$.




\begin{theorem}
  An element $g\in \Z_n$ generates $\Z_n$ if and only if $\gcd(g,n) =
  1$.
\end{theorem}










\section{Direct products of cyclic groups}

The group defined by $(a,b)$ has order $\gcd(a,b)$.



\end{document}
