\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{Groups}

\begin{document}
\begin{abstract}
  We introduce the notion of a group.
\end{abstract}
\maketitle

Let us not mince words:

\begin{definition}
  A \dfn{group} is a set $G$ with an operation $\star$ such that the
  following properties hold:
  \begin{description}
  \item[Closure] If $a,b\in G$, then $a\star b\in G$.
  \item[Associativity] If $a,b,c\in G$, then
    \[
    a\star(b\star c)  = (a\star b)\star c.
    \]
  \item[Identity] There exists $e\in G$ such that for all $a\in G$, 
    \[
    e\star a = a \star e  = a.
    \]
  \item[Inverse] For all $a\in G$, there exists $a^{-1}$ such that
    \[
    a\star a^{-1} = a^{-1}\star a = e.
    \]
  \end{description}
  We will often write a group as $G$, or $(G,\star)$ when the
  operation is not clear. If the operation is multiplication (or
  something like multiplication), we often write $1$ instead of $e$;
  if the operation is addition (or something like addition) then we
  often write the inverse of $a$ as ``$-a$'' and the identity as $0$.
\end{definition}

\begin{exercise} %% supposed to test closure
  Which of the following are groups? Select all that apply.
  \begin{selectAll}
    \choice[correct]{$(\Z,+)$ you may assume addition is associative.}
    \choice{$(\{1,2,\dots,n\},+)$.}
    \choice{$(\N,\cdot)$ where $\N = \{1,2,3,\dots\}$, you may assume multiplication is associative.}
    \choice[correct]{$(\Z\times\Z,+)$ where
      \[
      \Z\times \Z := \{(a,b): a,b\in \Z\}
      \]
      and
      \[
      (a_1,b_1) + (a_2,b_2) := (a_1+a_2,b_1+b_2).
      \]}
    \choice[correct]{Setting $\Q^\times := \Q-\{0\}$, $(\Q^\times,\cdot).$}
  \end{selectAll}
\end{exercise}

\begin{exercise} %% supposed to test associativity
  Which of the following are groups? Select all that apply.
  \begin{selectAll}
    \choice{$(\{x^n:n\in \N\},\div)$}
    \choice[correct]{$(\{x^n:n\in \Z\},\cdot)$}
    \choice{$(\{x^n:n\in \Z\},\circ)$}
    \choice{$(\{x^n:n\in\Q^\times\},\cdot)$}
    \choice[correct]{$(\{x^n:n\in\Q^\times\},\circ)$}
  \end{selectAll}
\end{exercise}


\begin{exercise} %% harder constructions
  Which of the following are groups? Select all that apply.
  \begin{selectAll}
    \choice[correct]{$(2\Z,+)$ where $2\Z$ is the set of even integers.}
    \choice{$(\Z_{\mathrm{odd}},\cdot)$ where $\Z_{\mathrm{odd}}$ is
      the set of odd integers.}
    \choice[correct]{$(I(\x),+)$ where $\x$ is a set of points in $\R^2$ and
        \[
        I(\x) = \{F:\text{where $F:\R^2\to\R$ and $F(\x) = 0$}\}.
        \]}
    \choice{$(\Q[x],\cdot)$ where $\Q[x]$ is the set of polynomials in
      the variable $x$ with coefficients in $\Q$.}
    \choice[correct]{$(C^\infty(\R,\R), +)$ where $C^\infty(\R,\R)$ is
      the set of continuous functions mapping $\R\to\R$ such that
      every derivative of the function is also continuous.}
  \end{selectAll}
\end{exercise}

\begin{exercise}
  Let $(G,\star)$ and $(H,\diamond)$ be groups. Prove that
  \[
  G\times H := \{(g,h): \text{$g\in G$ and $h\in H$}\}
  \]
  is a group under
  \[
  (g_1,h_1) \circledast (g_2,h_2) := (g_1\star g_2, h_1\diamond h_2).
  \]
  $G\times H$ is called the \dfn{direct sum} of $G$ and $H$.
\end{exercise}

\begin{exercise}
  Consider the set $\Z\times \Z$ under the operation
  \[
  (a,b) \star(c,d) := (ad+bc,bd).
  \]
  Is this a group? If so, give a proof. If not, explain why not.
\end{exercise}


\begin{exercise}
  Find the largest subset $S\subset \Q$ such that $(S,\star)$ is a
  group where
  \[
  x\star y = x\cdot y + x + y
  \]
  and prove your answer is correct.
\end{exercise}




\begin{lemma}[Identity is unique]
  Let $G$ be a group. There is a unique identity element $e\in G$.
  \begin{sketch}
    Suppose there are two identities.
  \end{sketch}
\end{lemma}


\begin{lemma}[Inverses are unique]
  Let $G$ be a group. Every element in $G$ has a unique inverse.
  \begin{sketch}
    Suppose that an element has two inverses.
  \end{sketch}
\end{lemma}


\begin{exercise}
  If $g_1,g_2,\dots, g_n$ are elements of a group, prove that
  \[
  (g_1g_2\cdots g_n)^{-1} = g_n^{-1}\cdots g_2^{-1}g_1^{-1}.
  \]
\end{exercise}







\section{Finite groups}


So far, every group we've seen has had an infinite number of
elements. Groups can be finite too.

\begin{definition}
  If $G$ is a group, the \dfn{order of a group} $G$, denoted $|G|$ is
  equal to the number of distinct elements in $G$.
\end{definition}


\begin{example}\label{E:C}
  Let $C = \{1,i,-1,-i\}$ where $i =\sqrt{-1}$. We claim that $C$ is a
  group under multiplication.
  \begin{proof}
    First note that $C$ is closed under multiplication. Now also note
    that we know that multiplication is a associative. The identity
    element is $1$, and each element has an inverse, behold:
    \[
    i\cdot (-i) = 1, \quad\text{and}\quad (-1)\cdot(-1) = 1.
    \]
    So $C$ is a group of order $4$.
  \end{proof}
\end{example}


\begin{definition}
  Given a finite group $(G,\star)$, a \dfn{group table}, or
  \dfn{Cayley table}, is a complete $\star$-``multiplication'' table
  for $G$.
\end{definition}

\begin{example}
  Let's consider the group $C =\{1,i,-1,-i\}$ under multiplication once more.
  Write with me  
  \[
  \renewcommand{\arraystretch}{1.6}
  \begin{array}{c!{\vline width 2pt}cccccc}
    (C,\cdot)& 1     & i\cellcolor{blue!12!white}     & -1\cellcolor{blue!24!white}    & -i\cellcolor{blue!36!white}  \\  \Xhline{2pt}
    1        & 1     & i\cellcolor{blue!12!white}     & -1\cellcolor{blue!24!white}    & -i\cellcolor{blue!36!white}  \\  
    i\cellcolor{blue!12!white}        & i\cellcolor{blue!12!white}     & -1\cellcolor{blue!24!white}    & -i\cellcolor{blue!36!white}    & 1   \\  
    -1\cellcolor{blue!24!white}       & -1\cellcolor{blue!24!white}    & -i\cellcolor{blue!36!white}    & 1     & i\cellcolor{blue!12!white}   \\  
    -i\cellcolor{blue!36!white}       & -i\cellcolor{blue!36!white}    & 1     & i\cellcolor{blue!12!white}     & -1\cellcolor{blue!24!white}  
  \end{array}
  \]
  We colored in the cells of the table for dramatic effect. You don't
  need to do that, but it might help to see symmetry.  Sometimes, it
  is best to write a Cayley table using fewer symbols.  Noting that
  \[
  i^2 = -1 \quad\text{and}\quad i^3 = -i,
  \]
  we can also write:
  \[
  \renewcommand{\arraystretch}{1.6}
  \begin{array}{c!{\vline width 2pt}cccccc}
    (C,\cdot)& 1     & i\cellcolor{blue!12!white}     & i^2\cellcolor{blue!24!white}    & i^3\cellcolor{blue!36!white}  \\  \Xhline{2pt}
    1        & 1     & i\cellcolor{blue!12!white}     & i^2\cellcolor{blue!24!white}    & i^3\cellcolor{blue!36!white}  \\  
    i\cellcolor{blue!12!white}        & i\cellcolor{blue!12!white}     & i^2\cellcolor{blue!24!white}    & i^3\cellcolor{blue!36!white}    & 1   \\  
    i^2\cellcolor{blue!24!white}       & i^2\cellcolor{blue!24!white}    & i^3\cellcolor{blue!36!white}    & 1     & i\cellcolor{blue!12!white}   \\  
    i^3\cellcolor{blue!36!white}       & i^3\cellcolor{blue!36!white}    & 1     & i\cellcolor{blue!12!white}     & i^2\cellcolor{blue!24!white}  
  \end{array}
  \]
  You must write the entries of the Cayley table using elements from
  the top-most row; those same elements should also appear in the
  left-most column.
\end{example}



%% \begin{exercise}
%%   The following operation table is \textbf{not} a Cayley table for a
%%   group:
%%   \[
%%   \renewcommand{\arraystretch}{1.6}
%%   \begin{array}{|c||c|c|c|c|c|c|}\hline
%%      * & a & b & c & d & e & f  \\  \hline\hline
%%     a  & a & b & c & d & e & f  \\  \hline
%%     b  & b & c & d & e & f & a  \\  \hline
%%     c  & c & d & a & f & b & e  \\  \hline
%%     d  & d & e & f & a & c & b  \\  \hline
%%     e  & e & f & b & c & a & d  \\  \hline
%%     f  & f & a & e & b & d & c  \\  \hline
%%     \end{array}
%%   \]
%%   Which properties of a group fail?
%% \end{exercise}


Sometimes, Cayley tables for two groups $G$ and $H$ look essentially
``the same'' except for the symbols used. In this case, we say the
group $G$ is isomorphic to the group $H$.

\begin{definition}\index{isomorphic!groups}
  Two finite groups $G$ and $H$ are called \textbf{isomorphic groups} if
  they have the ``same'' Cayley table, with possibly some reordering
  of the elements. When a group $G$ is isomorphic to a group $H$, we
  write $G\iso H$.
\end{definition}

We will formalize this idea of ``rearranging group tables and see if
they look the same'' later. The idea will be to use special type of
function that respects the group structure to bijectively identify
elements in the one group to elements in the other group.

\section{Groups of functions}

Groups are absolutely fundamental to all of modern mathematics; their
importance cannot be overstated. I remember reading a similar line to
the one that precedes this one as an undergraduate. I didn't believe
it. However, make no mistake. Group theory permeates all of modern
mathematics. Let me explain.

One of the most powerful insights of 20th-Century mathematics is that
to understand an object in mathematics, you should understand the
functions on the object.

As an example of this claim, consider the paper \link[\textit{Real
    Analysis in Reverse}]{https://arxiv.org/pdf/1204.4483}, here the
author shows that the completeness of the reals is equivalent to
various theorems in calculus, theorems about functions. Hence they are
perhaps saying that to understand the real numbers, one must
understand the functions on the real numbers.

At this point, you may be wondering, ``First I'm told groups are
important, then I'm told that functions are important, well\dots which
is it?''

And the answer is, ``both.'' This is because if you have a set $X$
then the set of all bijections from $X$ to $X$ is a group under
composition, this will be the content of
Theorem~\ref{E:aut0}. However, we are not quite ready for this
theorem. We must start with a fundamental lemma.



\begin{lemma}[Function composition is associative]\label{L:funCompAss}
  Suppose that $A$, $B$, $C$, and $D$ are sets and that
  \begin{align*}
    \alpha:A &\to B\\
    \beta:B &\to C\\
    \gamma:C &\to D
  \end{align*}
  are functions. Then,
  \[
  \gamma\circ(\beta\circ\alpha) = (\gamma\circ\beta)\circ\alpha.
  \]
  \begin{proof}
    Let's put our maps in a row and represent the functions as arrows
    between the sets:
    \[
    \begin{tikzcd}[row sep=-1ex]%, column sep=-5ex]
      A \ar[r,"\alpha"] & B \ar[r,"\beta"] & C \ar[r,"\gamma"] & D\\
      a \ar[mapsto,r] & \alpha(a) = b  \ar[mapsto,r] & \beta(b) = c  \ar[mapsto,r] & \gamma(c)=d 
    \end{tikzcd}
    \]
    The second row above shows where the elements of these sets map
    to. The funny arrow with a base, ``$\mapsto$,'' is how we show
    where elements map. The diagram above represents
    $\gamma\circ\beta\circ\alpha$. In particular, with our diagram, we have set
    \[
    \gamma\circ\beta\circ\alpha(a) = d.
    \]
    Now let's make a similar diagram to
    represent $\gamma\circ(\beta\circ\alpha)$:
    \[
    \begin{tikzcd}[row sep=-1ex]%, column sep=-5ex]
      A \ar[r,"\alpha"] & B \ar[r,"\beta"] & C \ar[r,"\gamma"] & D\\
      a \ar[mapsto,rr,"\beta\circ\alpha",swap] &  & c  \ar[mapsto,r] & d 
    \end{tikzcd}
    \]
    We see that
    \[
    \gamma\circ(\beta\circ\alpha)(a) = d.
    \]
    Finally, let's make a similar diagram to
    represent $(\gamma\circ\beta)\circ\alpha$:
    \[
    \begin{tikzcd}[row sep=-1ex]%, column sep=-5ex]
      A \ar[r,"\alpha"] & B \ar[r,"\beta"] & C \ar[r,"\gamma"] & D\\
      a \ar[mapsto,r] & b \ar[rr,mapsto,"\gamma\circ\beta",swap] &   & d 
    \end{tikzcd}
    \]
    Again, we see
    \[
    (\gamma\circ\beta)\circ\alpha(a) = d.
    \]
    Since $a$ is an arbitrary label, and $d$ is simply a label for
    $\gamma\circ\beta\circ\alpha(a)$, we have shown
    \[
    \gamma\circ(\beta\circ\alpha)(a) = \gamma\circ\beta\circ\alpha(a) = (\gamma\circ\beta)\circ\alpha(a).
    \]
    We conclude that functional composition is always associative.
  \end{proof}
\end{lemma}




\begin{exercise}
  We write functions on the left, $f(x)$, meaning that the ``$f$'' is
  left of the ``$(x)$.''  Some authors write functions on the right,
  meaning they would write ``$(x)f$.'' Why would an author do this? Hint:
  Rephrase the lemma above, writing the functions on the right, and
  see if you notice anything nice.
  \begin{hint}
    Think about this:
    \[
    \underbrace{\gamma\circ\beta\circ\alpha}_{\text{left}}(a) = (a)\underbrace{\alpha\circ\beta\circ\gamma}_{\text{right}}
    \]
  \end{hint}
\end{exercise}



\begin{exercise}
  Prove that matrix multiplication is associative.
  \begin{hint}
    An $n\times m$ matrix is a function mapping $\R^m\to \R^n$ when
    you write your vectors vertically and functions on the left.
  \end{hint}
\end{exercise}




\begin{exercise}\label{E:SG}
  Pick up a sock. Now choose one:
  \begin{enumerate}
  \item[$(n)$] Do nothing to the sock.
  \item[$(i)$] Pull the toe of the sock out the inside of the foot
    opening. Hence if the sock was rightside out, it is now inside
    out, and vice versa.
  \end{enumerate}
  Someone says,
  \begin{quote}
    Ah! The sock group! The set $S =\{r,i\}$ is a group of order $2$ with
    \begin{align*}
      r\circ r &= r\\
  r\circ i &= i\\
  i\circ i &= r.
  \end{align*}
  \end{quote}
  Explain what this ``groupie'' is taking about. Be very
  detailed. Make a Cayley table with entries $n$ and $r$. 
\end{exercise}

\begin{exercise}\label{E:rf}
  In the Ohio State University Marching Band, there is a fundamental
  movement called a ``right-face.'' With this turn, the band-member
  spins to the right, $90^\circ$, ending up facing right. Denote a
  right-face by $R$. Explain how
  \[
  RF=\{R^0,R^1,R^2,R^3\}
  \]
  forms a group of order $4$. Make a Cayley table for this group where the entries
  are $R^0$, $R^1$, $R^2$, and $R^3$.
\end{exercise}

\begin{exercise}\label{E:st}
  In the Ohio State University Marching Band, there is a fundamental
  movement called a ``270-spin-turn.'' With this turn, the band-member
  spins to the left, $270^\circ$, thus ends up facing right. Denote a
  $270$-spin-turn by $T$. Explain how
  \[
  ST=\{T^0,T^1,T^2,T^3\}
  \]
  forms a group of order $4$. Make a Cayley table for this group where
  the entries are $T^0$, $T^1$, $T^2$, and $T^3$.
\end{exercise}

\begin{exercise}
  Use Cayley tables to show that the group $RF$ (Exercise~\ref{E:rf}),
  the group $ST$ (Exercise~\ref{E:st}), and the group $C$
  (Exercise~\ref{E:C}) are isomorphic to each other.
\end{exercise}



\begin{exercise}\label{E:invg}
  Let
  \begin{align*}
    i(x) &= x,             & \l(x) &= \frac{1}{x}, \\
    f(x) &= \frac{1}{1-x}, &  m(x) &= 1-x, \\
    g(x) &= \frac{x-1}{x}, &  n(x) &= \frac{x}{x-1},
  \end{align*}
  be functions from $\R-\{0,1\}\to\R$.  Prove that $F = \{i,f,g,\l,m,n\}$
  forms a group under function composition.
\end{exercise}


\begin{definition}
  Let $G$ be group, a subset $\{g_1,\dots, g_n,\dots\}\subset G$ is a
  set of \dfn{generators} for $G$ if for all $a\in G$
  \[
  a = \prod_{\mathrm{finite}} g_i^{\alpha_i}
  \]
  where $\alpha_i\in \Z$ and the $g_i$'s are not necessarily distinct.
\end{definition}


\begin{exercise}
  Prove that $C =\{1,i,-1,-i\}$ is generated by $i$. See Example~\ref{E:C}
\end{exercise}

\begin{exercise}
  Prove that $(\Z,+)$ is generated by $1$.
\end{exercise}

\begin{exercise}
  Prove that $(\Z,+)$ is generated by $3$ and $5$.
\end{exercise}

\begin{exercise}
  If $G$ is a group, and $g\in G$, prove that
  \[
  g^m\cdot g^n = g^{m+n}.
  \]
\end{exercise}

\begin{exercise}
  Prove that the group $\{i,f,g,\l,m,n\}$, as defined in
  Exercise~\ref{E:invg}, is generated by $f$ and $\l$. Make a Cayley
  table for this group with entries
  \[
  i, f, f\circ f, \l, f\circ \l,f^2\circ \l
  \]
  where $i$ is the identity function.
\end{exercise}

\begin{exercise}
  Let $x \in G$ where $(G,\cdot)$ is a group with identity element
  $e$. If $x^2 \ne e$ and $x^6 = e$, prove that that $x^4 \ne e$ and
  $x^5 \ne e$.
\end{exercise}


\section{Abelian groups}


Finally, note that the group operation need not be commutative.

\begin{definition}
  A group $(G,\star)$ is called \dfn{Abelian} if
  \[
  a\star b = b\star a
  \]
  for all $a,b\in G$.
\end{definition}

Abelian groups are named after \link[Niels Henrik Abel]{https://en.wikipedia.org/wiki/Niels_Henrik_Abel},
a great mathematician of his day (1802--1829), who died too young. 

\begin{exercise}
  Give examples of Abelian groups. Can you give an example of a group
  that is not Abelian? Hint: Think matrix multiplication.
\end{exercise}

\begin{exercise}
  Prove that a group $G$ is Abelian if and only if
  \[
  aba^{-1}b^{-1} = e
  \]
  for all $a,b\in G$.
\end{exercise}


\begin{exercise}
  Suppose that $G$ is a group where for all $x\in G$, $x^2 =e$. Prove
  that $G$ is Abelian.
\end{exercise}

\begin{exercise}
  Let $G$ be a group with the following property: Whenever $a$, $b$,
  and $c$ belong to $G$ and $ab = ca$, then $b = c$. Prove that $G$ is
  Abelian.
\end{exercise}


Now that we have some examples of a group, we should ask ourselves:

\begin{question}
  What is a good example of a finite group?
\end{question}



%% \section{Solving equations}

%% When you ask the pros what algebra is ``about,'' a common answer is:
%% \begin{quote}
%%   Algebra is about solving equations.
%% \end{quote}

%% Let me see if I can convince you of this. Suppose you have a group $G$
%% with elements $a,b\in G$. If you wish to solve this equation
%% \[
%% ax = b
%% \]
%% the definition of a group allows you \textit{always} solve this
%% equation by multiplying by $a^{-1}$ on the left, obtaining
%% \[
%% x = a^{-1}b.
%% \]
%% The definition of a group allows you to always solve linear equations
%% consisting of group elements.

%% On the other hand,














\end{document}
