\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{The correspondence theorem}

\begin{document}
\begin{abstract}
  We now inverstigate the subgroup structure of quotient groups.
\end{abstract}
\maketitle

This next theorem says that subgroups of a quotient group $G/N$ have
the same relative structure as subgroups of $G$ that contain $N$.

\begin{theorem}[Correspondence theorem]
  If $G$ is a group, $N$ is a normal subgroup of $G$, and
  \begin{itemize}
  \item $\mathcal{G} = \{S:N\subgp S\subgp G\}$
  \item $\mathcal{\bar{G}} = \{\bar{T}:\{\bar{e}\}\subgp \bar{T}\subgp G/N\}$,
  \end{itemize}
  \[
  \begin{tikzcd}
    N \ar[r,-,"\subgp"] \ar[d,<->,dotted]& S\ar[r,-,"\subgp"] \ar[d,<->,dotted]& G \ar[d,<->,dotted]\\
    \{\bar{e}\} \ar[r,-,"\subgp"] & \bar{T} \ar[r,-,"\subgp"] & G/N
  \end{tikzcd}
  \]
  then there is a bijective order-preserving map of sets
  $\mathcal{G}\leftrightarrow\mathcal{\bar{G}}$.
  \begin{proof}
    \textbf{First define a map from $\pmb{\mathcal{G}}$ to
      $\pmb{\mathcal{N}}$.} Let $\pi:G\to G/N$ be the canonical
    surjection.  By Lemma~\ref{L:hps}, define
    \begin{align*}
    \pi_*:\mathcal{G} &\to \bar{\mathcal{G}}\\
    S &\mapsto \pi(S).
    \end{align*}
    From this construction we see that if
    \[
    N \subset S'\subset S \subset G,
    \]
    then
    \[
    \{\bar{e}\}\subset \pi_*(S') \subset \pi_*(S)\subset G/N.
    \]
    \textbf{Now define a map from $\pmb{\mathcal{H}}$ to
      $\pmb{\mathcal{G}}$.}
    By Lemma~\ref{L:pps}, define
    \begin{align*}
      \pi^*:\bar{\mathcal{G}} &\to \mathcal G\\ \bar{T} &\mapsto
      \pi^{-1}(\bar{T}).
    \end{align*}
    From this construction we see that if
    \[
    \{\bar{e}\} \subset \bar{T}\hspace{0em}' \subset \bar{T} \subset G/N,
    \]
    then
    \[
    N \subset \pi^*(\bar{T}\hspace{0em}')\subset \pi^*(\bar{T}) \subset G.
    \]
    \textbf{Now we'll show that the maps above are inverses of each
      other.} We will show that $\pi^*(\pi_*(S)) = S$ and
    $\pi_*(\pi^*(\bar T)) = \bar{T}$. Both of these follow from the
    definition.

    \textbf{Now we'll show that our maps are bijective}. First note
    that
    \[
    \pi_*(\pi^*(\bar{T})) = \bar{T}
    \]
    shows that $\pi_*$ is surjective.

    We claim that $\pi_*$ is injective. Seeking a contradiction,
    suppose there exists $S,S'\in \mathcal{G}$ such that
    \[
    T = \pi(S) =\pi(S').
    \]
    In this case, $\pi^*(\bar T) = S\cup S'$. This contradicts the
    fact that $\pi_*$ and $\pi^*$ are inverses of each other.
  \end{proof}
\end{theorem}


\begin{corollary}[Correspondence of normal subgroups]
  If in the theorem above, $S$ is normal then $\pi_*(S)$ is normal; if
  $\bar{T}$ is normal, then $\pi^*(\bar{T})$ is normal.
\end{corollary}


\begin{corollary}[Correspondence and homomorphisms]
  Let $\phi:G\to H$ be a homomorphism of groups. Set
  \begin{itemize}
  \item $\mathcal{G} = \{S:\ker(\phi)\subgp S\subgp G\}$
  \item $\mathcal{H} = \{T:\{e_H\}\subgp T\subgp \im(\phi)\subset H\}$,
  \end{itemize}
  \[
  \begin{tikzcd}
    \ker(\phi) \ar[r,-,"\subgp"] \ar[d,<->,dotted]& S\ar[r,-,"\subgp"] \ar[d,<->,dotted]& G \ar[d,<->,dotted]\\
    \{e\} \ar[r,-,"\subgp"] & T \ar[r,-,"\subgp"] & \im(\phi)
  \end{tikzcd}
  \]
  then there is a bijective order-preserving map of sets
  $\mathcal{G}\leftrightarrow\mathcal{H}$.
  \begin{proof}
    By Lemma~\ref{L:kerN}, $\ker(\phi)$ is a normal subgroup of $G$,
    and by Noether's isomorphism theorem, Theorem~\ref{T:NI},
    $\im(\phi)\iso G/\ker(\phi)$.
  \end{proof}
\end{corollary}

\end{document}
