\documentclass{ximera}

\input{../preamble.tex}

\author{Bart Snapp}

\title{The fraction isomorphism theorem}

\begin{document}
\begin{abstract}
  We introduce the fraction isomorphism theorem, also known as the
  third isomorphism theorem.
\end{abstract}
\maketitle

\index{third isomorphism theorem}
Previously we seen groups like this


\begin{tikzcd}[row sep=0in]
    & \Z_{30}/\<10\>\ar[ldd,-]  \ar[white!50!black,ddd,-] \ar[rdddd,-] &       \\
    &\phantom{MM} & \\
    \<2\> /\<10\> \ar[white!50!black,ddd,-] &   &  \\
    & \color{white!50!black}\<3\>/\<10\>\ar[white!50!black,ldd,-]\ar[white!50!black,rdddd,-] & \\
    &  &  \<5\> /\<10\>\ar[white!50!black,ddd,-]\\
     \color{white!50!black}\<6\>/\<10\>\ar[white!50!black,rdddd,-]& & \\
    &\<10\>/\<10\> \equiv \<0\> \ar[white!50!black,ddd,-]\ar[ruu,-,white,line width=2mm]\ar[ruu,-]\ar[luuuu,-,white,line width=2mm]\ar[luuuu,-] & \\
    & & \color{white!50!black} \<15\>/\<10\>\ar[white!50!black,ldd,-]\\
    &\phantom{MM} & \\
    &  \color{white!50!black}\<0\>/\<10\> &
\end{tikzcd}
But how do we understand groups like 


\begin{theorem}[Fraction isomorphism theorem]
  Let $M\normal N\normal G$. In this case,
  \[
  \frac{G/M}{N/M} \iso \frac{G}{N}.
  \]
  \begin{proof}
    Our plan is to define a surjective map
    \[
    \pi:G/M \onto G/N,
    \]
    we'll show that $\ker(\pi) = N/M$, and hence by Noether's
    isomorphism theorem, Theorem~\ref{T:NI},
    \[
    \frac{G/M}{N/M} \iso \frac{G}{N}.
    \]
    \textbf{First we define a map from $\boldsymbol{G/M}$ to
      $\boldsymbol{G/N}$.} Let $a\in G$ and define
    \begin{align*}
      \pi: G/M &\onto G/N \\
      aM &\mapsto aN
    \end{align*}
    \textbf{We must check this map is
      well-defined.}\index{well-defined} Suppose for $a,b\in G$
    \[
    a M = bM.
    \]
    we must show that $aN = bN$. First note that if $aM = bM$, then $a
    \in bM$. Hence $a=bm$ for some $m\in M$. Write with me
    \begin{align*}
      \pi(aM) &= aN\\
      &= bm N\\
      &= bN.  & &\text{(since $M\subset N$)}
    \end{align*}
    Since $\pi(a M) = \pi(bM)$, our map is well-defined.

    \textbf{Now we must show that $\boldsymbol\pi$ is a homomorphism.}
    Write with me
    \begin{align*}
      \pi(aMbM) &= \phi(ab M) & &\text{(by  Lemma~\ref{L:cpwd})} \\
      &= ab N \\
      &= aN bN & &\text{(by  Lemma~\ref{L:cpwd})} \\
      &= \pi(aM)\pi(bM).
    \end{align*}


    \textbf{We must show that $\boldsymbol{\pi}$ is surjective.}
    Consider any element $a\in G$. Then $aN\in \im(\pi)$ since
    $M\subset N$. In particular, this shows that $\im(\pi) = G/N$.

    
    \textbf{We must show that $\boldsymbol{\ker(\pi)=N/M}$.} We'll use
    double containment.

    $(\subset)$ We'll show that $\ker(\pi)\subset N/M$.  Suppose that
    $kM\in\ker(\pi)$. In this case, $k\in N/M$.
    

    $(\supset)$ We'll show that $N/M\subset \ker(\pi)$. Suppose that
    $nM\in N/M$. In this case, $\pi(nM) = nN = N$. Thus
    $\ker(\pi)=N/M$.


    \textbf{Use Noether's isomorphism theorem to complete the proof.} Since
    \[
    \ker(\pi) = N/M\quad\text{and}\quad\im(\pi) = G/N,
    \]
    by Noether's isomorphism theorem
    \begin{align*}
    \frac{G/M}{\ker(\pi)} &\iso \im(\pi)\\
    \frac{G/M}{N/M} &\iso \frac{G}{N}.
    \end{align*}
    This completes the proof.
  \end{proof}
\end{theorem}

\end{document}
